### 2016-04-05 paper/paper.tex
Started writing the paper.

\documentclass[a4paper,UKenglish]{lipics-v2016}

\usepackage{microtype}

\bibliographystyle{plainurl}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}}

\author[1]{Jouni Sirén}
\affil[1]{Wellcome Trust Sanger Institute, Hinxton, Cambridge, UK\\
  \texttt{jouni.siren@iki.fi}}
\authorrunning{J. Sirén}

\Copyright{Jouni Sirén}

\subjclass{E.1 Data Structures}
\keywords{Dummy keyword -- please provide 1--5 keywords}% FIXME
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Mathematics
\newcommand{\set}[1]{\ensuremath{\{ #1 \}}}
\newcommand{\abs}[1]{\ensuremath{\lvert #1 \rvert}}

% DNA
\newcommand{\dnaseq}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\baseA}{\dnaseq{A}}
\newcommand{\baseC}{\dnaseq{C}}
\newcommand{\baseG}{\dnaseq{G}}
\newcommand{\baseT}{\dnaseq{T}}
\newcommand{\baseN}{\dnaseq{N}}
\newcommand{\dnacomp}[1]{\ensuremath{\overline{#1}}}
\newcommand{\revcomp}[1]{\ensuremath{\overleftarrow{#1}}}

% Rank/select
\newcommand{\rank}{\textsf{rank}}
\newcommand{\select}{\textsf{select}}
\newcommand{\mrank}{\ensuremath{\mathsf{rank}}}
\newcommand{\mselect}{\ensuremath{\mathsf{select}}}

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}

% Shorthands
\newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
\newcommand{\orderk}[1]{order\nobreakdash-$#1$}

\begin{document}

\maketitle

\begin{abstract}% FIXME
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti.
 \end{abstract}

\section{Introduction}

1000GP: \cite{1000GP2015}

Graph genomes: \cite{Schneeberger2009}

FM-index: \cite{Ferragina2005a}

CST-NPR: \cite{Fischer2009a}

CST MEMs: \cite{Ohlebusch2010a}

XBW: \cite{Ferragina2009b}

GCSA: \cite{Siren2014}

Succinct DBG: \cite{Bowe2012}, variable-order \cite{Boucher2014}, alternative representation \cite{Roedland2013}

BWBBLE: \cite{Huang2013}

Contracted DBG: \cite{Cazaux2014}

Compressed DBG: \cite{Marcus2014}

Manifold DBG: \cite{Lin2014}

Hypertext index: \cite{Thachuk2013}

FM-index of alignment: \cite{Na2015}

HISAT / HISAT2: \cite{Kim2015}

\section{Background}

\subsection{Strings}

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively.

In this paper, we primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$. Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base. In addition to the endmarker $\$$, the alphabet may also contain other characters for technical purposes. Each character $c$ of the DNA alphabet has a \emph{complement} $\dnacomp{c}$ defined as $\dnacomp{\baseA} = \baseT$, $\dnacomp{\baseC} = \baseG$, $\dnacomp{\baseG} = \baseC$, $\dnacomp{\baseT} = \baseA$, and $\dnacomp{c} = c$ for other characters $c$. Given a DNA sequence $S$, its \emph{reverse complement} is the sequence $\revcomp{S}$ obtained by reversing the non-technical parts of the sequence and replacing each character with its complement. For example, $\revcomp{\dnaseq{GATTACA}\$} = \dnaseq{TGTAATC}\$$.

Given string $S[0, n-1]$, we define $S.\mrank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define $S.\mselect(i, c) = \arg \max_{j \le n} S.\mrank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions correspond to the ones used in the SDSL library \cite{Gog2014b}.} A \emph{bitvector} is a binary sequence supporting efficient \rank/\select{} queries. \emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support \rank/\select{} queries on arbitrary strings.

\subsection{Graphs}

A \emph{graph} $G = (V, E)$ consists of a set of \emph{nodes} $V = \set{0, \dotsc, \abs{V}-1}$ and a set of \emph{edges} $E \subseteq V \times V$. We say that $(u, v) \in E$ is an edge \emph{from} node $u$ \emph{to} node $v$, and assume that the edges are \emph{directed}: $(u, v) \ne (v, u)$ for $u \ne v$. The \emph{indegree} $G.\gindegree(v)$ of node $v$ is the number of \emph{incoming} edges to node $v$, while the \emph{outdegree} $G.\goutdegree(v)$ is the number of \emph{outgoing} edges from node $v$.

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for $0 \le i < \abs{P}-1$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$ not used elsewhere in the graph. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.

% FIXME example
Let $\mathcal{S}$ be a (potentially infinite) collection of strings over alphabet $\Sigma$, each of them ending with $k$ copies of the endmarker $\$$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$. Each node $v_{X}$ has a \emph{key} $G.\gkey(v_{X}) = X$ and a label $G.\glabel(v_{X}) = X[0]$. The edges of the de~Bruijn graph represent \kmer{k+1}s $Y$ occurring in the collection. For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$. We use the node corresponding to substring $\$^{k}$ as the sink node, and add the source node (with $G.\gkey(s) = \#$) and the associated edges in the usual way. We may also have edges from the source node to the nodes representing the prefixes of the collection.

\begin{lemma}[No false negatives]
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ not containing character $\$$. Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
\end{lemma}

\begin{proof}
Let $X = S[i, i+\abs{X}-1]$ for a string $S \in \mathcal{S}$. For $i \le j \le i+\abs{X}-1$, there is node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$. By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$. Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[No false positives with $\abs{P} \le k+1$]
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then path $P$ either contains the source node $s$ or its label is a substring of a string $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
Assume that path $P = v_{0} \dotsm v_{\abs{P}-1}$ does not contain node $s$. If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$. If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.

Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma \setminus \set{\#}$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

\subsection{Indexes}

\section{Path indexes}

\section{GCSA2}

\section{Construction}

\section{Experiments}

\section{Discussion}

% FIXME
\subparagraph*{Acknowledgements.}

I want to thank \dots

\bibliography{paper}

\end{document}

### 2016-04-06 paper/paper.tex
Sketched the sections on text indexes and path indexes.

@@
 
+Suffix tree \cite{Weiner1973}, suffix array \cite{Manber1993}
+
+BWT \cite{Burrows1994}, LF-mapping
+
+FM-index \cite{Ferragina2005a}, find, locate, extract
+
+LCP array \cite{Manber1993}, CST \cite{Sadakane2007}
+
+XBW \cite{Ferragina2009b}, GCSA \cite{Siren2014}
+
 \section{Path indexes}
+
+find the start nodes of paths with label matching the pattern
+
+general difficulty: exponential growth
+GCSA \cite{Siren2014}
+
+k-mer indexes based on hash tables, sorted k-mers
+
+similarities between k-mer indexes and de Bruijn graphs
+
+using de Bruijn graphs as indexes; false positives
+- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
+- add values (pointers to start nodes) to each node
+- valid for existence queries of length up to k+1 and locate queries of length up to k
+
+Path graphs
+- a generalization of DBG as a path index
+- keys must be prefix-free
+- k-index: has a node matching any k-path in the input graph
+- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
+- k-equivalent: union of values for nodes matching X identical for all |X| <= k
+- k-equivalent + edge condition: no false positives of length <= k; no false negatives
+
+pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
 

### 2016-04-07 paper/paper.tex
Finished the background section, still missing examples.

@@
 \newcommand{\abs}[1]{\ensuremath{\lvert #1 \rvert}}
+\newcommand{\Oh}[1]{\ensuremath{\mathsf{O}\!\left( #1 \right)}}
 
@@
 
-% Rank/select
-\newcommand{\rank}{\textsf{rank}}
-\newcommand{\select}{\textsf{select}}
-\newcommand{\mrank}{\ensuremath{\mathsf{rank}}}
-\newcommand{\mselect}{\ensuremath{\mathsf{select}}}
+% Queries
+\newcommand{\rank}{\ensuremath{\mathsf{rank}}}
+\newcommand{\select}{\ensuremath{\mathsf{select}}}
+\newcommand{\LF}{\ensuremath{\mathsf{LF}}}
 
@@
 \newcommand{\orderk}[1]{order\nobreakdash-$#1$}
+\newcommand{\LFmapping}{LF\nobreakdash-mapping}
+\newcommand{\FMindex}{FM\nobreakdash-index}
+
+% Structures
+\newcommand{\SA}{\ensuremath{\mathsf{SA}}}
+\newcommand{\BWT}{\ensuremath{\mathsf{BWT}}}
+\newcommand{\Carray}{\ensuremath{\mathsf{C}}}
+\newcommand{\LCP}{\ensuremath{\mathsf{LCP}}}
+\newcommand{\bvIN}{\ensuremath{\mathsf{IN}}}
+\newcommand{\bvOUT}{\ensuremath{\mathsf{OUT}}}
 
@@
 
-A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$.
+A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$.
 For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text.
@@
 
-Given string $S[0, n-1]$, we define $S.\mrank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
-We also define $S.\mselect(i, c) = \arg \max_{j \le n} S.\mrank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions correspond to the ones used in the SDSL library \cite{Gog2014b}.} A \emph{bitvector} is a binary sequence supporting efficient \rank/\select{} queries.
-\emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support \rank/\select{} queries on arbitrary strings.
+Given string $S[0, n-1]$, we define $S.\rank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
+We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions correspond to the ones used in the SDSL library \cite{Gog2014b}.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries.
+\emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support $\rank$/$\select$ queries on arbitrary strings.
 
@@
 
-% FIXME example
+% FIXME example of strings, DBG, long paths, false positives
 Let $\mathcal{S}$ be a (potentially infinite) collection of strings over alphabet $\Sigma$, each of them ending with $k$ copies of the endmarker $\$$.
@@
 
-\subsection{Indexes}
+\subsection{Text indexes}
 
-Suffix tree \cite{Weiner1973}, suffix array \cite{Manber1993}
+The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting arbitrary substring queries.
+It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges.
+Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.
 
-BWT \cite{Burrows1994}, LF-mapping
+% FIXME SA/BWT/LCP example
+\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree.
+The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}.
+Given a text and its suffix array, we can find the occurrences of \emph{pattern} string $X$ in the text by using \emph{binary search} in $\Oh{\abs{X} \log n}$ time.
+The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree.
 
-FM-index \cite{Ferragina2005a}, find, locate, extract
+The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text sharing the combinatorial structure of the suffix array.
+The BWT of text $T[0, n-1]$ is a string $\BWT[0, n-1]$ defined as $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
+Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$.
+Let
+$$
+\LF(i) = \Carray[\BWT[i]] + \BWT.\rank(i, \BWT[i]),
+$$
+where $\Carray[c]$ is the number of suffixes of the text starting with any character $c' < c$.
+Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$.
+We can also generalize the definition for any character $c \in \Sigma$:
+$$
+\LF(i, c) = \Carray[c] + \BWT.\rank(i, c).
+$$
+Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.
 
-LCP array \cite{Manber1993}, CST \cite{Sadakane2007}
+Given the similarities to the suffix array, we can use the BWT as a space-efficient text index.
+The \emph{\FMindex} \cite{Ferragina2005a} combines a plain or compressed representation of the BWT supporting rank/select queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
+It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}.
+If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, then the lexicographic range matching suffix $X[i, \abs{X}-1]$ of the pattern is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$.
+Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ rank queries.
 
-XBW \cite{Ferragina2009b}, GCSA \cite{Siren2014}
+We can use the sampled suffix array pointers to find the text positions containing the occurrences.
+If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer.
+If we find a sample at $\SA[\LF^{k}(i)]$, we know that
+$$
+\SA[i] = (\SA[\LF^{k}(i)] + k) \bmod n.
+$$
+If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries.
+If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.
+In a typical case, we expect an \FMindex{} to take less than $n \log \sigma$ bits of space and to be able to find about 100,000 pattern occurrences per second \cite{Ferragina2009a}.
+
+The \emph{longest-common-prefix array} (LCP array) \cite{Manber1993} is an integer array $\LCP[0, n-1]$, where each value $\LCP[i]$ tells the length of the longest common prefix of suffixes $T[\SA[i-1], n-1]$ and $T[\SA[i], n-1]$ (with $\LCP[0] = 0$).
+It is often used to augment the functionality of the suffix array and the \FMindex.
+In particular, if we augment the \FMindex{} with the LCP array and the topology of the suffix tree, we get the \emph{compressed suffix tree}, which supports the full functionality of the suffix tree in a space-efficient manner \cite{Sadakane2007}.
+
+\subsection{Generalized indexes}
+
+Suffix trees, suffix arrays, and \FMindex{} can be generalized to index multiple texts.
+There are also generalizations for other combinatorial structures.
+The \emph{XBW transform} \cite{Ferragina2009b} is essentially an \FMindex{} for \emph{labeled trees}.
+The nodes of the tree are sorted by path labels from the node to the root of the tree.
+$\BWT$ stores the labels of the children of each node (leaf nodes require special treatment).
+If a node has $k$ children, we encode that as binary sequence $0^{k-1} 1$.
+We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges.
+The labels of the children of node $i$ are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$ (with $0$ as the lower bound for $i = 0$).
+
+% FIXME LF(i, c) picture from the slides for the earlier DBG example
+The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and \emph{de Bruijn graphs}.
+Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the paths starting from them.
+The transformation is expensive and may increase the size of the graph exponentially.
+After sorting the nodes of the transformed graph, we encode them using three sequences.
+$\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform.
+\LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes into BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges into the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
 

### 2016-04-07 paper/paper.tex
Started writing the section on path indexes.

@@
 If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries.
-If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.
+If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.
 In a typical case, we expect an \FMindex{} to take less than $n \log \sigma$ bits of space and to be able to find about 100,000 pattern occurrences per second \cite{Ferragina2009a}.
@@
 
-find the start nodes of paths with label matching the pattern
+A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}.
+Given a path index for graph $G = (V, E)$, we want to use the index to find the \emph{start nodes} $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
 
-general difficulty: exponential growth
-GCSA \cite{Siren2014}
+The design of a path index is a trade-off between the maximum supported query length and the size of the index.
+In the worst case, a graph may have up to $(\sigma-2)^{k}$ distinct labels for paths of length $k$ (ignoring paths containing the source node or the sink node), and the length of possible paths is unlimited in cyclic graphs.
+GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary.
+The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.
 
+% FIXME from here
 k-mer indexes based on hash tables, sorted k-mers
 
-similarities between k-mer indexes and de Bruijn graphs
+similarities between k-mer indexes and de Bruijn graphs: order-k+1 index can simulate order-k DBG; order-k DBG can be used as an order-k index
+
+define de Bruijn graphs for graphs (infinite strings, no source node, repeat sink node instead)
 

### 2016-04-08 paper/paper.tex
Finished the section on basic path indexes.

@@
 
-\subsection{Graphs}
+\subsection{Graphs}\label{sect:graphs}
 
@@
 The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$.
-A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for $0 \le i < \abs{P}-1$.
+A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$.
 The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
+We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$, where set $Z$ is one of $\mathbb{Z}$, $\set{x \in \mathbb{Z} \mid x \ge 0}$, and $\set{x \in \mathbb{Z} \mid x \le 0}$.
+The labels of such paths are infinite sequences of characters instead of strings.
 
 We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
-To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$ not used elsewhere in the graph.
+To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$.
+The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes.
 We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
 We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
+However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.
 
 % FIXME example of strings, DBG, long paths, false positives
-Let $\mathcal{S}$ be a (potentially infinite) collection of strings over alphabet $\Sigma$, each of them ending with $k$ copies of the endmarker $\$$.
+Let $\mathcal{S}$ be a collection of strings over alphabet $\Sigma$.
+We assume that each string $S \in \mathcal{S}$ begins with $k$ copies of $\#$ and ends with $k$ copies of $\$$, with no other occurrences of these characters in the string.
 The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$.
@@
 For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$.
-We use the node corresponding to substring $\$^{k}$ as the sink node, and add the source node (with $G.\gkey(s) = \#$) and the associated edges in the usual way.
-We may also have edges from the source node to the nodes representing the prefixes of the collection.
+We use the node corresponding to substring $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
 
-\begin{lemma}[No false negatives]
-Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ not containing character $\$$.
+\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
+Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ containing at most one $\$$.
 Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
@@
 
-\begin{lemma}[No false positives with $\abs{P} \le k+1$]
+\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
 Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$.
-Then path $P$ either contains the source node $s$ or its label is a substring of a string $S \in \mathcal{S}$.
+Then the label of the path is a substring of a string $S \in \mathcal{S}$.
 \end{lemma}
@@
 \begin{proof}
-Assume that path $P = v_{0} \dotsm v_{\abs{P}-1}$ does not contain node $s$.
 If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$.
 If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.
-
-Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma \setminus \set{\#}$ and $1 \le \abs{X} \le k$.
+Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$.
 By induction, string $X$ is a prefix of $G.\gkey(v_{1})$.
@@
 % FIXME LF(i, c) picture from the slides for the earlier DBG example
-The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and \emph{de Bruijn graphs}.
+The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
 Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the paths starting from them.
@@
 The design of a path index is a trade-off between the maximum supported query length and the size of the index.
-In the worst case, a graph may have up to $(\sigma-2)^{k}$ distinct labels for paths of length $k$ (ignoring paths containing the source node or the sink node), and the length of possible paths is unlimited in cyclic graphs.
+In the worst case, a graph may have up to $\sigma^{k}$ distinct labels for paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and the length of possible paths is unlimited in cyclic graphs.
 GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary.
@@
 
+\subsection{Basic indexes}
+
+The \emph{\kmer{k} index} is the simplest path index.
+In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, W)$, where $X \in \Sigma^{k}$ and $W \subseteq V$.
+Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $W$ as matches if we find it.
+Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.
+
+Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns.
+Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order.
+Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, W_{i})$, where pattern $X$ is a prefix of the key $X_{i}$.
+We then list the union $\bigcup_{i=sp}^{ep} W_{i}$ as the set of matches.
+
+As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other.
+Before proceeding, we first have to define the de~Bruijn graph of a labeled graph.
+Let $P$ be a (finite or infinite) path in graph $G = (V, E)$.
+We say that path $P$ is \emph{maximal}, if starts at the source node or extends infinitely to the left, and ends at the sink node or extends infinitely to the right.
+We construct a collection of character sequences $\mathcal{S}$ by taking all maximal paths $P$ in graph $G$ and creating a sequence $S$ based on $G.\glabel(P)$ for each such path.
+If path $P$ is infinite in both directions, we use the label directly.
+If the path starts at the source node, we adjust the number of characters $\#$ at the beginning to match the order $k$ of the de~Bruijn graph.
+Similarly, if the path ends at the sink node, we adjust the number of characters $\$$ at the end.
+We can then generalize the definition of de~Bruijn graphs in Section~\ref{sect:graphs} to create a de~Bruijn graph of sequence collection $\mathcal{S}$.
+
+Given a \kmer{k+1} index of graph $G = (V, E)$, we can use it to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
+As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges.
+If we are interested in node $v_{X} \in V'$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$.
+Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.
+
+On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ directly as a \kmer{k} index of $G$.
+For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node.
+We can then search for patterns of any length in the de~Bruijn graph.
+As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} with short enough patterns.
+With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives.
+We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.
+
+\subsection{Path graphs}
+
 % FIXME from here
-k-mer indexes based on hash tables, sorted k-mers
-
-similarities between k-mer indexes and de Bruijn graphs: order-k+1 index can simulate order-k DBG; order-k DBG can be used as an order-k index
-
-define de Bruijn graphs for graphs (infinite strings, no source node, repeat sink node instead)
-
 using de Bruijn graphs as indexes; false positives
 - string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
-- add values (pointers to start nodes) to each node
 - valid for existence queries of length up to k+1 and locate queries of length up to k

### 2016-04-08 paper/paper.tex
Continuing with path graphs and rewriting old text.

@@
 \subjclass{E.1 Data Structures}
-\keywords{Dummy keyword -- please provide 1--5 keywords}% FIXME
+\keywords{Burrows-Wheeler transform, de Bruijn graphs, path indexes, variation graphs}
 % Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
+\newcommand{\kcollection}[1]{$#1$\nobreakdash-collection}
 \newcommand{\orderk}[1]{order\nobreakdash-$#1$}
@@
 
-\subsection{Strings}
+\subsection{Strings}\label{sect:strings}
 
@@
 A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$.
-We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively.
+We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s.
+We say that string $S'$ is a substring of string collection $\mathcal{S}$, if there is a string $S \in \mathcal{S}$ such that $S'$ is a substring of $S$.
+
+In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous subset of $\mathbb{Z}$.
+The notion of substrings generalizes for such infinite sequences in a natural way.
+We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
+A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.
 
@@
 The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
-We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$, where set $Z$ is one of $\mathbb{Z}$, $\set{x \in \mathbb{Z} \mid x \ge 0}$, and $\set{x \in \mathbb{Z} \mid x \le 0}$.
-The labels of such paths are infinite sequences of characters instead of strings.
+We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.
 
@@
 
+\begin{definition}[\kcollection{k}]
+Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value.
+We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if each sequence $S \in \mathcal{S}$
+\begin{enumerate}
+\item is left-infinite or begins with $\#^{k}$;
+\item is right-infinite or ends with $\$^{k}$; and
+\item contains no other occurrences of characters $\#$ and $\$$,
+\end{enumerate}
+and the collection contains substrings $\#^{k}$ and $\$^{k}$.
+\end{definition}
+
 % FIXME example of strings, DBG, long paths, false positives
-Let $\mathcal{S}$ be a collection of strings over alphabet $\Sigma$.
-We assume that each string $S \in \mathcal{S}$ begins with $k$ copies of $\#$ and ends with $k$ copies of $\$$, with no other occurrences of these characters in the string.
-The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$.
-Each node $v_{X}$ has a \emph{key} $G.\gkey(v_{X}) = X$ and a label $G.\glabel(v_{X}) = X[0]$.
-The edges of the de~Bruijn graph represent \kmer{k+1}s $Y$ occurring in the collection.
-For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$.
-We use the node corresponding to substring $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+% FIXME some comments about DBGs
+\begin{definition}[de~Bruijn graph]
+Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
+\begin{enumerate}
+\item each node $v_{X} \in V$ represent a distinct \kmer{k} occurring in $\mathcal{S}$;
+\item $G.\glabel(v_{X}) = X[0]$ and $G.\gkey(v_{X}) = X$ for all $v_{X} \in V$; and
+\item each edge $(v_{X}, v_{X'}) \in E$ represents a distinct \kmer{k+1} $Y = Xc = c'X'$ (where $c, c' \in \Sigma$) occurring in $\mathcal{S}$.
+\end{enumerate}
+We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+\end{definition}
 
 \begin{lemma}[No false negatives]\label{lemma:dbg-fn}
-Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ containing at most one $\$$.
+Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $X$ be a finite substring of $\mathcal{S}$ containing at most one $\$$.
 Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
@@
 \begin{proof}
-Let $X = S[i, i+\abs{X}-1]$ for a string $S \in \mathcal{S}$.
-For $i \le j \le i+\abs{X}-1$, there is node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$.
+Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$.
+For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$.
 By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$.
@@
 \begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
-Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$.
-Then the label of the path is a substring of a string $S \in \mathcal{S}$.
+Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$.
+Then the label of the path is a substring of a sequence $S \in \mathcal{S}$.
 \end{lemma}
@@
 \begin{proof}
-If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$.
+If $\abs{P} = 0$, the label is an empty string and hence a substring of each sequence $S \in \mathcal{S}$.
 If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.
@@
 
-As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other.
-Before proceeding, we first have to define the de~Bruijn graph of a labeled graph.
-Let $P$ be a (finite or infinite) path in graph $G = (V, E)$.
-We say that path $P$ is \emph{maximal}, if starts at the source node or extends infinitely to the left, and ends at the sink node or extends infinitely to the right.
-We construct a collection of character sequences $\mathcal{S}$ by taking all maximal paths $P$ in graph $G$ and creating a sequence $S$ based on $G.\glabel(P)$ for each such path.
-If path $P$ is infinite in both directions, we use the label directly.
-If the path starts at the source node, we adjust the number of characters $\#$ at the beginning to match the order $k$ of the de~Bruijn graph.
-Similarly, if the path ends at the sink node, we adjust the number of characters $\$$ at the end.
-We can then generalize the definition of de~Bruijn graphs in Section~\ref{sect:graphs} to create a de~Bruijn graph of sequence collection $\mathcal{S}$.
+As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. We now define the de~Bruijn graph of a graph for that purpose.
+
+\begin{definition}[Maximal paths]
+Let $G = (V, E)$ be a labeled graph.
+We say that path $P$ in the graph is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
+\end{definition}
+
+We can build a \kcollection{k} of maximal path labels by starting with the collection $\mathcal{S}$ of maximal path labels.
+In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.
+We also add string $\#^{k} \$^{k}$ into the collection if necessary.
+
+\begin{definition}[de~Bruijn graph of a graph]
+Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$.
+The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is also the \orderk{k} de~Bruijn graph of graph $G$.
+\end{definition}
 
@@
 
+De~Bruijn graphs are a special case of path graphs.
+They represent paths of length $k$ in the \emph{input graph} as nodes and paths overlapping on $k-1$ nodes of the input graph as edges.
+We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.
+
+\begin{definition}[Prefix-matching strings]
+Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
+\end{definition}
+
+\begin{definition}[Prefix-free set]
+Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
+\end{definition}
+
+\begin{definition}[Path graph]
+Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$.
+We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$.
+Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
+\begin{enumerate}
+\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$;
+\item $G'.\glabel(v_{K}) = K[0]$ and $G'.\gkey(v_{K}) = K$ for all $v_{K} \in V$; and
+\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
+\end{enumerate}
+We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+\end{definition}
+
 % FIXME from here
-using de Bruijn graphs as indexes; false positives
-- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
-- valid for existence queries of length up to k+1 and locate queries of length up to k
-
-Path graphs
-- a generalization of DBG as a path index
-- keys must be prefix-free
 - k-index: has a node matching any k-path in the input graph
@@
 pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
+
+using pruned de Bruijn graphs as indexes; false positives
+- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
+- valid for existence queries of length up to k+1 and locate queries of length up to k
 

### 2016-04-12 paper/paper.tex
Rewrote parts of the existing text.

@@
 
-\begin{abstract}% FIXME
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+\begin{abstract}
+Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes.
+Path indexes are one of the most important tools for working with variation graphs.
+They generalize text indexes for graphs, allowing one to find the paths matching the query string.
+We propose using pruned de Bruijn graphs as path indexes, encoding them space-efficiently with the Burrows-Wheeler transform.
+We also generalize many ideas from text indexing literature to work with graphs.
+The proposed approach has been implemented in the variation graph toolkit vg.
 \end{abstract}
@@
 
-Manifold DBG: \cite{Lin2014}
-
 Hypertext index: \cite{Thachuk2013}
@@
 
-In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous subset of $\mathbb{Z}$.
+In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
 The notion of substrings generalizes for such infinite sequences in a natural way.
@@
 
-\subsection{Graphs}\label{sect:graphs}
+The following definitions are useful when discussing text indexes and path indexes.
 
-A \emph{graph} $G = (V, E)$ consists of a set of \emph{nodes} $V = \set{0, \dotsc, \abs{V}-1}$ and a set of \emph{edges} $E \subseteq V \times V$.
-We say that $(u, v) \in E$ is an edge \emph{from} node $u$ \emph{to} node $v$, and assume that the edges are \emph{directed}: $(u, v) \ne (v, u)$ for $u \ne v$.
-The \emph{indegree} $G.\gindegree(v)$ of node $v$ is the number of \emph{incoming} edges to node $v$, while the \emph{outdegree} $G.\goutdegree(v)$ is the number of \emph{outgoing} edges from node $v$.
-
-The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$.
-A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$.
-The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
-We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.
-
-We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
-To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$.
-The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes.
-We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
-We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
-However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.
-
-\begin{definition}[\kcollection{k}]
-Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value.
-We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if each sequence $S \in \mathcal{S}$
-\begin{enumerate}
-\item is left-infinite or begins with $\#^{k}$;
-\item is right-infinite or ends with $\$^{k}$; and
-\item contains no other occurrences of characters $\#$ and $\$$,
-\end{enumerate}
-and the collection contains substrings $\#^{k}$ and $\$^{k}$.
+\begin{definition}[Prefix-matching strings]
+Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
 \end{definition}
 
-% FIXME example of strings, DBG, long paths, false positives
-% FIXME some comments about DBGs
-\begin{definition}[de~Bruijn graph]
-Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
-\begin{enumerate}
-\item each node $v_{X} \in V$ represent a distinct \kmer{k} occurring in $\mathcal{S}$;
-\item $G.\glabel(v_{X}) = X[0]$ and $G.\gkey(v_{X}) = X$ for all $v_{X} \in V$; and
-\item each edge $(v_{X}, v_{X'}) \in E$ represents a distinct \kmer{k+1} $Y = Xc = c'X'$ (where $c, c' \in \Sigma$) occurring in $\mathcal{S}$.
-\end{enumerate}
-We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+\begin{definition}[Prefix-free set]
+Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
 \end{definition}
-
-\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
-Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $X$ be a finite substring of $\mathcal{S}$ containing at most one $\$$.
-Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
-\end{lemma}
-
-\begin{proof}
-Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$.
-For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$.
-By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$.
-Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
-\end{proof}
-
-\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
-Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$.
-Then the label of the path is a substring of a sequence $S \in \mathcal{S}$.
-\end{lemma}
-
-\begin{proof}
-If $\abs{P} = 0$, the label is an empty string and hence a substring of each sequence $S \in \mathcal{S}$.
-If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.
-Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$.
-By induction, string $X$ is a prefix of $G.\gkey(v_{1})$.
-As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$.
-Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
-\end{proof}
 
@@
 
-The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text sharing the combinatorial structure of the suffix array.
-The BWT of text $T[0, n-1]$ is a string $\BWT[0, n-1]$ defined as $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
+The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array.
+Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
 Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$.
@@
 
+\subsection{Graphs}\label{sect:graphs}
+
+A \emph{graph} $G = (V, E)$ consists of a set of \emph{nodes} $V = \set{0, \dotsc, \abs{V}-1}$ and a set of \emph{edges} $E \subseteq V \times V$.
+We say that $(u, v) \in E$ is an edge \emph{from} node $u$ \emph{to} node $v$, and assume that the edges are \emph{directed}: $(u, v) \ne (v, u)$ for $u \ne v$.
+The \emph{indegree} $G.\gindegree(v)$ of node $v$ is the number of \emph{incoming} edges to $v$, while the \emph{outdegree} $G.\goutdegree(v)$ is the number of \emph{outgoing} edges from $v$.
+
+The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$.
+A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$.
+We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path.
+The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
+We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.
+
+We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
+To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$.
+The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes.
+We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
+We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
+However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.
+
+In this paper, we work with de~Bruijn graphs and their generalizations.
+In order to do so, we need to define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.
+
+\begin{definition}[\kcollection{k}]
+Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value.
+We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if the collection contains substrings $\#^{k}$ and $\$^{k}$, and each sequence $S \in \mathcal{S}$
+\begin{enumerate}
+\item is left-infinite or begins with $\#^{k}$;
+\item is right-infinite or ends with $\$^{k}$; and
+\item contains no other occurrences of characters $\#$ and $\$$.
+\end{enumerate}
+\end{definition}
+
+% FIXME example of strings, DBG, long paths, false positives
+\begin{definition}[de~Bruijn graph]
+Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
+\begin{enumerate}
+\item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
+\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
+\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
+\end{enumerate}
+We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+\end{definition}
+
+The following lemmas summarize the key properties of de~Bruijn graphs for indexing purposes.
+
+\begin{lemma}[Keys prefix-match with path labels]\label{lemma:dbg-key}
+Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph, and let $P$ be a path starting from node $v_{0} \in V$. Then strings $G.\gkey(v_{0})$ and $G.\glabel(P)$ prefix-match.
+\end{lemma}
+
+\begin{proof}
+Let $P[0, k'-1] = v_{0} \dotsm v_{k'-1}$ be a prefix of path $P$, where $1 \le k' \le \min(k, \abs{P})$.
+If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.
+Now assume that $k' > 1$.
+By induction, we know that $G.\glabel(P[1, k'-1])$ is a prefix of $G.\gkey(v_{1})$.
+As edge $(v_{i}, v_{i+1}) \in E$ exists, $G.\glabel(v_{0}) \cdot G.\gkey(v_{1}) = G.\gkey(v_{0}) \cdot c$ (where $c \in \Sigma$), and hence $G.\glabel(P[0, k'-1])$ is a prefix of $G.\gkey(v_{0})$.
+As the first $\min(k, \abs{P})$ characters of $G.\gkey(v_{0})$ and $G.\glabel(P)$ are identical, the result follows.
+\end{proof}
+
+\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
+Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $X$ be a finite substring of $\mathcal{S}$ containing at most one $\$$.
+Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
+\end{lemma}
+
+\begin{proof}
+Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$.
+For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$.
+By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$.
+Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
+\end{proof}
+
+\begin{lemma}[No false positives with $\abs{P} \le k$]\label{lemma:dbg-fp}
+Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P = v_{0} \dotsm v_{\abs{P}-1}$ be a path in $G$ with $\abs{P} \le k$.
+Then the label of the path is a prefix of $G.\gkey(v_{0})$ and hence a substring of a sequence $S \in \mathcal{S}$.
+\end{lemma}
+
+\begin{proof}
+The label $G.\glabel(P)$ prefix-matches $G.\gkey(v_{0})$ by Lemma~\ref{lemma:dbg-key}. As $\abs{P} \le k$, string $G.\glabel(P)$ is a prefix of $G.\gkey(v_{0})$
+\end{proof}
+
+Note that the labels of all paths of length $k+1$ are also substrings of collection $\mathcal{S}$.
+
 \subsection{Generalized indexes}
@@
 A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}.
-Given a path index for graph $G = (V, E)$, we want to use the index to find the \emph{start nodes} $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
+Given a path index for graph $G = (V, E)$, we want to use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
 
@@
 
-We can build a \kcollection{k} of maximal path labels by starting with the collection $\mathcal{S}$ of maximal path labels.
+We can build a \kcollection{k} based on the labels of maximal paths by starting with the collection $\mathcal{S}$ of such labels.
 In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.
-We also add string $\#^{k} \$^{k}$ into the collection if necessary.
 
@@
 Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$.
-The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is also the \orderk{k} de~Bruijn graph of graph $G$.
+Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
 \end{definition}
@@
 
-On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ directly as a \kmer{k} index of $G$.
+On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length.
 For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node.
-We can then search for patterns of any length in the de~Bruijn graph.
-As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} with short enough patterns.
+
+\begin{definition}[de~Bruijn graph as an index]
+Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be its \orderk{k} de~Bruijn graph.
+Pattern $X \in \Sigma^{\ast}$ matches node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$.
+The set of occurrences for pattern $X$ is the union of sets $G.\gvalue(v)$ for the matching nodes $v \in V'$.
+\end{definition}
+
+As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$.
 With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives.
@@
 
-\begin{definition}[Prefix-matching strings]
-Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
-\end{definition}
-
-\begin{definition}[Prefix-free set]
-Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
-\end{definition}
-
 \begin{definition}[Path graph]
@@
 \begin{enumerate}
-\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$;
-\item $G'.\glabel(v_{K}) = K[0]$ and $G'.\gkey(v_{K}) = K$ for all $v_{K} \in V$; and
+\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
+\item each node has a key $G'.\gkey(v_{K}) = K$; and
 \item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
@@
 % FIXME from here
+- using path graphs as indexes
 - k-index: has a node matching any k-path in the input graph
@@
 pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
-
-using pruned de Bruijn graphs as indexes; false positives
-- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
-- valid for existence queries of length up to k+1 and locate queries of length up to k
+similar to manifold de Bruijn graphs \cite{Lin2014}
 
@@
 
-I want to thank \dots
+I thank Richard Durbin, Erik Garrison, and Adam Novak for \dots
 

### 2016-04-12 paper/paper.tex
Almost finished with pruned de Bruijn graphs.

@@
 \newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
+\newcommand{\gnode}{\ensuremath{\mathsf{node}}}
 
@@
 The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
-We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.
+
+We generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
+We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
 
@@
 If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.
+
 Now assume that $k' > 1$.
@@
 The \emph{\kmer{k} index} is the simplest path index.
-In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, W)$, where $X \in \Sigma^{k}$ and $W \subseteq V$.
-Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $W$ as matches if we find it.
+In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ and $V_{X} \subseteq V$.
+Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $V_{X}$ as matches if we find it.
 Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.
@@
 Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order.
-Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, W_{i})$, where pattern $X$ is a prefix of the key $X_{i}$.
-We then list the union $\bigcup_{i=sp}^{ep} W_{i}$ as the set of matches.
+Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{X_{i}})$, where pattern $X$ is a prefix of the key $X_{i}$.
+We then list the union $\bigcup_{i=sp}^{ep} V_{X_{i}}$ as the set of matches.
 
-As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. We now define the de~Bruijn graph of a graph for that purpose.
-
-\begin{definition}[Maximal paths]
-Let $G = (V, E)$ be a labeled graph.
-We say that path $P$ in the graph is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
-\end{definition}
-
-We can build a \kcollection{k} based on the labels of maximal paths by starting with the collection $\mathcal{S}$ of such labels.
-In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.
+As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other.
+In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$.
+If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$.
+We then transform $\mathcal{S}$ into a \kcollection{k} by adjusting the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.
+If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = (s, j)$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
 
 \begin{definition}[de~Bruijn graph of a graph]
-Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$.
+Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$.
 Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
@@
 On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length.
-For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node.
-
-\begin{definition}[de~Bruijn graph as an index]
-Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be its \orderk{k} de~Bruijn graph.
-Pattern $X \in \Sigma^{\ast}$ matches node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$.
-The set of occurrences for pattern $X$ is the union of sets $G.\gvalue(v)$ for the matching nodes $v \in V'$.
-\end{definition}
-
+For each key-value pair $(X, V_{X})$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = V_{X}$ to the node.
+We say that a pattern matches node $v \in V'$ if it prefix-matches string $G.\gkey(v)$.
 As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$.
@@
 De~Bruijn graphs are a special case of path graphs.
-They represent paths of length $k$ in the \emph{input graph} as nodes and paths overlapping on $k-1$ nodes of the input graph as edges.
+They represent the paths of length $k$ in the \emph{input graph} as nodes and the pairs of paths overlapping on $k-1$ nodes of the input graph as edges.
 We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.
@@
 \item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
-\item each node has a key $G'.\gkey(v_{K}) = K$; and
+\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i)$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
 \item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
@@
 
-% FIXME from here
-- using path graphs as indexes
-- k-index: has a node matching any k-path in the input graph
-- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
-- k-equivalent: union of values for nodes matching X identical for all |X| <= k
-- k-equivalent + edge condition: no false positives of length <= k; no false negatives
+\begin{definition}[Path graph as an index]
+Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$.
+Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$.
+If $V_{X}$ is the set of nodes $v \in V'$ matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\gvalue(X) = G'.\gvalue(V_{X}) = \bigcup_{v \in V_{X}} G'.\gvalue(v)$.
+\end{definition}
 
-pruned de Bruijn graphs: merge nodes of DBG while keeping the path graph k-equivalent
-similar to manifold de Bruijn graphs \cite{Lin2014}
+In the general case, queries using a path graph may result in false positives for patterns $X$ of length $\abs{X} > \abs{K}$, where $K$ is the shortest key in the key set $\mathcal{K}$.
+To avoid this, we restrict our attention to path graphs that are equivalent to de~Bruijn graphs.
+
+\begin{definition}[Equivalent path graphs]
+Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs. We say that graphs $G$ and $G'$ are \emph{equivalent}, if we have $G.\gvalue(X) = G'.\gvalue(X)$ for all patterns $X \in \Sigma^{\ast}$.
+\end{definition}
+
+\begin{definition}[Pruned de~Bruijn graph]
+Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$.
+Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is equivalent to the \orderk{k} de~Bruijn graph of $G$, and $\abs{K} \le k$ for all $K \in \mathcal{K}$.
+\end{definition}
+
+Pruned de~Bruijn graphs can be built by merging nodes of a de~Bruijn graph. They are similar to manifold de Bruijn graphs \cite{Lin2014}.
+
+\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
+If $\abs{V_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$ after we remove $G'.\gkey(v)$ for all $v \in V_{K}$ from the key set and replace them with $K$.
+\end{lemma}
+
+\begin{proof}
+Let $X \in \Sigma^{\ast}$ be a pattern, and let $V_{X} \subseteq V'$ be the set of nodes matching it.
+If the pattern does not prefix-match string $K$, we have $V_{X} \cap V_{K} = \emptyset$, and the changes in the key set do not affect the set of occurrences for pattern $X$.
+
+Otherwise let $G'' = (V'', E'')$ be the path graph with adjusted key set, and let $v_{K} \in V''$ be the node corresponding to key $K$. Then
+\begin{align*}
+G''.\gvalue(X) & = G''.\gvalue(V_{X} \setminus V_{K}) \cup G''.\gvalue(v_{K}) \\
+ & = G'.\gvalue(V_{X} \setminus V_{K}) \cup G'.\gvalue(V_{X} \cap V_{K}) = G'.\gvalue(X).
+\end{align*}
+\end{proof}
+
+\begin{lemma}[Correctness]\label{lemma:dbl-correct}
+Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$.
+For every pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$, let $V_{X}$ be the set of start nodes of paths $P$ in graph $G$ with $G.\glabel(P) = X$.
+Then $V_{X} \subseteq G'.\gvalue(X)$, and
+$V_{X} = G'.\gvalue(X)$ if $\abs{X} \le k$.
+\end{lemma}
+
+% FIXME
+% Introduce the de Bruijn graph
+% Set of nodes of DBG matching the pattern -> set of kmers in S prefix-matching the pattern
+% -> set of start nodes in G for paths matching the first min(k, |X|) characters of the pattern.
+% If |X| > k, this may be a superset of start nodes matching the entire pattern.
+% Pruned DBG equivalent to DBG -> holds with G' as well.
+\begin{proof}
+Foo
+\end{proof}
 

### 2016-04-13 paper/paper.tex
Continued mutilating the definitions.

@@
 \newcommand{\LF}{\ensuremath{\mathsf{LF}}}
+\newcommand{\find}{\ensuremath{\mathsf{find}}}
+\newcommand{\locate}{\ensuremath{\mathsf{locate}}}
 
@@
 \newcommand{\kcollection}[1]{$#1$\nobreakdash-collection}
+\newcommand{\kequivalent}[1]{$#1$\nobreakdash-equivalent}
 \newcommand{\orderk}[1]{order\nobreakdash-$#1$}
@@
 \begin{definition}[\kcollection{k}]
-Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value.
+Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k > 0$ be a parameter value.
 We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if the collection contains substrings $\#^{k}$ and $\$^{k}$, and each sequence $S \in \mathcal{S}$
@@
 
-The following lemmas summarize the key properties of de~Bruijn graphs for indexing purposes.
-
-\begin{lemma}[Keys prefix-match with path labels]\label{lemma:dbg-key}
-Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph, and let $P$ be a path starting from node $v_{0} \in V$. Then strings $G.\gkey(v_{0})$ and $G.\glabel(P)$ prefix-match.
-\end{lemma}
-
-\begin{proof}
-Let $P[0, k'-1] = v_{0} \dotsm v_{k'-1}$ be a prefix of path $P$, where $1 \le k' \le \min(k, \abs{P})$.
-If $k' = 1$, we know that $G.\glabel(P[0]) = G.\glabel(v_{0})$ is a prefix of $G.\gkey(v_{0})$.
-
-Now assume that $k' > 1$.
-By induction, we know that $G.\glabel(P[1, k'-1])$ is a prefix of $G.\gkey(v_{1})$.
-As edge $(v_{i}, v_{i+1}) \in E$ exists, $G.\glabel(v_{0}) \cdot G.\gkey(v_{1}) = G.\gkey(v_{0}) \cdot c$ (where $c \in \Sigma$), and hence $G.\glabel(P[0, k'-1])$ is a prefix of $G.\gkey(v_{0})$.
-As the first $\min(k, \abs{P})$ characters of $G.\gkey(v_{0})$ and $G.\glabel(P)$ are identical, the result follows.
-\end{proof}
-
-\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
-Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $X$ be a finite substring of $\mathcal{S}$ containing at most one $\$$.
-Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
-\end{lemma}
-
-\begin{proof}
-Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$.
-For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$.
-By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$.
-Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
-\end{proof}
-
-\begin{lemma}[No false positives with $\abs{P} \le k$]\label{lemma:dbg-fp}
-Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P = v_{0} \dotsm v_{\abs{P}-1}$ be a path in $G$ with $\abs{P} \le k$.
-Then the label of the path is a prefix of $G.\gkey(v_{0})$ and hence a substring of a sequence $S \in \mathcal{S}$.
-\end{lemma}
-
-\begin{proof}
-The label $G.\glabel(P)$ prefix-matches $G.\gkey(v_{0})$ by Lemma~\ref{lemma:dbg-key}. As $\abs{P} \le k$, string $G.\glabel(P)$ is a prefix of $G.\gkey(v_{0})$
-\end{proof}
-
-Note that the labels of all paths of length $k+1$ are also substrings of collection $\mathcal{S}$.
+De~Bruijn graphs have several properties that make them useful for indexing purposes.
+Node keys are prefixes of the labels of all paths of length at least $k$ starting from the node.
+This makes it possible to sort the nodes unambiguosly by path labels.
+Every substring of the \kcollection{k} is the label of a path in the de~Bruijn graph, and every path label of length at most $k+1$ is a substring of the collection.
+In Section~\ref{sect:path-indexes}, we develop an index structure based on a generalization of de~Bruijn graphs.
 
@@
 The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
-Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the paths starting from them.
-The transformation is expensive and may increase the size of the graph exponentially.
-After sorting the nodes of the transformed graph, we encode them using three sequences.
+Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them.
+In the worst case, the transformation increases the size of the graph exponentially.
+After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences.
 $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform.
@@
 
-\section{Path indexes}
+\section{Path indexes}\label{sect:path-indexes}
 
 A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}.
-Given a path index for graph $G = (V, E)$, we want to use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
+Given a path index for \emph{input graph} $G = (V, E)$, we want to use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
 
@@
 Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order.
-Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{X_{i}})$, where pattern $X$ is a prefix of the key $X_{i}$.
-We then list the union $\bigcup_{i=sp}^{ep} V_{X_{i}}$ as the set of matches.
+Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where pattern $X$ is a prefix of the key $X_{i}$.
+We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.
 
@@
 If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$.
-We then transform $\mathcal{S}$ into a \kcollection{k} by adjusting the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence.
+We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence.
 If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = (s, j)$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
+Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
 
@@
 Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$.
-Then the \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
+The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
 \end{definition}
 
-Given a \kmer{k+1} index of graph $G = (V, E)$, we can use it to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
+We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
 As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges.
-If we are interested in node $v_{X} \in V'$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$.
+If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$.
 Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.
@@
 On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length.
-For each key-value pair $(X, V_{X})$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = V_{X}$ to the node.
-We say that a pattern matches node $v \in V'$ if it prefix-matches string $G.\gkey(v)$.
-As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} (path labels that exist in the de~Bruijn graph but not in the input graph) with patterns of length at most $k$.
-With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives.
-We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.
+Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph.
+For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}:
+$$
+G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
+$$
+Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$.
+Due to the nature of de~Bruijn graphs, this index will not produce any false negatives.
+There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.
 
@@
 \begin{definition}[Path graph]
-Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$.
+Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings of length at most $k$ containing $\#^{k}$ and $\$^{k}$.
 We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$.
-Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
+The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
 \begin{enumerate}
 \item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
-\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i)$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
+\item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
 \item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
@@
 \begin{definition}[Path graph as an index]
-Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$.
-Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$ if it prefix-matches string $G'.\gkey(v)$.
-If $V_{X}$ is the set of nodes $v \in V'$ matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\gvalue(X) = G'.\gvalue(V_{X}) = \bigcup_{v \in V_{X}} G'.\gvalue(v)$.
+Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
+\begin{itemize}
+\item Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$, if there is a path $P$ in $G'$ with $G'.\glabel(P) = X$.
+We use $G'.\find(X)$ to denote the set of nodes $V'_{X} \subseteq V'$ matching the pattern.
+\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = G'.\gvalue(V'_{X}) = \bigcup_{v \in V'_{X}} G'.\gvalue(v)$.
+We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
+\end{itemize}
 \end{definition}
 
-In the general case, queries using a path graph may result in false positives for patterns $X$ of length $\abs{X} > \abs{K}$, where $K$ is the shortest key in the key set $\mathcal{K}$.
-To avoid this, we restrict our attention to path graphs that are equivalent to de~Bruijn graphs.
+\begin{lemma}[No false negatives]\label{lemma:pg-fn}
+Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern string.
+Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
+\end{lemma}
+
+\begin{proof}
+Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
+The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
+
+For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
+By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
+Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
+As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
+\end{proof}
+
+When used as an index, a path graph will not produce false negatives.
+On the other hand, we may encounter false positives already at path length $k'+1$, where $k'$ is the length of the shortest key in the key set.
+This is not very convenient for a path index.
+In the next section, we consider a class of path graphs that can be proven to be equivalent to de~Bruijn graphs.
+
+\subsection{Pruned de Bruijn graphs}
+
+De~Bruijn graphs often have nodes that are redundant when the graph is used as a path index.
+Removing these nodes from the graph would make the index smaller and more practical.
+On the other hand, we still want to guarantee that the index works correctly with patterns of length $k$ or less, where $k$ is the order of the de~Bruijn graph.
 
 \begin{definition}[Equivalent path graphs]
-Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs. We say that graphs $G$ and $G'$ are \emph{equivalent}, if we have $G.\gvalue(X) = G'.\gvalue(X)$ for all patterns $X \in \Sigma^{\ast}$.
+Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value.
+We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $\abs{X} \le k$.
 \end{definition}
@@
 \begin{definition}[Pruned de~Bruijn graph]
-Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be the \orderk{k} path graph of $G$ with key set $\mathcal{K}$.
-Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is equivalent to the \orderk{k} de~Bruijn graph of $G$, and $\abs{K} \le k$ for all $K \in \mathcal{K}$.
+Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} path graph of $G$.
+Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
 \end{definition}
@@
 \begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
-If $\abs{V_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$ after we remove $G'.\gkey(v)$ for all $v \in V_{K}$ from the key set and replace them with $K$.
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
+If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
 \end{lemma}
@@
 \begin{proof}
-Let $X \in \Sigma^{\ast}$ be a pattern, and let $V_{X} \subseteq V'$ be the set of nodes matching it.
-If the pattern does not prefix-match string $K$, we have $V_{X} \cap V_{K} = \emptyset$, and the changes in the key set do not affect the set of occurrences for pattern $X$.
+Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $\abs{X} \le k$, let $G'' = (V'', E'')$ be the path graph after the changes in the key set, and let $v''_{K} \in V''$ be the node with $G''.\gkey(v''_{K}) = K$.
 
-Otherwise let $G'' = (V'', E'')$ be the path graph with adjusted key set, and let $v_{K} \in V''$ be the node corresponding to key $K$. Then
-\begin{align*}
-G''.\gvalue(X) & = G''.\gvalue(V_{X} \setminus V_{K}) \cup G''.\gvalue(v_{K}) \\
- & = G'.\gvalue(V_{X} \setminus V_{K}) \cup G'.\gvalue(V_{X} \cap V_{K}) = G'.\gvalue(X).
-\end{align*}
+Each node $v \in V' \setminus V'_{K}$ is also a node of graph $G''$ with $G''.\gvalue(v) = G'.\gvalue(v)$.
+As each occurrence of substring $K$ in the \kcollection{k} $\mathcal{S}$ can be extended into an occurrence of substring $G'.\gkey(v)$ for a node $v \in V'_{K}$, we also have $G''.\gvalue(v''_{K}) = G'.\gvalue(v)$ for all $v \in V'_{K}$.
+
+We now prove that for every path matching the pattern in graph $G'$, we can find a matching path in graph $G''$ with the same set of occurrences, and the other way around.
+As graph $G'$ is a pruned de~Bruijn graph and we prove that graph $G''$ is \kequivalent{k} to it, graph $G''$ must also be a pruned de~Bruijn graph.
+
+(From $G'$ to $G''$) Consider a path $P' = v'_{0} \dotsm v'_{\abs{X}-1}$ matching the pattern in graph $G'$.
+Let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be the sequence of nodes of graph $G''$ created by replacing every node $v \in V'_{K}$ in path $P$ with node $v''_{K}$.
+We note that $G''.\gkey(v''_{i})$ is a prefix of $G'.\gkey(v'_{i})$ for $0 \le i < \abs{X}$, and hence $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$ is a prefix of $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$ for $0 \le i < \abs{X}-1$.
+As key $G'.\gkey(v'_{i})$ prefix-matches $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$, key $G''.\gkey(v''_{i})$ must also prefix-match $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$, giving us an edge $(v''_{i}, v''_{i+1}) \in E''$.
+Hence sequence $P''$ is a path in $G''$, with $G''.\glabel(P'') = G'.\glabel(P')$ and $G''.\gvalue(v''_{0}) = G'.\gvalue(v'_{0})$.
+
+% FIXME the same for P'' -> P'
+% by contradiction: assume that G'' contains X but G' does not
+% let X[i, ...] be the rightmost suffix not in G'
+% X[i, ...] does not occur in the k-collection but X[i+1, ...] does
+% X[i] G''.key(i+1) does occur, must be a proper prefix of X[i, ...]
+% ...
+(From $G''$ to $G'$)
 \end{proof}
 
-\begin{lemma}[Correctness]\label{lemma:dbl-correct}
-Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$.
-For every pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$, let $V_{X}$ be the set of start nodes of paths $P$ in graph $G$ with $G.\glabel(P) = X$.
-Then $V_{X} \subseteq G'.\gvalue(X)$, and
-$V_{X} = G'.\gvalue(X)$ if $\abs{X} \le k$.
+% Pruning generalizes the graph in the same way as DBG generalizes it?
+
+\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P$ be a path in $G'$ with $\abs{P'} \le k$.
+Then there is a path $P$ in graph $G$ with $G.\glabel(P) = G'.\glabel(P')$.
 \end{lemma}
 
-% FIXME
-% Introduce the de Bruijn graph
-% Set of nodes of DBG matching the pattern -> set of kmers in S prefix-matching the pattern
-% -> set of start nodes in G for paths matching the first min(k, |X|) characters of the pattern.
-% If |X| > k, this may be a superset of start nodes matching the entire pattern.
-% Pruned DBG equivalent to DBG -> holds with G' as well.
+% FIXME proof
 \begin{proof}
@@
 
-\section{Construction}
+\subsection{GCSA for path indexes}
+
+\subsection{Index construction}
+
+\subsection{Extensions}
 

### 2016-04-14 paper/paper.tex
The lemmas seem to work. Finally.

@@
 Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value.
-We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $\abs{X} \le k$.
+We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
 \end{definition}
@@
 
-Pruned de~Bruijn graphs can be built by merging nodes of a de~Bruijn graph. They are similar to manifold de Bruijn graphs \cite{Lin2014}.
+Pruned de~Bruijn graphs are similar to manifold de Bruijn graphs \cite{Lin2014}.
 
-\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
-If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u) = G'.\gvalue(v)$ for all $u, v \in V_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
+\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P' = v'_{0} \dotsm v'_{\abs{P}-1}$ be a path in graph $G'$ with $1 \le \abs{P'} \le k$ and $G'.\glabel(P') \in (\Sigma \setminus \set{\#, \$})^{\ast}$.
+Then there is a path $P$ in $G$ with $G.\glabel(P) = G'.\glabel(P')$.
+If $G'$ is a de~Bruijn graph, $G'.\glabel(P)$ is a prefix of $G'.\gkey(v'_{0})$.
 \end{lemma}
@@
 \begin{proof}
-Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $\abs{X} \le k$, let $G'' = (V'', E'')$ be the path graph after the changes in the key set, and let $v''_{K} \in V''$ be the node with $G''.\gkey(v''_{K}) = K$.
+We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
 
-Each node $v \in V' \setminus V'_{K}$ is also a node of graph $G''$ with $G''.\gvalue(v) = G'.\gvalue(v)$.
-As each occurrence of substring $K$ in the \kcollection{k} $\mathcal{S}$ can be extended into an occurrence of substring $G'.\gkey(v)$ for a node $v \in V'_{K}$, we also have $G''.\gvalue(v''_{K}) = G'.\gvalue(v)$ for all $v \in V'_{K}$.
-
-We now prove that for every path matching the pattern in graph $G'$, we can find a matching path in graph $G''$ with the same set of occurrences, and the other way around.
-As graph $G'$ is a pruned de~Bruijn graph and we prove that graph $G''$ is \kequivalent{k} to it, graph $G''$ must also be a pruned de~Bruijn graph.
-
-(From $G'$ to $G''$) Consider a path $P' = v'_{0} \dotsm v'_{\abs{X}-1}$ matching the pattern in graph $G'$.
-Let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be the sequence of nodes of graph $G''$ created by replacing every node $v \in V'_{K}$ in path $P$ with node $v''_{K}$.
-We note that $G''.\gkey(v''_{i})$ is a prefix of $G'.\gkey(v'_{i})$ for $0 \le i < \abs{X}$, and hence $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$ is a prefix of $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$ for $0 \le i < \abs{X}-1$.
-As key $G'.\gkey(v'_{i})$ prefix-matches $G'.\glabel(v'_{i}) \cdot G'.\gkey(v'_{i+1})$, key $G''.\gkey(v''_{i})$ must also prefix-match $G''.\glabel(v''_{i}) \cdot G''.\gkey(v''_{i+1})$, giving us an edge $(v''_{i}, v''_{i+1}) \in E''$.
-Hence sequence $P''$ is a path in $G''$, with $G''.\glabel(P'') = G'.\glabel(P')$ and $G''.\gvalue(v''_{0}) = G'.\gvalue(v'_{0})$.
-
-% FIXME the same for P'' -> P'
-% by contradiction: assume that G'' contains X but G' does not
-% let X[i, ...] be the rightmost suffix not in G'
-% X[i, ...] does not occur in the k-collection but X[i+1, ...] does
-% X[i] G''.key(i+1) does occur, must be a proper prefix of X[i, ...]
-% ...
-(From $G''$ to $G'$)
+If $\abs{P} = 1$, every node $v \in G'.\gvalue(v'_{0})$ is a path in graph $G$ with label $X = G'.\glabel(P')$.
+Otherwise every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$.
+Furthermore, the substring has $G'.\gkey(v'_{i})$ as its prefix.
+We can show by induction from $\abs{P}-2$ to $0$ that $S_{0}[j, j+k]$ has $X$ as its prefix.
+The sequence of nodes $P = \mathcal{S}.\gnode(S_{0}, j+i) \dotsm \mathcal{S}.\gnode(S_{0}, j+\abs{X}-1)$ is a path in graph $G$ with $G.\glabel(P) = X$.
 \end{proof}
 
-% Pruning generalizes the graph in the same way as DBG generalizes it?
-
-\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P$ be a path in $G'$ with $\abs{P'} \le k$.
-Then there is a path $P$ in graph $G$ with $G.\glabel(P) = G'.\glabel(P')$.
+\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
+If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
 \end{lemma}
 
-% FIXME proof
 \begin{proof}
-Foo
+Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys.
+For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.
+
+Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
+By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
+We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
+The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
+If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
+
+Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
+If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
+Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, there must be node $u \in V$ with $G.\glabel(u) = X[j]$ and with an edge $(u, v) \in E$ to a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
+Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
+As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
+When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
+
+Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
+We can also start from path $P''$ and determine the node $v_{0}$.
+As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
 \end{proof}
+
+De~Bruijn graphs are a special case of pruned de~Bruijn graphs.
+Hence we may compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
 

### 2016-04-15 paper/paper.tex
Maximally pruned DBGs are needed for ST functionality

@@
 \newcommand{\glabel}{\ensuremath{\mathsf{label}}}
+\newcommand{\gpred}{\ensuremath{\mathsf{pred}}}
 \newcommand{\gkey}{\ensuremath{\mathsf{key}}}
@@
 
+Given a graph $G = (V, E)$, we may want to reason about the \emph{predecessors} of a node with the given label.
+Let $v \in V$ be a node and $c \in \Sigma$ be a character.
+We write $G.\gpred(v, c)$ to denote the set of nodes $u \in V$ such that $G.\glabel(u) = c$ and there is an edge $(u, v) \in E$.
+
 In this paper, we work with de~Bruijn graphs and their generalizations.
@@
 \begin{definition}[de~Bruijn graph of a graph]
-Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$.
-The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
+Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
 \end{definition}
@@
 \begin{definition}[Equivalent path graphs]
-Let $G = (V, E)$ and $G' = (V', E')$ be two path graphs, and let $k > 0$ be a parameter value.
+Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value.
 We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
@@
 \begin{definition}[Pruned de~Bruijn graph]
-Let $G = (V, G)$ be a graph, and let $G' = (V', E')$ be an \orderk{k} path graph of $G$.
-Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
+Let $G$ be a graph, and let $G'$ be an \orderk{k} path graph of $G$. Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
 \end{definition}
@@
 Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, there must be node $u \in V$ with $G.\glabel(u) = X[j]$ and with an edge $(u, v) \in E$ to a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
 Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
@@
 Hence we may compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
+
+\begin{definition}[Maximally pruned de~Bruijn graph]
+Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$.
+We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune}, and the keys in $\mathcal{K}$ cannot be made shorter without changing the graph structure.
+\end{definition}
+
+\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
+Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph.
+Then (a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and (b) the key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
+\end{lemma}
+
+\begin{proof}
+(a) Let $v' \in V'$ be a node, and let $c \in \Sigma$ be a character.
+For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
+If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct.
+In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match.
+Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge set $G'.\gpred(v', c)$ into a single node.
+This contradicts the assumption that graph $G'$ is maximally pruned, so it must be that $\abs{G'.\gpred(v', c)} \le 1$.
+
+(b) If $\abs{X} = 0$, every key prefix-matches the pattern.
+If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
+Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
+By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
+As the keys are as short as possible, key $G'.\gkey(u')$ must be a prefix of $X[0] \cdot G'.\gkey(v')$.
+Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
+\end{proof}
 

### 2016-04-15 paper/paper.tex
Clarified lemma 14 a bit.

@@
 We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if the collection contains substrings $\#^{k}$ and $\$^{k}$, and each sequence $S \in \mathcal{S}$
-\begin{enumerate}
-\item is left-infinite or begins with $\#^{k}$;
-\item is right-infinite or ends with $\$^{k}$; and
-\item contains no other occurrences of characters $\#$ and $\$$.
-\end{enumerate}
+(a) is left-infinite or begins with $\#^{k}$;
+(b) is right-infinite or ends with $\$^{k}$; and
+(c) contains no other occurrences of characters $\#$ and $\$$.
 \end{definition}
@@
 Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
-\begin{enumerate}
+\begin{itemize}
 \item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
@@
 \item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
-\end{enumerate}
+\end{itemize}
 We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
@@
 The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
-\begin{enumerate}
+\begin{itemize}
 \item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
@@
 \item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
-\end{enumerate}
+\end{itemize}
 We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
@@
 \begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
-Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph.
-Then (a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and (b) the key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
+Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
+(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
+(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
+(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
 \end{lemma}
@@
 \begin{proof}
-(a) Let $v' \in V'$ be a node, and let $c \in \Sigma$ be a character.
-For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
+(a) For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
 If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct.
 In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match.
-Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge set $G'.\gpred(v', c)$ into a single node.
-This contradicts the assumption that graph $G'$ is maximally pruned, so it must be that $\abs{G'.\gpred(v', c)} \le 1$.
+Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
+Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
 
-(b) If $\abs{X} = 0$, every key prefix-matches the pattern.
+(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
+Key $G'.\gkey(u')$ must have $c \cdot G'.\gkey(v')$ as a prefix.
+If there is a substring $S[i,j] = c \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', c)$.
+By (a), $u'$ is the only such node, and hence string $c \cdot G'.\gkey(v')$ can also serve as its key.
+
+(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
 If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
@@
 By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
-As the keys are as short as possible, key $G'.\gkey(u')$ must be a prefix of $X[0] \cdot G'.\gkey(v')$.
+By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
 Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.

### 2016-04-15 paper/paper.tex
Started section 4; extended one of the lemmas.

@@
 \newcommand{\FMindex}{FM\nobreakdash-index}
+\newcommand{\patternset}{\ensuremath{(\Sigma \setminus \set{\#, \$})^{\ast}}}
 
@@
 FM-index: \cite{Ferragina2005a}
-
-CST-NPR: \cite{Fischer2009a}
-
-CST MEMs: \cite{Ohlebusch2010a}
 
@@
 $$
-where $\Carray[c]$ is the number of suffixes of the text starting with any character $c' < c$.
-Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$.
-We can also generalize the definition for any character $c \in \Sigma$:
+where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
 $$
@@
 \begin{lemma}[No false negatives]\label{lemma:pg-fn}
-Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern string.
+Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in \patternset$ be a pattern string.
 Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
@@
 Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value.
-We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ with $1 \le \abs{X} \le k$.
+We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
 \end{definition}
@@
 \begin{lemma}[No short false positives]\label{lemma:dbg-fp}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $P' = v'_{0} \dotsm v'_{\abs{P}-1}$ be a path in graph $G'$ with $1 \le \abs{P'} \le k$ and $G'.\glabel(P') \in (\Sigma \setminus \set{\#, \$})^{\ast}$.
-Then there is a path $P$ in $G$ with $G.\glabel(P) = G'.\glabel(P')$.
-If $G'$ is a de~Bruijn graph, $G'.\glabel(P)$ is a prefix of $G'.\gkey(v'_{0})$.
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$. Then
+(a) $G'.\locate(X)$ is a set of start nodes $v_{0} \in V$ of paths $P$ matching the pattern in graph $G$; and
+(b) if $G'$ is a de~Bruijn graph, $X$ is prefix of $G'.\gkey(v'_{0})$ for all $v'_{0} \in G'.\find(X)$.
 \end{lemma}
@@
 \begin{proof}
-We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
+If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
 
-If $\abs{P} = 1$, every node $v \in G'.\gvalue(v'_{0})$ is a path in graph $G$ with label $X = G'.\glabel(P')$.
-Otherwise every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$.
-Furthermore, the substring has $G'.\gkey(v'_{i})$ as its prefix.
-We can show by induction from $\abs{P}-2$ to $0$ that $S_{0}[j, j+k]$ has $X$ as its prefix.
-The sequence of nodes $P = \mathcal{S}.\gnode(S_{0}, j+i) \dotsm \mathcal{S}.\gnode(S_{0}, j+\abs{X}-1)$ is a path in graph $G$ with $G.\glabel(P) = X$.
+(b) If $\abs{X} = 1$, the statement is true by definition.
+Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
+Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
+By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
+
+(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
+Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
+By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
 \end{proof}
@@
 
-Let $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
+Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
 By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
@@
 (b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
-(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in (\Sigma \setminus \set{\#, \$})^{\ast}$ for all nodes $v' \in G'.\find(X)$.
+(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
 \end{lemma}
@@
 
-\subsection{GCSA for path indexes}
+\subsection{GCSA for path graphs}
+
+Path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}.
+As our technical details are different from the ones used in that paper, we now describe how the GCSA of a path graph works.
+
+% FIXME refer to the figure in Section 2.4
+Let $G = (V, E)$ be a path graph.
+We sort the nodes $V$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$.
+For each node $v \in V$, we append
+\begin{itemize}
+\item $\BWT$ with the predecessor labels $G.\glabel(u)$ for all edges $(u, v) \in E$;
+\item $\bvIN$ with the indegree $x = G.\gindegree(v)$ encoded as a binary sequence $0^{x-1} 1$; and
+\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v)$ encoded as a binary sequence $0^{x-1} 1$.
+\end{itemize}
+If node $v \in V$ has lexicographic rank $i$, the range of incoming edges $(u, v) \in E$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$, with $sp_{in} = 0$ if $i = 0$.
+The labels of predecessor nodes $u$ are encoded in $\BWT[sp_{in}, ep_{in}]$.
+Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u, v)$, is equivalent to sorting them by strings $G.\glabel(u) \cdot G.\gkey(v)$.
+As multiple edges may have the same sort keys, our sorting algorithm must be stable.
+We can achieve the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.
+
+Consider now the outgoing edges.
+If the lexicographic rank of node $u \in V$ is $i'$, the range of outgoing edges $(u, v) \in E$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$, with $sp_{out} = 0$ if $i' = 0$.
+The edges are already sorted by keys $G.\gkey(u)$.
+All outgoing edges from the same node have the same sort key, as there is nothing to distinguish them from one another.
+Because graph $G$ is a path graph, we know that key $G.\gkey(u)$ prefix-matches string $G.\glabel(u) \cdot G.\gkey(v)$.
+The sorting orders are therefore compatible.
+For every $j \in [sp_{in}, ep_{in}]$ for node $v \in V$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u \in V$ implies an edge $(u, v) \in E$.
+If we sort the incoming edges $(u, v) \in E$ stably by strings $G.\glabel(u) \cdot G.\gkey(v)$, we have $(u, v) < (u', v')$ in that order, if $G.\gkey(u) < G.\gkey(u')$.
+
+We implement query $G.\find(X)$ with backward searching.
+Let $X \in \patternset$ be a pattern.
+If $\abs{X} = 0$, query $G.\find(X)$ returns the lexicographic range $[0, \abs{V}-1]$ containing all nodes of the path graph.
+Now assume that $\abs{X} \ge 1$ and that query $G.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$.
+We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$.
+We start by mapping the node range $[sp_{i+1}, ep_{i+1}]$ into range
+$$
+[sp_{in}, ep_{in}] = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]
+$$
+of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$). Then we map that range into the corresponding range of outgoing edges
+$$
+[sp_{out}, ep_{out}] = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1].
+$$
+The lexicographic range containing nodes $G.\find(X[i, \abs{X}-1])$ is
+$$
+[sp_{i}, ep_{i}] = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+$$
+
+% FIXME from here
+Samples
+- sampling conditions
+- sample encoding
+- locate()
+
+Two encodings: general and the one for $\abs{G.\gpred(v, c)} \le 1$. Difference: no general LF in the latter one, must go forward for samples.
 
@@
 \subsection{Extensions}
+
+These are both based on ST, which requires a maximally pruned DBG
+
+count()
+- SadaCount \cite{Sadakane2007a}
+- compressible \cite{Gagie2015}
+
+parent()
+- CST-NPR: \cite{Fischer2009a}
+- CST MEMs: \cite{Ohlebusch2010a}
+- used for MEM mapping: \cite{Li2013}
 

### 2016-04-15 paper/paper.tex
Moved proofs to an appendix

@@
 HISAT / HISAT2: \cite{Kim2015}
+
+Mention that proofs of lemmas are in the appendix
 
@@
 $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform.
-\LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes into BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges into the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
+\LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
 
@@
 Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph.
-For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}:
-$$
-G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
-$$
-Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$.
+For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$.
 Due to the nature of de~Bruijn graphs, this index will not produce any false negatives.
@@
 
-\begin{proof}
-Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
-The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
-
-For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
-By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
-Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
-As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
-Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
-\end{proof}
-
 When used as an index, a path graph will not produce false negatives.
@@
 
-\begin{proof}
-If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
-
-(b) If $\abs{X} = 1$, the statement is true by definition.
-Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
-Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
-By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
-
-(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
-Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
-By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
-\end{proof}
-
-\begin{lemma}[Pruning a de~Bruijn graph]\label{lemma:dbg-prune}
+\begin{lemma}[Pruning]\label{lemma:dbg-prune}
 Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
@@
 \end{lemma}
-
-\begin{proof}
-Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
-Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys.
-For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.
-
-Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
-By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
-We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
-The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
-If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
-
-Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
-If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
-Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
-Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
-As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
-When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
-
-Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
-We can also start from path $P''$ and determine the node $v_{0}$.
-As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
-\end{proof}
 
@@
 
-\begin{lemma}[Consequences of maximal pruning]\label{lemma:dbg-maximal}
+\begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
 Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
@@
 \end{lemma}
-
-\begin{proof}
-(a) For each node $u' \in G'.\gpred(v', c)$, key $G.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
-If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct.
-In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match.
-Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
-Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
-
-(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
-Key $G'.\gkey(u')$ must have $c \cdot G'.\gkey(v')$ as a prefix.
-If there is a substring $S[i,j] = c \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', c)$.
-By (a), $u'$ is the only such node, and hence string $c \cdot G'.\gkey(v')$ can also serve as its key.
-
-(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
-If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
-Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
-By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
-By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
-Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
-\end{proof}
 
@@
 We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$.
-We start by mapping the node range $[sp_{i+1}, ep_{i+1}]$ into range
-$$
-[sp_{in}, ep_{in}] = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]
-$$
-of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$). Then we map that range into the corresponding range of outgoing edges
-$$
-[sp_{out}, ep_{out}] = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1].
-$$
-The lexicographic range containing nodes $G.\find(X[i, \abs{X}-1])$ is
-$$
-[sp_{i}, ep_{i}] = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
-$$
+We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G.\find(X[i, \abs{X}-1])$:
+\begin{align*}
+[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
+[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1];\ \textrm{and} \\
+[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+\end{align*}
 
@@
 
+\clearpage
+\appendix
+\section{Proofs of lemmas}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:pg-fn} (No false negatives)]
+Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
+The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
+
+For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
+By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
+Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
+As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
+If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
+
+(b) If $\abs{X} = 1$, the statement is true by definition.
+Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
+Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
+By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
+
+(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
+Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
+By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-prune} (Pruning)]
+Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys.
+For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.
+
+Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
+By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
+We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
+The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
+If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
+
+Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
+If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
+Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
+Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
+As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
+When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
+
+Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
+We can also start from path $P''$ and determine the node $v_{0}$.
+As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
+(a) The key $G.\gkey(u')$ of a node $u' \in G'.\gpred(v', c)$ prefix-matches string $c \cdot G.\gkey(v')$.
+If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct.
+In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match.
+Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
+Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
+
+(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
+Key $G'.\gkey(u')$ must have $c \cdot G'.\gkey(v')$ as a prefix.
+If there is a substring $S[i,j] = c \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', c)$.
+By (a), $u'$ is the only such node, and hence string $c \cdot G'.\gkey(v')$ can also serve as its key.
+
+(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
+If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
+Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
+By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
+By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
+Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
+\end{proof}
+
 \end{document}

### 2016-04-17 paper/paper.tex
Continued Section 4

@@
 Given string $S[0, n-1]$, we define $S.\rank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
-We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions correspond to the ones used in the SDSL library \cite{Gog2014b}.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries.
+We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}.
+We assume for convenience that $S.\select(0, c) = -1$.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries.
 \emph{Wavelet trees} \cite{Grossi2003} are space-efficient data structures that use bitvectors to support $\rank$/$\select$ queries on arbitrary strings.
@@
 Given the similarities to the suffix array, we can use the BWT as a space-efficient text index.
-The \emph{\FMindex} \cite{Ferragina2005a} combines a plain or compressed representation of the BWT supporting rank/select queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
+The \emph{\FMindex} \cite{Ferragina2005a} combines a plain or compressed representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
 It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}.
@@
 We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges.
-The labels of the children of node $i$ are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$ (with $0$ as the lower bound for $i = 0$).
+The labels of the children of node $i$ are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.
 
@@
 $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform.
-\LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
+\LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
 
@@
 % FIXME refer to the figure in Section 2.4
-Let $G = (V, E)$ be a path graph.
-We sort the nodes $V$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$.
-For each node $v \in V$, we append
+Let $G' = (V', E')$ be a path graph.
+We sort the nodes $V'$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$.
+For each node $v' \in V'$, we append
 \begin{itemize}
-\item $\BWT$ with the predecessor labels $G.\glabel(u)$ for all edges $(u, v) \in E$;
-\item $\bvIN$ with the indegree $x = G.\gindegree(v)$ encoded as a binary sequence $0^{x-1} 1$; and
-\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v)$ encoded as a binary sequence $0^{x-1} 1$.
+\item $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$;
+\item $\bvIN$ with the indegree $x = G.\gindegree(v')$ encoded as a binary sequence $0^{x-1} 1$; and
+\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v')$ encoded as a binary sequence $0^{x-1} 1$.
 \end{itemize}
-If node $v \in V$ has lexicographic rank $i$, the range of incoming edges $(u, v) \in E$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$, with $sp_{in} = 0$ if $i = 0$.
-The labels of predecessor nodes $u$ are encoded in $\BWT[sp_{in}, ep_{in}]$.
-Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u, v)$, is equivalent to sorting them by strings $G.\glabel(u) \cdot G.\gkey(v)$.
+If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$.
+The labels of predecessor nodes $u'$ are encoded in $\BWT[sp_{in}, ep_{in}]$.
+Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v')$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$.
 As multiple edges may have the same sort keys, our sorting algorithm must be stable.
-We can achieve the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.
+We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.
 
 Consider now the outgoing edges.
-If the lexicographic rank of node $u \in V$ is $i'$, the range of outgoing edges $(u, v) \in E$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$, with $sp_{out} = 0$ if $i' = 0$.
-The edges are already sorted by keys $G.\gkey(u)$.
+If the lexicographic rank of node $u' \in V'$ is $i'$, the range of outgoing edges $(u', v') \in E'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$.
+The edges are already sorted by keys $G'.\gkey(u')$.
 All outgoing edges from the same node have the same sort key, as there is nothing to distinguish them from one another.
-Because graph $G$ is a path graph, we know that key $G.\gkey(u)$ prefix-matches string $G.\glabel(u) \cdot G.\gkey(v)$.
+Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
 The sorting orders are therefore compatible.
-For every $j \in [sp_{in}, ep_{in}]$ for node $v \in V$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u \in V$ implies an edge $(u, v) \in E$.
-If we sort the incoming edges $(u, v) \in E$ stably by strings $G.\glabel(u) \cdot G.\gkey(v)$, we have $(u, v) < (u', v')$ in that order, if $G.\gkey(u) < G.\gkey(u')$.
+For every $j \in [sp_{in}, ep_{in}]$ for node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u' \in V'$ implies an edge $(u', v') \in E'$.
+If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.
 
-We implement query $G.\find(X)$ with backward searching.
+We implement query $G'.\find(X)$ with \emph{backward searching}.
 Let $X \in \patternset$ be a pattern.
-If $\abs{X} = 0$, query $G.\find(X)$ returns the lexicographic range $[0, \abs{V}-1]$ containing all nodes of the path graph.
-Now assume that $\abs{X} \ge 1$ and that query $G.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$.
-We want to find the lexicographic range corresponding to the union of sets $G.\gpred(v, X[i])$ over nodes $v \in G.\find(X[i+1, \abs{X}-1])$.
-We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G.\find(X[i, \abs{X}-1])$:
+If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes of the path graph.
+Now assume that $\abs{X} \ge 1$ and that query $G'.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$.
+We want to find the lexicographic range corresponding to the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$.
+We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
 \begin{align*}
@@
 \end{align*}
+We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i}], X[i])$.
 
-% FIXME from here
-Samples
-- sampling conditions
-- sample encoding
-- locate()
+Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$.
+Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above.
+To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM-indexes.
+We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind.
+If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.
 
-Two encodings: general and the one for $\abs{G.\gpred(v, c)} \le 1$. Difference: no general LF in the latter one, must go forward for samples.
+Let $v' \in V'$ be a node of the path graph.
+We sample the values $G'.\gvalue(v')$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $(u', v') \in E'$ is the only incoming edge to node $v'$, and $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$.
+We may also sample the values for some nodes on long unary paths for performance reasons.
+If the values $G'.\gvalue(v')$ has not been sampled, we can derive them from sampled values by following the incoming edges backwards.
+
+Following edges backwards is based on a similar generalization of \LFmapping{} as backward searching. If node $v' \in V'$ has lexicographic rank $i$, the lexicographic rank of its only predecessor is
+$G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$.
+If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.
+
+Let $B_{S}[0, \abs{V'}-1]$ be a bitvector marking the sampled nodes.
+If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S} = 1$.
+We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$.
+For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$.
+We then store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each.
+The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.
+
+If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph \cite{Siren2014}.
+We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
+If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$.
+Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
+$$
+G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
+$$
+We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector.
+On the other hand, computing $G'.\LF()$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
+The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.
 
 \subsection{Index construction}
+
+Prefix-doubling; k to 2k, 4k, 8k. original GCSA did not have to store labels, as it built an unlimited order graph
+
+prune, extend, full prune, merge
+
+paths: start, extension, predecessors, key; encode key as a sequence of kmer ranks
+
+prune: only if start node is the same; full prune as before
+
+extend: join when extension matches start; do not extend if already pruned; if right side is pruned, the extended path is also pruned
+
+merge: invariant: path labels are the key set of a pruned de Bruijn graph; full pruning produces a maximally pruned graph
+
+disk-based: each chromosome in a separate file sorted by keys; extend by file; prune/merge essentially merge sorted lists
+
+edges and construction: use \LFmapping{} in GCSA of the de~Bruijn graph of the kmers to determine the rank sequence of $c \cdot G'.\gkey(v')$; need $\sigma + 1$ sequential pointers in each file
 

### 2016-04-17 paper/paper.tex
Clafifications

@@
 
-The following definitions are useful when discussing text indexes and path indexes.
+The following definitions are useful when discussing path indexes.
 
@@
 
-De~Bruijn graphs are a special case of pruned de~Bruijn graphs.
-Hence we may compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
+We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
+Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of graph $G$, and let $G' = (V', E')$ be a pruned de~Bruijn graph produced by that process.
+Each node $v' \in V'$ is an \emph{equivalence class} of nodes $V_{d}(v') \subseteq V_{d}$ corresponding to a shared prefix $G'.\gkey(v')$ of keys.
+For all $v \in V_{d}(v')$, we have $G_{d}.\gvalue(v) = G'.\gvalue(v')$.
 
@@
 
-These are both based on ST, which requires a maximally pruned DBG
+Path graphs; LCP array; compact trie of keys; LCP interval tree
 
@@
 Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
-Each node $v'' \in V''$ is an equivalence class of nodes $V_{d}(v'') \subseteq V_{d}$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys.
-For all $v \in V_{d}(v'')$, we have $G_{d}.\gvalue(v) = G''.\gvalue(v'')$.
 
@@
 Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
+
+Note that any key set satisfying (a) and (b) also satisfies (c).
+If $G'' = (V'', E'')$ is a pruned de~Bruijn graph and $v'' \in V''$ is a node, any pattern matching node $v''$ prefix-matches the key of the corresponding equivalence class in graph $G'$.
 \end{proof}

### 2016-04-18 paper/paper.tex
Finished the sections on the index and the experiments. No proofreading though.

@@
 \newcommand{\locate}{\ensuremath{\mathsf{locate}}}
+\newcommand{\parent}{\ensuremath{\mathsf{parent}}}
+\newcommand{\countq}{\ensuremath{\mathsf{count}}}
 
@@
 \newcommand{\gnode}{\ensuremath{\mathsf{node}}}
+\newcommand{\gext}{\ensuremath{\mathsf{ext}}}
 
@@
 
-In this paper, we primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$.
+We primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$.
 Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base.
@@
 
-The following definitions are useful when discussing path indexes.
-
-\begin{definition}[Prefix-matching strings]
-Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
-\end{definition}
-
-\begin{definition}[Prefix-free set]
-Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
-\end{definition}
+Let $S$ be a string and $S'$ be be a string or an infinite character sequence over alphabet $\Sigma$.
+We say that sequences $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
+Set $\mathcal{S}$ of strings is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
 
@@
 
-The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting arbitrary substring queries.
+The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries.
 It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges.
+If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$.
 Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.
@@
 If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.
-In a typical case, we expect an \FMindex{} to take less than $n \log \sigma$ bits of space and to be able to find about 100,000 pattern occurrences per second \cite{Ferragina2009a}.
 
@@
 
-Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$.
+Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$ and removing duplicates.
 Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above.
-To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM-indexes.
+To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM\nobreakdash-indexes.
 We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind.
@@
 
-If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph \cite{Siren2014}.
+If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph.
 We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
@@
 
-Prefix-doubling; k to 2k, 4k, 8k. original GCSA did not have to store labels, as it built an unlimited order graph
+GCSA construction \cite{Siren2014} is based on the original \emph{prefix-doubling} algorithm for suffix array construction \cite{Manber1993}.
+The original GCSA started from paths of length $1$ in the input graph, and then repeatedly \emph{joined} paths of length $k$ into paths of length $2k$, until each path had a distinct label.
+The resulting path graph was essentially an unlimited-order pruned de~Bruijn graph and supported queries of arbitrary length.
 
-prune, extend, full prune, merge
+We use a variant of that algorithm with GCSA2.
+Let $G = (V, E)$ be the input graph.
+We extract all paths of length $k$ (typically for $k = 16$) from graph $G$.
+For each path $P = v_{0} \dotsm v_{\abs{P}-1}$, we store several fields.
+\emph{Key} $P.\gkey$ encodes $G.\glabel(P)$ as a sequence of lexicographic ranks of \kmer{k}s.
+If $\abs{P}$ is not an integer multiple of $k$, the key consists of the \kmer{k} ranks for the lexicographically smallest \kmer{(\lceil \abs{P}/k \rceil \cdot k)} having $G.\glabel(P)$ as a prefix, followed by the rank of the last \kmer{k} in the largest such \kmer{(\lceil \abs{P}/k \rceil \cdot k)}.
+\emph{Value} $P.\gvalue$ is the start node $v_{0}$ of the path.
+We store the set of \emph{predecessor labels} $\set{c \in \Sigma \mid \abs{G.\gpred(v_{0}, c)} > 0}$ as $P.\gpred$.
+For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a new copy of the path and store the node as $P.\gext = v$.
 
-paths: start, extension, predecessors, key; encode key as a sequence of kmer ranks
+The construction uses several supporting structures.
+We build an \orderk{k} \emph{de~Bruijn graph} $G_{d} = (V_{d}, E_{d})$ of the path labels and encode it as a GCSA, using the predecessor labels $P.\gpred$ for determining the edges.
+Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys.
+We use two additional arrays: \emph{the LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$.
+The LCP array is stored in a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.
 
-prune: only if start node is the same; full prune as before
+Because we have to store path labels explicitly in the keys, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$.
+After $d$ doubling steps, the length of the paths is $2^{d} k$, and we can use them to build a maximally pruned \orderk{(2^{d} k)} de~Bruijn graph.
+Each doubling step consists of a \emph{pruning} step, followed by an \emph{extension} step.
+The pruning step uses a limited form of Lemma~\ref{lemma:dbg-prune} and applies it to all possible ranges of paths.
+Given two paths $P$ and $P'$, we can determine the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array.
+If all paths with a certain prefix of their labels start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.
 
-extend: join when extension matches start; do not extend if already pruned; if right side is pruned, the extended path is also pruned
+The extension step transforms the current set of paths of length (up to) $k'$ and transforms it into a set of paths of length (up to) $2k'$.
+If $P$ is a path with $P.\gext = -1$, we use it as such.
+If $P.\gext = P'.\gvalue$ for paths $P$ and $P'$, we create a new path $PP'$.
+We set $(PP').\gkey$ according to the concatenation of the path labels, take $\gvalue$ and $\gpred$ from path $P$, and take $\gext$ from path $P'$.
+If we have another path $Q$ with $Q.\gkey = P.\gkey$ such that $QP'$ is a path, and if $P'.\gext = -1$, all possible \kmer{2k'} extensions of label $G.\glabel(PP')$ are also labels of paths starting from node $Q.\gvalue$, and the other way around.
+Hence paths $PP'$ and $QP'$ can be represented by a single node in a pruned de~Bruijn graph.
 
-merge: invariant: path labels are the key set of a pruned de Bruijn graph; full pruning produces a maximally pruned graph
+The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.
+First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
+If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$.
+We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$.
+Then we apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.
 
-disk-based: each chromosome in a separate file sorted by keys; extend by file; prune/merge essentially merge sorted lists
+Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs.
+To avoid that, we store them on \emph{disk} when possible.
+Each chromosome forms a subgraph that is disjoint from the other chromosomes, except near the source node and the sink node.
+Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order.
+A doubling step can be done separately for each of the chromosomes.
+A pruning step takes a set of sorted lists paths and merges the lists.
+It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged.
+The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer.
+The merging step is mostly the same as a pruning step, but the output format is different.
 
-edges and construction: use \LFmapping{} in GCSA of the de~Bruijn graph of the kmers to determine the rank sequence of $c \cdot G'.\gkey(v')$; need $\sigma + 1$ sequential pointers in each file
+After storing the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$ on disk, we proceed to \emph{building the index}.
+Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing.
+Because graph $G''$ is a path graph, there is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$.
+We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers.
+The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a $c \in \Sigma$.
+We can also sample the nodes for $\locate$ queries during the same scans.
+
+Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V, E)$ and the last character array $L$.
+If the lexicographic rank of \kmer{k} $X$ is $i$, the lexicographic rank of \kmer{k} $(cX)[0, k-1]$ is $G_{d}.\LF(i, c)$.
+If \kmer{2k} $X$ is encoded with \kmer{k} ranks $(i, j)$, we can encode string $cX$ as a lexicographic range of \kmer{k} rank sequences, with $(G_{d}.\LF(i, c), G_{d}.\LF(j, L[i]), G_{d}.\LF(0, L[j]))$ as the lower bound and $G_{d}.\LF(\abs{V_{d}}-1, L[j])$ as the last rank in the upper bound.
+String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.
 
@@
 
-Path graphs; LCP array; compact trie of keys; LCP interval tree
+GCSA is a generalization of FM\nobreakdash-indexes for path graphs.
+Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.
 
-count()
-- SadaCount \cite{Sadakane2007a}
-- compressible \cite{Gagie2015}
+Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
+We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
+The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
+The compacted trie can be considered the \emph{suffix tree of the path graph}.
 
-parent()
-- CST-NPR: \cite{Fischer2009a}
-- CST MEMs: \cite{Ohlebusch2010a}
-- used for MEM mapping: \cite{Li2013}
+The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
+Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
+These ranges can be determined from the LCP array.
+If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
+
+% FIXME justify why we can use parent queries
+% we need a lemma stating that we always match entire equivalence classes
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
+However, if pattern $X$ of length $\abs{X} \le k$ matches lexicographic range $[sp, ep]$, we can safely use a $\parent$ query for the corresponding node in the suffix tree.
+
+BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference.
+It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions.
+GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$).
+Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
+We have implemented a GCSA-based read aligner using that algoritm in vg \cite{Garrison2014-2016}.
+
+In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
+In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes.
+Determining the  number of distinct start nodes with a $\locate$ query can be slow.
+Hence we need additional structures to support efficient \emph{counting queries}.
+The problem is similar to determining the \emph{document frequency} of a pattern.
+Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
+We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+
+Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
+If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
+Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times.
+We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$.
+If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
+For subsequent visits to the same node, we set $R[j] = 0$.
+Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$.
+We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$.
+If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
+$$
+\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
+$$
+
+For value counting, we use array $R$ to store the number of redundant values in each suffix tree node.
+We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$.
+We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above.
+The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
+$$
+G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
+$$
+where $[sp, ep]$ is the lexicographic range returned by the $\find$ query.
+The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
 
@@
 
+GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}.
+The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
+As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.
+
+The \emph{variation} graphs used in vg have strings as node labels.
+The average length of a node label is typically around 10~bases.
+Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
+When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
+Input graph nodes corresponding to first offsets of variation graph nodes are always selected as samples for performance reasons.
+
+The following experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
+Input files, final indexes, and temporary files were all stored on a distributed Lustre file system.
+We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the query benchmarks.
+
+We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
+To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases.
+We then extracted paths of length $16$ from the forward strands of all chromosomes.
+There were a total of 4.80~billion paths with 1.53~billion distinct labels.
+We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively.
+The results are summarized in Table~\ref{table:construction}.
+
+\begin{table}[t!]
+\begin{center}
+\caption{GCSA2 construction with different orders of the path graph.
+Billions of nodes in the de~Bruijn graph and the pruned de~Bruijn graph; index size in gigabytes; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
+\begin{tabular}{c|ccc|ccc|cc}
+\hline
+\textbf{Order} & \textbf{Nodes} & \textbf{Pruned} & \textbf{Index} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
+\hline
+ 32 & 6.23G & 4.39G & 13.7 GB & 9.85 h & 63.2 GB & 401 GB & 1.55 TB & 1.05 TB \\
+ 64 & 16.9G & 5.27G & 14.2 GB & 13.6 h & 56.1 GB & 424 GB & 2.23 TB & 1.71 TB \\
+128 &  118G & 5.76G & 15.3 GB & 18.0 h & 56.9 GB & 489 GB & 3.00 TB & 2.47 TB \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+A single-strand whole-genome index can be built overnight with 96~gigabytes of memory.
+Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower.
+The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.
+The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.
+
+We extracted \kmer{16}s and \kmer{32}s from the vg graphs for benchmarking purposes using \texttt{vg sim}.
+We then queried for them in the \orderk{64} index using a single thread.
+The results can be seen in Table~\ref{table:benchmark}.
+Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction.
+The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index.
+The main reason is that we use to use $\select$ queries, which are much slower than $\rank$ queries on plain bitvectors \cite{Gog2014}.
+
+\begin{table}[t!]
+\begin{center}
+\caption{Performance benchmarks using an \orderk{64} GCSA2.
+Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in milliseconds.}\label{table:benchmark}
+\begin{tabular}{c|cc|cc|cccc}
+\hline
+\textbf{Length} & \textbf{Patterns} & \textbf{Matches} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+\hline
+16 & 351584 & 347453 & 1830M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
+32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 0.47 ms \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \section{Discussion}
 
-% FIXME
 \subparagraph*{Acknowledgements.}
 
-I thank Richard Durbin, Erik Garrison, and Adam Novak for \dots
+I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index.
+Many extensions over the original GCSA and clarifications of the assumptions I made are a direct result of those discussions.
 

### 2016-04-19 paper/paper.tex
Almost finished the paper. Probably full of typos and confusion.

@@
 
-1000GP: \cite{1000GP2015}
+Sequencing pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species.
+As reference genomes are usually assembled from the genomes of a small number of individuals, they are biased towards those individuals.
+This \emph{reference bias} may affect the results of subsequent analysis, especially when the sequenced individuals are from different populations than the ones behind the reference genome.
 
-Graph genomes: \cite{Schneeberger2009}
+Recently there has been much interest in \emph{variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
+These graphs are eventually expected to replace reference sequences as reference genomes.
+The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.
 
-FM-index: \cite{Ferragina2005a}
+The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools.
+This paper describes \emph{GCSA2}, the path index developed for vg.
+A \emph{path index} is a generalization of text indexes for labeled graphs.
+They are used for finding paths with labels that match the query string.
+Indexing graphs for path queries is an inherently hard problem, as the number of paths increases exponentially with path length.
+Hence the design of a path index is a trade-off between maximum supported query length, the size of the index, and pruning complex regions of the graph before indexing.
 
-XBW: \cite{Ferragina2009b}
+Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}.
+Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index.
+The original GCSA could only index directed acyclic graphs and depended on pruning the complex regions, as it did not place any limits on query length.
+Kim et~al.~\cite{Kim2015-2016} combined GCSA with the ideas developed for the HISAT read aligner \cite{Kim2015}.
+The resulting HISAT2 aligner was the first practical graph-based read aligner, though it only uses the graph for more accurate mapping to a linear reference.
 
-GCSA: \cite{Siren2014}
+The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} used a similar generalization of the FM\nobreakdash-index as in GCSA for representing \emph{de~Bruijn graphs} space-efficiently.
+Rødland \cite{Roedland2013} proposed another similar generalization.
+The succinct de~Bruijn graphs have been generalized to represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
+Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space savings by representing unary paths in the graph as a single node.
+A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.
 
-Succinct DBG: \cite{Bowe2012}, variable-order \cite{Boucher2014}, alternative representation \cite{Roedland2013}
+Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences.
+Simple variants are encoded by using the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site.
+This is usually enough to keep the index small, at the price of much higher query times.
+The hypertext index \cite{Thachuk2013} is another similar idea.
+The nodes of the graph are labeled with strings, which are indexed using an ordinary FM-index.
+Partial matches in the strings are combined into full matches with range queries in the edge matrix.
+While matches crossing one edge are easy to find, the approach becomes complex and potentially slow with matches crossing multiple edges.
+There are also several indexes (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient representation for similar sequences but index the sequences instead of the graph.
 
-BWBBLE: \cite{Huang2013}
+GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs.
+Conceptually it uses a de~Bruijn graph as a path index of a variation graph.
+The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely.
+The pruned graph is then encoded with a generalization of the FM\nobreakdash-index.
+GCSA2 includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
+These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
 
-Contracted DBG: \cite{Cazaux2014}
-
-Compressed DBG: \cite{Marcus2014}
-
-Hypertext index: \cite{Thachuk2013}
-
-FM-index of alignment: \cite{Na2015}
-
-HISAT / HISAT2: \cite{Kim2015}
-
-Mention that proofs of lemmas are in the appendix
+This paper contains several lemmas describing the properties of generalizations of de~Bruijn graphs. Their proofs are included in Appendix~\ref{appendix:proofs}.
 
@@
 
-\section{Path indexes}\label{sect:path-indexes}
+\section{Path Indexes}\label{sect:path-indexes}
 
@@
 
+\begin{lemma}[Context length]\label{lemma:pg-context}
+Let $G' = (V', E')$ be a path graph, and let $X \in \patternset$ be a pattern.
+The set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$.
+\end{lemma}
+
 When used as an index, a path graph will not produce false negatives.
@@
 De~Bruijn graphs often have nodes that are redundant when the graph is used as a path index.
-Removing these nodes from the graph would make the index smaller and more practical.
-On the other hand, we still want to guarantee that the index works correctly with patterns of length $k$ or less, where $k$ is the order of the de~Bruijn graph.
+By carefully pruning the graphs, we get smaller graphs that are equivalent to \orderk{k} de~Bruijn graphs with patterns of length $k$ or less.
+These pruned de~Bruijn graphs naturally arise from GCSA construction.
+They are similar to manifold de Bruijn graphs \cite{Lin2014}.
 
@@
 
-Pruned de~Bruijn graphs are similar to manifold de Bruijn graphs \cite{Lin2014}.
-
 \begin{lemma}[No short false positives]\label{lemma:dbg-fp}
@@
 Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
-If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
+If $\abs{V'_{K}} > 1$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \cup \set{K}) \setminus V'_{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
 \end{lemma}
@@
 We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
-Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of graph $G$, and let $G' = (V', E')$ be a pruned de~Bruijn graph produced by that process.
-Each node $v' \in V'$ is an \emph{equivalence class} of nodes $V_{d}(v') \subseteq V_{d}$ corresponding to a shared prefix $G'.\gkey(v')$ of keys.
-For all $v \in V_{d}(v')$, we have $G_{d}.\gvalue(v) = G'.\gvalue(v')$.
+Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $G'' = (V'', E'')$ be the same graph after further pruning.
+Each node $v'' \in V''$ is an \emph{equivalence class} of nodes $V'(v'') \subseteq V'$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys.
+For all $v' \in V'(v'')$, we have $G'.\gvalue(v') = G''.\gvalue(v'')$.
 
@@
 \begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
-Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
+Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of graph $G = (V, E)$. Then
 (a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
-(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
+(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$; and
 (c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
@@
 
+The path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}.
+As many of the technical details are different, we now take a look at how a GCSA of a path graph works.
+We also extend GCSA with techniques based on suffix trees.
+Appendix~\ref{appendix:construction} describes the construction algorithm used with GCSA2.
+While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.
+
 \subsection{GCSA for path graphs}
-
-Path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}.
-As our technical details are different from the ones used in that paper, we now describe how the GCSA of a path graph works.
 
@@
 
-\subsection{Index construction}
+\subsection{Extensions}
+
+GCSA is a generalization of FM\nobreakdash-indexes for path graphs.
+Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.
+
+Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
+We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
+The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
+The compacted trie can be considered the \emph{suffix tree of the path graph}.
+
+The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
+Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
+These ranges can be determined from the LCP array.
+If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
+
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
+By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
+Hence we can safely use $\parent$ queries with such ranges.
+
+BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference.
+It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions.
+GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$).
+Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
+We have implemented a GCSA-based read aligner using that algoritm in vg \cite{Garrison2014-2016}.
+
+In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
+In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes.
+Determining the  number of distinct start nodes with a $\locate$ query can be slow.
+Hence we need additional structures to support efficient \emph{counting queries}.
+The problem is similar to determining the \emph{document frequency} of a pattern.
+Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
+We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+
+Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
+If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
+Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times.
+We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$.
+If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
+For subsequent visits to the same node, we set $R[j] = 0$.
+Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$.
+We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$.
+If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
+$$
+\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
+$$
+
+For value counting, we use array $R$ to store the number of redundant values in each suffix tree node.
+We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$.
+We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above.
+The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
+$$
+G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
+$$
+where $[sp, ep]$ is the lexicographic range returned by the $\find$ query.
+The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
+
+\section{Implementation and Experiments}
+
+GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}.
+The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
+As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.
+
+The \emph{variation graphs} used in vg have strings as node labels.
+The average length of a node label is typically around 10~bases.
+Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
+When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
+Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.
+
+The following experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
+Input files, final indexes, and temporary files were all stored on a distributed Lustre file system.
+We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the query benchmarks.
+
+We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
+To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases.
+We then extracted paths of length $16$ from the forward strands of all chromosomes.
+There were a total of 4.80~billion paths with 1.53~billion distinct labels.
+We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively.
+The results are summarized in Table~\ref{table:construction}.
+
+\begin{table}[t!]
+\begin{center}
+\caption{GCSA2 construction with different orders of the path graph.
+Billions of nodes in the de~Bruijn graph and the pruned de~Bruijn graph; index size in gigabytes; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
+\begin{tabular}{c|ccc|ccc|cc}
+\hline
+\textbf{Order} & \textbf{Nodes} & \textbf{Pruned} & \textbf{Index} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
+\hline
+ 32 & 6.23G & 4.39G & 13.7 GB & 9.85 h & 63.2 GB & 401 GB & 1.55 TB & 1.05 TB \\
+ 64 & 16.9G & 5.27G & 14.2 GB & 13.6 h & 56.1 GB & 424 GB & 2.23 TB & 1.71 TB \\
+128 &  118G & 5.76G & 15.3 GB & 18.0 h & 56.9 GB & 489 GB & 3.00 TB & 2.47 TB \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+A single-strand whole-genome index can be built overnight with 96~gigabytes of memory.
+Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower.
+The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
+It consists of a pruning step, followed by merging the paths with identical labels.} The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.
+
+We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread.
+The results can be seen in Table~\ref{table:benchmark}.
+Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction.
+The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index.
+There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out.
+Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.
+
+\begin{table}[t!]
+\begin{center}
+\caption{Performance benchmarks using an \orderk{64} GCSA2.
+Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
+\begin{tabular}{c|cc|cc|cccc}
+\hline
+\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+\hline
+16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
+32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
+64 & 351567 & 326101 & 0.33M & 1.35M &  142 \textmu{}s & 0.34 \textmu{}s & 0.57 \textmu{}s & 12.4 \textmu{}s \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\section{Conclusions}
+
+GCSA2 is a path index for variation graphs.
+It uses a structurally compressed de~Bruijn graph of the paths in the input graph as an index, and encodes that with a generalization of the FM\nobreakdash-index.
+The index also incorporates extensions based on a generalization of suffix trees.
+GCSA2 has been integrated in the variation graph toolkit vg, where it is used e.g.~for read alignment based on maximal exact matches.
+
+We can build a forward strand whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
+The index itself requires less than 16~gigabytes of memory.
+Query performance is roughly 5\nobreakdash--10 times slower than for a fast FM\nobreakdash-index for sequences.
+GCSA2 can search for patterns at roughly 0.8\nobreakdash--1.0~MB/s and locate 100,000\nobreakdash--300,000 occurrences per second using a single CPU core.
+As with other FM\nobreakdash-indexes, the speedup from distributing the queries on multiple cores is quite high.
+
+The compressed suffix tree used in GCSA2 may be useful for indexing texts as well, as it is both simple to implement and fast in practice.
+As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently.
+We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.
+
+% FIXME from here
+% additional edges using hypertext index; haplotypes and potential recombinations
+
+% graph formalism, grammars
+
+\subparagraph*{Acknowledgements.}
+
+I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index.
+Many extensions over the original GCSA and clarifications of the assumptions I made are a direct result of those discussions.
+
+\bibliography{paper}
+
+\clearpage
+\appendix
+\section{Proofs of Lemmas}\label{appendix:proofs}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:pg-fn} (No false negatives)]
+Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
+The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
+
+For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
+By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
+Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
+As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:pg-context} (Context length)]
+If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
+Now assume that $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ is the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of $G'.\gkey(v')$.
+
+Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i])$, and assume that it is nonempty.
+Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
+Hence $u' \in M_{i}$ if and only if key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$.
+If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
+Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
+If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
+
+(b) If $\abs{X} = 1$, the statement is true by definition.
+Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
+Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
+By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
+
+(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
+Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
+By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-prune} (Pruning)]
+Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+
+Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
+By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
+We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
+The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
+If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
+
+Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
+If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
+Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
+Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
+As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
+When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
+
+Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
+We can also start from path $P''$ and determine the node $v_{0}$.
+As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
+(a) For each node $u' \in G'.\gpred(v', c)$, the key $G'.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
+If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G'.\gkey(v')$ or they will prefix-match one another.
+In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
+Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
+Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
+
+(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
+Then key $G'.\gkey(u')$ must have string $G'.\glabel(u') \cdot G'.\gkey(v')$ as a prefix.
+If there is a substring $S[i,j] = G'.\glabel(u') \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', G'.\glabel(u'))$.
+By (a), $u'$ is the only such node, and there cannot be an edge $(u', w') \in E'$ to a node $w' \ne v'$.
+Hence string $G'.\glabel(u') \cdot G'.\gkey(v')$ can also serve as the key of node $u' \in V'$.
+
+(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
+If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
+Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
+By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
+By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
+Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
+\end{proof}
+
+\section{Index Construction}\label{appendix:construction}
 
@@
 
-\subsection{Extensions}
-
-GCSA is a generalization of FM\nobreakdash-indexes for path graphs.
-Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.
-
-Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
-We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
-The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
-The compacted trie can be considered the \emph{suffix tree of the path graph}.
-
-The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
-Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
-These ranges can be determined from the LCP array.
-If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
-
-% FIXME justify why we can use parent queries
-% we need a lemma stating that we always match entire equivalence classes
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
-Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
-However, if pattern $X$ of length $\abs{X} \le k$ matches lexicographic range $[sp, ep]$, we can safely use a $\parent$ query for the corresponding node in the suffix tree.
-
-BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference.
-It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions.
-GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$).
-Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
-We have implemented a GCSA-based read aligner using that algoritm in vg \cite{Garrison2014-2016}.
-
-In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
-In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes.
-Determining the  number of distinct start nodes with a $\locate$ query can be slow.
-Hence we need additional structures to support efficient \emph{counting queries}.
-The problem is similar to determining the \emph{document frequency} of a pattern.
-Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
-We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
-
-Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
-If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
-Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times.
-We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$.
-If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
-For subsequent visits to the same node, we set $R[j] = 0$.
-Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$.
-We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$.
-If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
-$$
-\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
-$$
-
-For value counting, we use array $R$ to store the number of redundant values in each suffix tree node.
-We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$.
-We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above.
-The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
-$$
-G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
-$$
-where $[sp, ep]$ is the lexicographic range returned by the $\find$ query.
-The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
-
-\section{Experiments}
-
-GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}.
-The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
-As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.
-
-The \emph{variation} graphs used in vg have strings as node labels.
-The average length of a node label is typically around 10~bases.
-Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
-When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
-Input graph nodes corresponding to first offsets of variation graph nodes are always selected as samples for performance reasons.
-
-The following experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
-Input files, final indexes, and temporary files were all stored on a distributed Lustre file system.
-We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the query benchmarks.
-
-We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
-To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases.
-We then extracted paths of length $16$ from the forward strands of all chromosomes.
-There were a total of 4.80~billion paths with 1.53~billion distinct labels.
-We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively.
-The results are summarized in Table~\ref{table:construction}.
-
-\begin{table}[t!]
-\begin{center}
-\caption{GCSA2 construction with different orders of the path graph.
-Billions of nodes in the de~Bruijn graph and the pruned de~Bruijn graph; index size in gigabytes; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
-\begin{tabular}{c|ccc|ccc|cc}
-\hline
-\textbf{Order} & \textbf{Nodes} & \textbf{Pruned} & \textbf{Index} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
-\hline
- 32 & 6.23G & 4.39G & 13.7 GB & 9.85 h & 63.2 GB & 401 GB & 1.55 TB & 1.05 TB \\
- 64 & 16.9G & 5.27G & 14.2 GB & 13.6 h & 56.1 GB & 424 GB & 2.23 TB & 1.71 TB \\
-128 &  118G & 5.76G & 15.3 GB & 18.0 h & 56.9 GB & 489 GB & 3.00 TB & 2.47 TB \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-A single-strand whole-genome index can be built overnight with 96~gigabytes of memory.
-Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower.
-The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.
-The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.
-
-We extracted \kmer{16}s and \kmer{32}s from the vg graphs for benchmarking purposes using \texttt{vg sim}.
-We then queried for them in the \orderk{64} index using a single thread.
-The results can be seen in Table~\ref{table:benchmark}.
-Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction.
-The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index.
-The main reason is that we use to use $\select$ queries, which are much slower than $\rank$ queries on plain bitvectors \cite{Gog2014}.
-
-\begin{table}[t!]
-\begin{center}
-\caption{Performance benchmarks using an \orderk{64} GCSA2.
-Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in milliseconds.}\label{table:benchmark}
-\begin{tabular}{c|cc|cc|cccc}
-\hline
-\textbf{Length} & \textbf{Patterns} & \textbf{Matches} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
-\hline
-16 & 351584 & 347453 & 1830M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
-32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 0.47 ms \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{Discussion}
-
-\subparagraph*{Acknowledgements.}
-
-I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index.
-Many extensions over the original GCSA and clarifications of the assumptions I made are a direct result of those discussions.
-
-\bibliography{paper}
-
-\clearpage
-\appendix
-\section{Proofs of lemmas}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:pg-fn} (No false negatives)]
-Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
-The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
-
-For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
-By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
-Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
-As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
-Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
-If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
-
-(b) If $\abs{X} = 1$, the statement is true by definition.
-Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
-Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
-By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
-
-(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
-Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
-By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-prune} (Pruning)]
-Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
-
-Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
-By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
-We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
-The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
-If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
-
-Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
-If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
-Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
-Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
-As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
-When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
-
-Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
-We can also start from path $P''$ and determine the node $v_{0}$.
-As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
-(a) The key $G.\gkey(u')$ of a node $u' \in G'.\gpred(v', c)$ prefix-matches string $c \cdot G.\gkey(v')$.
-If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct.
-In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match.
-Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
-Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
-
-(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
-Key $G'.\gkey(u')$ must have $c \cdot G'.\gkey(v')$ as a prefix.
-If there is a substring $S[i,j] = c \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', c)$.
-By (a), $u'$ is the only such node, and hence string $c \cdot G'.\gkey(v')$ can also serve as its key.
-
-(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
-If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
-Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
-By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
-By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
-Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
-
-Note that any key set satisfying (a) and (b) also satisfies (c).
-If $G'' = (V'', E'')$ is a pruned de~Bruijn graph and $v'' \in V''$ is a node, any pattern matching node $v''$ prefix-matches the key of the corresponding equivalence class in graph $G'$.
-\end{proof}
-
 \end{document}

### 2016-04-19 paper/paper.tex
Simplified lemmas.

@@
 We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s.
+Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$.
 We say that string $S'$ is a substring of string collection $\mathcal{S}$, if there is a string $S \in \mathcal{S}$ such that $S'$ is a substring of $S$.
@@
 Let $G' = (V', E')$ be a path graph, and let $X \in \patternset$ be a pattern.
-The set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$.
+The set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$, which depends on the graph and the pattern.
+\end{lemma}
+
+\begin{lemma}[Short keys]\label{lemma:pg-keys}
+Let $G' = (V', E')$ be a path graph, where $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}$ for all edges $(u', v') \in E'$. Then
+(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and
+(b) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
 \end{lemma}
@@
 \begin{lemma}[No short false positives]\label{lemma:dbg-fp}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$. Then
-(a) $G'.\locate(X)$ is a set of start nodes $v_{0} \in V$ of paths $P$ matching the pattern in graph $G$; and
-(b) if $G'$ is a de~Bruijn graph, $X$ is prefix of $G'.\gkey(v'_{0})$ for all $v'_{0} \in G'.\find(X)$.
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$.
+Then $G'.\locate(X)$ is a set of start nodes $v\in V$ of paths $P$ matching the pattern in graph $G$.
 \end{lemma}
@@
 \begin{lemma}[Pruning]\label{lemma:dbg-prune}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
-If $\abs{V'_{K}} > 1$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \cup \set{K}) \setminus V'_{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v)$.
+If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
 \end{lemma}
@@
 \begin{definition}[Maximally pruned de~Bruijn graph]
-Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$.
-We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune}, and the keys in $\mathcal{K}$ cannot be made shorter without changing the graph structure.
+Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune} any further.
 \end{definition}
@@
 \begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
-Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of graph $G = (V, E)$. Then
-(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
-(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$; and
-(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
+Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of graph $G = (V, E)$. Then $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$.
 \end{lemma}
@@
 
-The path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}.
+Path graphs are essentially equivalent to the prefix-sorted automata used in the original paper on GCSA \cite{Siren2014}.
 As many of the technical details are different, we now take a look at how a GCSA of a path graph works.
 We also extend GCSA with techniques based on suffix trees.
-Appendix~\ref{appendix:construction} describes the construction algorithm used with GCSA2.
+Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$.
 While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.
@@
 We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector.
-On the other hand, computing $G'.\LF()$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
+On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
 The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.
@@
 
+\begin{proof}[Proof of Lemma~\ref{lemma:pg-keys} (Short keys)]
+(a) If node $v'$ has multiple predecessors with label $c$, the keys of the predecessors must be longer than string $c \cdot G'.\gkey(v')$, as the key set is prefix-free.
+
+(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
+\end{proof}
+
 \begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
-If (a) holds for de~Bruijn graphs, it also holds for pruned de~Bruijn graphs. From now on, let $G'$ be a de~Bruijn graph, and let $\mathcal(S)$ be the \kcollection{k} used for building graph $G'$.
+We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
 
-(b) If $\abs{X} = 1$, the statement is true by definition.
-Now let $\abs{X} \ge 2$, and let $P' = v'_{0} \dotsm v'_{\abs{X}}$ be a path matching $X$ in graph $G'$.
-Every edge $(v'_{i}, v'_{i+1}) \in E'$ corresponds to a substring $S_{i}[j, j+k] = X[i] \cdot G'.\gkey(v'_{i+1})$ in $\mathcal{S}$, and the substring has $G'.\gkey(v'_{i})$ as a prefix.
-By induction from $\abs{X}-2$ to $0$, substring $S_{i}[j, j+k]$ has $X[i, \abs{X}-1]$ as a prefix.
-
-(a) Now let $S[i, i+k-1] = G'.\gkey(v'_{0})$ be a substring of $\mathcal{S}$, where $v'_{0} \in G'.\find(X)$.
-Every node $v \in G'.\gvalue(v'_{0})$ is the node $\mathcal{S}.\gnode(S, i)$ for one of such substrings.
-By (b), $S[i, i+\abs{X}-1] = X$, which corresponds to a path $P$ matching $X$ in graph $G$, with $\mathcal{S}.\gnode(S, i)$ as the start node.
+Let $v' \in G'.\find(X)$ be a node.
+By Lemma~\ref{lemma:pg-keys}, pattern $X$ prefix-matches key $G'.\gkey(v')$.
+For every node $v \in G'.\gvalue(v')$, there is a substring $S[i, i+k-1] = G'.\gkey(v'_{0})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$, with $\mathcal{S}.\gnode(S, i) = v$.
+Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
 \end{proof}
@@
 \begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
-(a) For each node $u' \in G'.\gpred(v', c)$, the key $G'.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$.
-If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G'.\gkey(v')$ or they will prefix-match one another.
-In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
-Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node.
-Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.
+Let $G_{d} = (V_{d}, E_{d})$ the de~Bruijn graph of graph $G$ with the same order $k$ as graph $G'$, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+If $v' \in V'$ is a node, then $G_{d}.\gvalue(v) = G'.\gvalue(v')$ for all nodes $v \in V_{d}(v')$.
 
-(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$.
-Then key $G'.\gkey(u')$ must have string $G'.\glabel(u') \cdot G'.\gkey(v')$ as a prefix.
-If there is a substring $S[i,j] = G'.\glabel(u') \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', G'.\glabel(u'))$.
-By (a), $u'$ is the only such node, and there cannot be an edge $(u', w') \in E'$ to a node $w' \ne v'$.
-Hence string $G'.\glabel(u') \cdot G'.\gkey(v')$ can also serve as the key of node $u' \in V'$.
+Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$.
+Then string $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$.
+There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
 
-(c) If $\abs{X} = 0$, every key prefix-matches the pattern.
-If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$.
-Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$.
-By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$.
-By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$.
-Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
+Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}.
+Because graph $G'$ is a pruned de~Bruijn graph, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
+As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.
+
+The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$.
+Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
 \end{proof}

### 2016-04-20 paper/paper.tex
Finished the text; figures next.

@@
 They generalize text indexes for graphs, allowing one to find the paths matching the query string.
-We propose using pruned de Bruijn graphs as path indexes, encoding them space-efficiently with the Burrows-Wheeler transform.
+We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform.
 We also generalize many ideas from text indexing literature to work with graphs.
@@
 
-In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
+Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
 The notion of substrings generalizes for such infinite sequences in a natural way.
@@
 [sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
-[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, out) - 1];\ \textrm{and} \\
+[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
 [sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
@@
 The results can be seen in Table~\ref{table:benchmark}.
-Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction.
+Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned before index construction.
 The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index.
@@
 
-The compressed suffix tree used in GCSA2 may be useful for indexing texts as well, as it is both simple to implement and fast in practice.
+The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice.
 As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently.
@@
 
-% FIXME from here
-% additional edges using hypertext index; haplotypes and potential recombinations
+We often have to prune complex regions of the input graph before indexing it.
+This causes false negatives: paths that exist in the input graph but not in the index.
+We could adapt techniques from the hypertext index \cite{Thachuk2013} to avoid the false negatives.
+We create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
+Then we create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph.
+We then search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index.
+For all $0 < i < \abs{X}$, we can combine the partial matches $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ into complete matches crossing one recombination edge with a range query in the edge matrix.
 
-% graph formalism, grammars
+While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately.
+For example, if sequence $S$ can occur in different positions of the genome (e.g.~$XSYZ$ and $XYSZ$), we can have a single copy or multiple copies of $S$ in the reference.
+Neither of the options is good for a reference genome.
+With a single copy, we create paths in the reference that do not correspond to any valid genome.
+With the second option, we duplicate sequences that are not only identical but also functionally the same.
+One solution is to switch from graphs to \emph{context-free grammars}.
+As long as the grammar is non-nested, we can handle it with the hypertext index.
+We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
 

### 2016-04-20 paper/paper.tex
Finished the figures; proofreading next.

@@
 
-% FIXME SA/BWT/LCP example
 \emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree.
@@
 The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree.
+See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.
+
+\begin{figure}[t!]
+\includegraphics{gcsa2_text_indexes.pdf}
+\caption{\LFmapping, BWT, suffix array, and LCP array for text $\dnaseq{GCATCATA}\$$.}\label{figure:indexes}
+\end{figure}
 
@@
 We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
-We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
-However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.
+We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
+However, this edge is not considered a real edge, and no path can cross it.
 
@@
 
-% FIXME example of strings, DBG, long paths, false positives
 \begin{definition}[de~Bruijn graph]
@@
 
-% FIXME LF(i, c) picture from the slides for the earlier DBG example
 The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
@@
 We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence.
-If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = (s, j)$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
+If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
 Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
@@
 
+See Figure~\ref{figure:graph-dbg} for an example of a de~Bruijn graph of a graph.
 We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
@@
 Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.
+
+\begin{figure}[t!]
+\includegraphics[width=396pt]{gcsa2_graph_dbg.pdf}
+\caption{Left: Input graph $G = (V, E)$, with each node $v \in V$ labeled with $v:G.\glabel(v)$.
+Right: The \orderk{3} de~Bruijn graph $G' = (V', E')$ of graph $G$, with each node $v' \in V'$ labeled with $G'.\gkey(v')$ and $G'.\gvalue(v')$.
+Both: Edges $(t, s)$ are not shown.
+The highlighted path in the de~Bruijn graph is a false positive, as it consists of two disjoint paths in the input graph.}\label{figure:graph-dbg}
+\end{figure}
 
@@
 For all $v' \in V'(v'')$, we have $G'.\gvalue(v') = G''.\gvalue(v'')$.
+See Figure~\ref{figure:pruned-index} for an example of a pruned de~Bruijn graph.
+
+\begin{figure}[t!]
+\includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
+\caption{Left: An \orderk{3} pruned de~Bruijn graph $G''$ \kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}.
+Right: GCSA for graph $G''$.
+Leftward arrows show how backward searching proceeds, with the highlighted arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$.
+Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for the node with key $\dnaseq{CAT}$.
+The keys are not stored explicitly in the index.}\label{figure:pruned-index}
+\end{figure}
 
@@
 We also extend GCSA with techniques based on suffix trees.
+See Figure~\ref{figure:pruned-index} for an example of a GCSA.
 Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$.
@@
 
-% FIXME refer to the figure in Section 2.4
 Let $G' = (V', E')$ be a path graph.
@@
 While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately.
-For example, if sequence $S$ can occur in different positions of the genome (e.g.~$XSYZ$ and $XYSZ$), we can have a single copy or multiple copies of $S$ in the reference.
+For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
 Neither of the options is good for a reference genome.

### 2016-04-20 paper/paper.tex
Added a simple CSA benchmark for the paper.

@@
 Node keys are prefixes of the labels of all paths of length at least $k$ starting from the node.
-This makes it possible to sort the nodes unambiguosly by path labels.
+This makes it possible to sort the nodes unambiguously by path labels.
 Every substring of the \kcollection{k} is the label of a path in the de~Bruijn graph, and every path label of length at most $k+1$ is a substring of the collection.
@@
 Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
-We have implemented a GCSA-based read aligner using that algoritm in vg \cite{Garrison2014-2016}.
+We have implemented a GCSA-based read aligner using that algorithm in vg \cite{Garrison2014-2016}.
 
@@
 32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
-64 & 351567 & 326101 & 0.33M & 1.35M &  142 \textmu{}s & 0.34 \textmu{}s & 0.57 \textmu{}s & 12.4 \textmu{}s \\
+64 & 351567 & 326101 & 0.33M & 1.35M & 88.9 \textmu{}s & 0.36 \textmu{}s & 0.42 \textmu{}s & 10.3 \textmu{}s \\
 \hline

### 2016-04-20 csa_builder.cpp

@@ -1,0 +1,61 @@
+/*
+  Copyright (c) 2016 Genome Research Ltd.
+
+  Author: Jouni Siren <jouni.siren@iki.fi>
+
+  Permission is hereby granted, free of charge, to any person obtaining a copy
+  of this software and associated documentation files (the "Software"), to deal
+  in the Software without restriction, including without limitation the rights
+  to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
+  copies of the Software, and to permit persons to whom the Software is
+  furnished to do so, subject to the following conditions:
+
+  The above copyright notice and this permission notice shall be included in all
+  copies or substantial portions of the Software.
+
+  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+  IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+  FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+  AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+  LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+  OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
+  SOFTWARE.
+*/
+
+#include <sdsl/suffix_arrays.hpp>
+
+#include "utils.h"
+
+using namespace gcsa;
+
+//------------------------------------------------------------------------------
+
+int
+main(int argc, char** argv)
+{
+  if(argc < 2)
+  {
+    std::cerr << "Usage: csa_builder base_name" << std::endl;
+    std::cerr << std::endl;
+    std::exit(EXIT_SUCCESS);
+  }
+
+  std::string base_name = argv[1];
+  std::string index_name = base_name + ".csa";
+
+  double start = readTimer();
+  sdsl::csa_wt<sdsl::wt_huff<>, 17> index;
+  sdsl::construct(index, base_name, 1);
+  sdsl::store_to_file(index, index_name);
+  double seconds = readTimer() - start;
+
+  double megabytes = inMegabytes(sdsl::size_in_bytes(index));
+  double bpc = sdsl::size_in_bytes(index) * 8.0 / index.size();
[…]

### 2016-04-20 csa_query.cpp

@@ -1,0 +1,127 @@
+/*
+  Copyright (c) 2016 Genome Research Ltd.
+
+  Author: Jouni Siren <jouni.siren@iki.fi>
+
+  Permission is hereby granted, free of charge, to any person obtaining a copy
+  of this software and associated documentation files (the "Software"), to deal
+  in the Software without restriction, including without limitation the rights
+  to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
+  copies of the Software, and to permit persons to whom the Software is
+  furnished to do so, subject to the following conditions:
+
+  The above copyright notice and this permission notice shall be included in all
+  copies or substantial portions of the Software.
+
+  THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+  IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+  FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+  AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+  LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+  OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
+  SOFTWARE.
+*/
+
+#include <sdsl/suffix_arrays.hpp>
+
+#include "utils.h"
+
+using namespace gcsa;
+
+//------------------------------------------------------------------------------
+
+size_type filter(std::vector<std::string>& patterns);
+
+int
+main(int argc, char** argv)
+{
+  if(argc < 3)
+  {
+    std::cerr << "usage: csa_query base_name patterns" << std::endl;
+    std::cerr << std::endl;
+    std::exit(EXIT_SUCCESS);
+  }
+
+  std::string base_name = argv[1];
+  std::string pattern_name = argv[2];
+  std::cout << "CSA query benchmark" << std::endl;
+  std::cout << std::endl;
+  printHeader("Base name"); std::cout << base_name << std::endl;
+  printHeader("Pattern file"); std::cout << pattern_name << std::endl;
+  std::cout << std::endl;
+
+  sdsl::csa_wt<sdsl::wt_huff<>, 17> index;
[…]

### 2016-04-20 paper/paper.tex
Proofreading and pruning the paper; some changes to benchmarks.

@@
 We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform.
-We also generalize many ideas from text indexing literature to work with graphs.
+We also generalize ideas from text indexing literature and use them with graphs.
 The proposed approach has been implemented in the variation graph toolkit vg.
@@
 
-Sequencing pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species.
+Sequence analysis pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species.
 As reference genomes are usually assembled from the genomes of a small number of individuals, they are biased towards those individuals.
@@
 
-Recently there has been much interest in \emph{variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
+\emph{Variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
 These graphs are eventually expected to replace reference sequences as reference genomes.
@@
 The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools.
-This paper describes \emph{GCSA2}, the path index developed for vg.
+This paper describes \emph{GCSA2}, the path index used in vg.
 A \emph{path index} is a generalization of text indexes for labeled graphs.
-They are used for finding paths with labels that match the query string.
-Indexing graphs for path queries is an inherently hard problem, as the number of paths increases exponentially with path length.
-Hence the design of a path index is a trade-off between maximum supported query length, the size of the index, and pruning complex regions of the graph before indexing.
+Given a query string, they find the paths with a label matching the query.
+Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length.
+The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.
 
@@
 Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index.
-The original GCSA could only index directed acyclic graphs and depended on pruning the complex regions, as it did not place any limits on query length.
-Kim et~al.~\cite{Kim2015-2016} combined GCSA with the ideas developed for the HISAT read aligner \cite{Kim2015}.
-The resulting HISAT2 aligner was the first practical graph-based read aligner, though it only uses the graph for more accurate mapping to a linear reference.
+The original GCSA could only index directed acyclic graphs.
+It depended on pruning the complex regions, as there were no limits on query length.
+Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}.
+The result was the first practical graph-based read aligner, though it uses the graph for more accurate mapping to a linear reference.
 
-The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} used a similar generalization of the FM\nobreakdash-index as in GCSA for representing \emph{de~Bruijn graphs} space-efficiently.
+The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index as GCSA for representing \emph{de~Bruijn graphs} space-efficiently.
 Rødland \cite{Roedland2013} proposed another similar generalization.
-The succinct de~Bruijn graphs have been generalized to represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
-Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space savings by representing unary paths in the graph as a single node.
+Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
+Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space usage by representing unary paths in the graph as single nodes.
 A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.
@@
 Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences.
-Simple variants are encoded by using the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site.
-This is usually enough to keep the index small, at the price of much higher query times.
-The hypertext index \cite{Thachuk2013} is another similar idea.
-The nodes of the graph are labeled with strings, which are indexed using an ordinary FM-index.
+Simple variants are encoded with the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site.
+Query times are much higher than in ordinary FM\nobreakdash-indexes due to the powerset alphabet.
+The hypertext index \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings.
+The strings are indexed using an ordinary FM-index.
 Partial matches in the strings are combined into full matches with range queries in the edge matrix.
-While matches crossing one edge are easy to find, the approach becomes complex and potentially slow with matches crossing multiple edges.
-There are also several indexes (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient representation for similar sequences but index the sequences instead of the graph.
+While matches crossing one edge are easy to find, the approach becomes impractical with matches crossing multiple edges.
+There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient way of indexing similar sequences.
 
@@
 A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$.
-For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text.
+We often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text.
 \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$.
@@
 Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$.
-We say that string $S'$ is a substring of string collection $\mathcal{S}$, if there is a string $S \in \mathcal{S}$ such that $S'$ is a substring of $S$.
+We say that string $S'$ is a substring of string collection $\mathcal{S}$, if it is a substring of a string $S \in \mathcal{S}$.
 
@@
 
-Given string $S[0, n-1]$, we define $S.\rank(i, c)$ to be the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
+Given string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
 We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}.
@@
 If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$.
-Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.
+Although fast and versatile, suffix trees are impractical with large texts, as they require much more space than the text itself.
 
@@
 The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}.
-Given a text and its suffix array, we can find the occurrences of \emph{pattern} string $X$ in the text by using \emph{binary search} in $\Oh{\abs{X} \log n}$ time.
-The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree.
+We can find the occurrences of \emph{pattern} string $X$ in the text in $\Oh{\abs{X} \log n}$ time by using \emph{binary search} in the suffix array.
+The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text, while its functionality is more limited than that of the suffix tree.
 See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.
@@
 
-The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text with the same combinatorial structure as the suffix array.
-Given text $T[0, n-1]$, its Burrows-Wheeler transform is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
+The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text.
+Given text $T[0, n-1]$, its BWT is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
 Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$.
@@
 
-Given the similarities to the suffix array, we can use the BWT as a space-efficient text index.
-The \emph{\FMindex} \cite{Ferragina2005a} combines a plain or compressed representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
+We can use the BWT as a space-efficient text index.
+The \emph{\FMindex} \cite{Ferragina2005a} combines a representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
 It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}.
-If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, then the lexicographic range matching suffix $X[i, \abs{X}-1]$ of the pattern is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$.
-Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ rank queries.
+If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, the range matching suffix $X[i, \abs{X}-1]$ is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$.
+Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ $\rank$ queries.
 
-We can use the sampled suffix array pointers to find the text positions containing the occurrences.
+We use the sampled suffix array pointers for finding the text positions containing the occurrences.
 If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer.
@@
 
-The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$.
-A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$.
+The graphs we use are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$.
+A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for $0 \le i < \abs{P}-1$.
 We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path.
@@
 
-We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
+We assume that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
 To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$.
-The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes.
+These characters are only used to label technical nodes similar to nodes $s$ and $t$.
 We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
@@
 
-In this paper, we work with de~Bruijn graphs and their generalizations.
-In order to do so, we need to define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.
+We work with de~Bruijn graphs and their generalizations. For that purpose, we define collections of (finite or infinite) sequences suitable for constructing \orderk{k} de~Bruijn graphs.
 
@@
 De~Bruijn graphs have several properties that make them useful for indexing purposes.
-Node keys are prefixes of the labels of all paths of length at least $k$ starting from the node.
+Node keys prefix-match the labels of all paths starting from the node.
 This makes it possible to sort the nodes unambiguously by path labels.
@@
 Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them.
-In the worst case, the transformation increases the size of the graph exponentially.
+The transformation increases the size of the graph exponentially in the worst case.
 After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences.
@@
 
-The design of a path index is a trade-off between the maximum supported query length and the size of the index.
-In the worst case, a graph may have up to $\sigma^{k}$ distinct labels for paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and the length of possible paths is unlimited in cyclic graphs.
+The design of a path index is a trade-off between maximum query length and the size of the index.
+In the worst case, a graph may have up to $\sigma^{k}$ distinct labels of paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and cyclic graphs have paths of unlimited length.
 GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary.
@@
 
-Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns.
-Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order.
-Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where pattern $X$ is a prefix of the key $X_{i}$.
+Another representation trades query performance for the ability to search for short patterns.
+We sort the key-value pairs by their keys and store them in array $A$.
+If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of the key $X_{i}$, with binary search.
 We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.
 
-As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other.
-In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$.
+We can use de~Bruijn graphs and \kmer{k} indexes to simulate each other.
+For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph.
 If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$.
-We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence.
-If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
+We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence.
+If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
 Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
 
-\begin{definition}[de~Bruijn graph of a graph]
+\begin{definition}[de~Bruijn graph of a graph]\label{def:dbg}
 Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
@@
 
-See Figure~\ref{figure:graph-dbg} for an example of a de~Bruijn graph of a graph.
-We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
-As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges.
+A \kmer{(k+1)} index of graph $G = (V, E)$ can simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
+We represent the nodes of the de~Bruijn graph with the adjacent edges.
 If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$.
 Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.
+
+We can also use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$.
+Let $\mathcal{S}$ be the \kcollection{k} in Definition~\ref{def:dbg}.
+We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ matching pattern $G'.\gkey(v)$ in graph $G$.
+There will be no false negatives, regardless of pattern length.
+There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.
+See Figure~\ref{figure:graph-dbg} for an example.
 
@@
 
-On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$ that supports queries of arbitrary length.
-Let $\mathcal{S}$ be the \kcollection{k} used for building the de~Bruijn graph.
-For each node $v \in V'$, we attach a set of nodes of graph $G$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ in graph $G$ with $G.\glabel(P) = G'.\gkey(v)$.
-Due to the nature of de~Bruijn graphs, this index will not produce any false negatives.
-There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.
-
 \subsection{Path graphs}
 
-De~Bruijn graphs are a special case of path graphs.
-They represent the paths of length $k$ in the \emph{input graph} as nodes and the pairs of paths overlapping on $k-1$ nodes of the input graph as edges.
-We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.
+Path graphs generalize de~Bruijn graphs by allowing node keys of different lengths.
 
 \begin{definition}[Path graph]
-Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings of length at most $k$ containing $\#^{k}$ and $\$^{k}$.
-We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$.
-The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
+Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
+Assume that each right-maximal subtring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that $\#^{k}, \$^{k} \in \mathcal{K}$.
+The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
 \begin{itemize}
@@
 \begin{definition}[Path graph as an index]
-Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
+Let $G' = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
 \begin{itemize}
-\item Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$, if there is a path $P$ in $G'$ with $G'.\glabel(P) = X$.
+\item Pattern $X \in \Sigma^{\ast}$ \emph{matches} node $v \in V'$, if there is a path $P'$ in $G'$ with $G'.\glabel(P') = X$.
 We use $G'.\find(X)$ to denote the set of nodes $V'_{X} \subseteq V'$ matching the pattern.
-\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = G'.\gvalue(V'_{X}) = \bigcup_{v \in V'_{X}} G'.\gvalue(v)$.
+\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v \in V'_{X}} G'.\gvalue(v)$.
 We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
@@
 Let $G' = (V', E')$ be a path graph, and let $X \in \patternset$ be a pattern.
-The set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$, which depends on the graph and the pattern.
+Set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$, which depends on the graph and the pattern.
 \end{lemma}
@@
 Leftward arrows show how backward searching proceeds, with the highlighted arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$.
-Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for the node with key $\dnaseq{CAT}$.
-The keys are not stored explicitly in the index.}\label{figure:pruned-index}
+Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
 \end{figure}

### 2016-04-20 csa_query.cpp

@@ -95,7 +95,7 @@ main(int argc, char** argv)
     printTime("locate()", ranges.size(), seconds);
     printHeader("locate()");
     std::cout << total << " occurrences (" <<
-              (total / seconds) << " / second)" << std::endl;
+              inMicroseconds(seconds / total) << " µs/occurrence)" << std::endl;
     std::cout << std::endl;
   }
 

### 2016-04-20 query_gcsa.cpp

@@ -146,7 +146,7 @@ main(int argc, char** argv)
     printTime("locate()", ranges.size(), seconds);
     printHeader("locate()");
     std::cout << total << " occurrences (" <<
-              (total / seconds) << " / second)" << std::endl;
+              inMicroseconds(seconds / total) << " µs/occurrence)" << std::endl;
     std::cout << std::endl;
   }
 

### 2016-04-21 paper/paper.tex
The paper is almost finished.

@@
 Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}.
-Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index.
-The original GCSA could only index directed acyclic graphs.
+Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs.
 It depended on pruning the complex regions, as there were no limits on query length.
@@
 
-This paper contains several lemmas describing the properties of generalizations of de~Bruijn graphs. Their proofs are included in Appendix~\ref{appendix:proofs}.
+The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendixes contain more material from the full paper.
 
@@
 \begin{lemma}[Short keys]\label{lemma:pg-keys}
-Let $G' = (V', E')$ be a path graph, where $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}$ for all edges $(u', v') \in E'$. Then
+Let $G' = (V', E')$ be a path graph with $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$. Then
 (a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$; and
@@
 \begin{lemma}[Pruning]\label{lemma:dbg-prune}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v)$.
-If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
+Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v' \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v')$.
+If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u', v' \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
 \end{lemma}
@@
 
-Path graphs are essentially equivalent to the prefix-sorted automata used in the original paper on GCSA \cite{Siren2014}.
-As many of the technical details are different, we now take a look at how a GCSA of a path graph works.
+We now show how a GCSA of a path graph works.
+As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different.
 We also extend GCSA with techniques based on suffix trees.
-See Figure~\ref{figure:pruned-index} for an example of a GCSA.
-Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$.
+See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
+Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$.
 While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.
@@
 Let $G' = (V', E')$ be a path graph.
-We sort the nodes $V'$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$.
-For each node $v' \in V'$, we append
-\begin{itemize}
-\item $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$;
-\item $\bvIN$ with the indegree $x = G.\gindegree(v')$ encoded as a binary sequence $0^{x-1} 1$; and
-\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v')$ encoded as a binary sequence $0^{x-1} 1$.
-\end{itemize}
+We sort the nodes $V'$ by their keys in lexicographic order and generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$ from the nodes in that order.
+For each node $v' \in V'$, we append $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$; $\bvIN$ with the indegree encoded as $0^{G.\gindegree(v')-1} 1$; and $\bvOUT$ with the outdegree as $0^{G.\goutdegree(v')-1} 1$.
+
 If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$.
-The labels of predecessor nodes $u'$ are encoded in $\BWT[sp_{in}, ep_{in}]$.
-Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v')$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$.
+The labels of predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$.
+Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$.
 As multiple edges may have the same sort keys, our sorting algorithm must be stable.
@@
 
-Consider now the outgoing edges.
-If the lexicographic rank of node $u' \in V'$ is $i'$, the range of outgoing edges $(u', v') \in E'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$.
+The range of outgoing edges $(u', v') \in E'$ for node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$.
 The edges are already sorted by keys $G'.\gkey(u')$.
-All outgoing edges from the same node have the same sort key, as there is nothing to distinguish them from one another.
 Because graph $G'$ is a path graph, we know that key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
 The sorting orders are therefore compatible.
-For every $j \in [sp_{in}, ep_{in}]$ for node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for node $u' \in V'$ implies an edge $(u', v') \in E'$.
+For every $j \in [sp_{in}, ep_{in}]$ for a node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for a node $u' \in V'$ implies an edge $(u', v') \in E'$.
 If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.
 
-We implement query $G'.\find(X)$ with \emph{backward searching}.
+We use \emph{backward searching} for query $G'.\find(X)$.
 Let $X \in \patternset$ be a pattern.
-If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes of the path graph.
-Now assume that $\abs{X} \ge 1$ and that query $G'.\find(X[i+1, \abs{X}-1])$ has returned range $[sp_{i+1}, ep_{i+1}]$.
-We want to find the lexicographic range corresponding to the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$.
-We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges (with $sp_{in} = 0$ if $sp_{i+1} = 0$); the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
+If $\abs{X} = 0$, query $G'.\find(X)$ returns the lexicographic range $[0, \abs{V'}-1]$ containing all nodes.
+Now assume that $\abs{X} \ge 1$ and that $G'.\find(X[i+1, \abs{X}-1]) = [sp_{i+1}, ep_{i+1}]$.
+We want to find the lexicographic range $G'.\find(X[i, \abs{X}-1])$, which is the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$.
+We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
 \begin{align*}
@@
 \end{align*}
-We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i}], X[i])$.
+We can think this as a generalization of \LFmapping: $[sp_{i}, ep_{i}] = G'.\LF([sp_{i+1}, ep_{i+1}], X[i])$.
 
-Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$ and removing duplicates.
-Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above.
-To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM\nobreakdash-indexes.
-We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind.
+Query $G'.\locate(X)$ retrieves the values $G'.\gvalue(v')$ for nodes $v' \in G'.\find(X)$ and filters out duplicates.
+Storing the values explicitly for all nodes would require too much space.
+Like FM\nobreakdash-indexes, GCSA uses a \emph{sampling} scheme to avoid that.
+We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are conveniently chosen integers.
 If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.
 
-Let $v' \in V'$ be a node of the path graph.
-We sample the values $G'.\gvalue(v')$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $(u', v') \in E'$ is the only incoming edge to node $v'$, and $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$.
+We sample the values $G'.\gvalue(v')$ for a node $v' \in V'$, (a) if there are multiple incoming edges to node $v'$; (b) if $v'$ is the source node $s$; or (c) if $G'.\gvalue(v') \ne \set{u+1 \mid u \in G'.\gvalue(u')}$ for the only incoming edge $(u', v') \in E'$.
 We may also sample the values for some nodes on long unary paths for performance reasons.
-If the values $G'.\gvalue(v')$ has not been sampled, we can derive them from sampled values by following the incoming edges backwards.
+If the values $G'.\gvalue(v')$ have not been sampled, we can derive them from sampled values by following the incoming edges.
 
-Following edges backwards is based on a similar generalization of \LFmapping{} as backward searching. If node $v' \in V'$ has lexicographic rank $i$, the lexicographic rank of its only predecessor is
-$G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$.
+If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$ .
 If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.
 
-Let $B_{S}[0, \abs{V'}-1]$ be a bitvector marking the sampled nodes.
-If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S} = 1$.
+Let $B_{S}[0, \abs{V'}-1]$ be a bitvector.
+If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S}[i] = 1$.
 We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$.
 For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$.
-We then store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each.
+We store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each.
 The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.
-
-If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph.
-We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
-If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$.
-Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
-$$
-G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
-$$
-We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector.
-On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
-The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.
 
@@
 
-GCSA is a generalization of FM\nobreakdash-indexes for path graphs.
-Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.
+GCSA is a generalization of the FM\nobreakdash-index for path graphs.
+Because path graphs are similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them for path graphs.
 
-Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
+Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys.
 We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
 The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
-The compacted trie can be considered the \emph{suffix tree of the path graph}.
+The compacted trie can be considered the \emph{suffix tree} of graph $G'$.
 
@@
 
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
 Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
@@
 BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference.
-It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions.
-GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$).
-Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
-We have implemented a GCSA-based read aligner using that algorithm in vg \cite{Garrison2014-2016}.
+It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions.
+GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$).
+However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
 
 In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
-In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes.
-Determining the  number of distinct start nodes with a $\locate$ query can be slow.
+In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes.
+Determining the number of distinct values with a $\locate$ query can be slow.
 Hence we need additional structures to support efficient \emph{counting queries}.
@@
 Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
-We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+We can augment an FM\nobreakdash-index with a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
 
-Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
-If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
-Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times.
-We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$.
+Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
+If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, the number of \emph{redundant} documents in them is $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
+We create an array $R[0, n-2]$ based on the \emph{inorder traversal} of the suffix tree.
 If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
-For subsequent visits to the same node, we set $R[j] = 0$.
-Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$.
-We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$.
-If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
+We set $R[j] = 0$ for any subsequent visits to the same node.
+Range $R[sp, ep-1]$ covers the internal nodes in the subtree with node $v$ as the root.
+We can determine document frequencies as $\countq(v) = (ep+1-sp) - \sum R[sp, ep-1]$.
+If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute the sums with $\select$ queries:
 $$
-\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
+\sum R[a, b] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
 $$
 
-For value counting, we use array $R$ to store the number of redundant values in each suffix tree node.
-We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$.
-We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above.
-The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
+For value counting, we use array $R$ to store the number of redundant values as above.
+We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above.
+The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
 $$
-G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
+G'.\countq(X) = G'.\countq([sp, ep]) = (ep+1-sp) + \sum A[sp, ep] - \sum R[sp, ep-1].
 $$
-where $[sp, ep]$ is the lexicographic range returned by the $\find$ query.
 The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
@@
 
-GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}.
-The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
-As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+GCSA2 has been developed as the path index library of vg \cite{Garrison2014-2016}.
+The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
+As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
 As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.
@@
 The \emph{variation graphs} used in vg have strings as node labels.
-The average length of a node label is typically around 10~bases.
-Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
+The average length of a label is typically around 10~bases.
+Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
 When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
@@
 
-We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
-To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases.
-We then extracted paths of length $16$ from the forward strands of all chromosomes.
+We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
+To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial with \texttt{vg mod -p -l -e 4} and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
+We extracted paths of length $16$ from the forward strands of all chromosomes.
 There were a total of 4.80~billion paths with 1.53~billion distinct labels.
-We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively.
+We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
 The results are summarized in Table~\ref{table:construction}.
+We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
+The de~Bruijn graph had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes.
+The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.
 
@@
 
-A single-strand whole-genome index can be built overnight with 96~gigabytes of memory.
-Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower.
-The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
-It consists of a pruning step, followed by merging the paths with identical labels.} The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.
-
-We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread.
+We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread.
 The results can be seen in Table~\ref{table:benchmark}.
-Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned before index construction.
-The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index.
-There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out.
-Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.
+The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index.
+We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries generate duplicates, which must be filtered out.
+A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
 
@@
 \begin{center}
-\caption{Performance benchmarks using an \orderk{64} GCSA2.
-Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
+\caption{Query benchmarks using an \orderk{64} index.
+Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
 \begin{tabular}{c|cc|cc|cccc}
@@
 GCSA2 is a path index for variation graphs.
-It uses a structurally compressed de~Bruijn graph of the paths in the input graph as an index, and encodes that with a generalization of the FM\nobreakdash-index.
-The index also incorporates extensions based on a generalization of suffix trees.
-GCSA2 has been integrated in the variation graph toolkit vg, where it is used e.g.~for read alignment based on maximal exact matches.
+It uses a pruned de~Bruijn graph as an index of the variation graph, and encodes that with a generalization of the FM\nobreakdash-index.
+The index includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
+GCSA2 has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
 
-We can build a forward strand whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The index itself requires less than 16~gigabytes of memory.
-Query performance is roughly 5\nobreakdash--10 times slower than for a fast FM\nobreakdash-index for sequences.
+We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
+The resulting index takes less than 16~gigabytes.
+Query performance is several times slower than with a fast FM\nobreakdash-index for sequences.
 GCSA2 can search for patterns at roughly 0.8\nobreakdash--1.0~MB/s and locate 100,000\nobreakdash--300,000 occurrences per second using a single CPU core.
-As with other FM\nobreakdash-indexes, the speedup from distributing the queries on multiple cores is quite high.
 
@@
 
-We often have to prune complex regions of the input graph before indexing it.
-This causes false negatives: paths that exist in the input graph but not in the index.
-We could adapt techniques from the hypertext index \cite{Thachuk2013} to avoid the false negatives.
-We create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
-Then we create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph.
-We then search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index.
-For all $0 < i < \abs{X}$, we can combine the partial matches $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ into complete matches crossing one recombination edge with a range query in the edge matrix.
-
-While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately.
-For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
-Neither of the options is good for a reference genome.
-With a single copy, we create paths in the reference that do not correspond to any valid genome.
-With the second option, we duplicate sequences that are not only identical but also functionally the same.
-One solution is to switch from graphs to \emph{context-free grammars}.
-As long as the grammar is non-nested, we can handle it with the hypertext index.
-We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
-
 \subparagraph*{Acknowledgements.}
 
-I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index.
-Many extensions over the original GCSA and clarifications of the assumptions I made are a direct result of those discussions.
+I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index, and for the new ideas I got from the discussions.
 
@@
 
-The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.
-First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
+The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
+It consists of a pruning step, followed by merging the paths with identical labels.} First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
 If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$.
@@
 
+\section{Alternate GCSA Encoding}\label{appendix:encoding}
+
+Let $G' = (V', E')$ be a path graph with $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$.
+This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}).
+We can use the simplified encoding of the original GCSA \cite{Siren2014} for the GCSA of such path graphs.
+
+We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
+If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$.
+Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
+$$
+G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
+$$
+We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector.
+On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
+The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.
+
+\section{Using the Hypertext Index}\label{appendix:hypertext}
+
+The hypertext index \cite{Thachuk2013} is based on graphs $G = (V, E)$, where the label $G.\glabel(v)$ of a node $v \in V$ is a string over alphabet $\Sigma$.
+The labels are indexes in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$.
+When we query for patterns $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross the edges.
+In order to find matches crossing one edge, we search for suffixes $X[i, \abs{X}-1]$ in the forward index $F$ and reverses of prefixes $X[0, i-1]$ in the reverse index $R$, for all $1 \le i \le \abs{X}-1$.
+For each value of $i$, we then combine the partial matches into complete matches with a two-dimensional range query $R.\find(X[0, i-1]) \times F.\find(X[i, \abs{X}-1])$ in the edge matrix $E$.
+Matches crossing multiple edges can be found with similar but more complicated techniques.
+
+We often have to prune complex regions of the input graph before indexing it with GCSA.
+This causes false negatives: paths that exist in the input graph but not in the index.
+We can generalize the hypertext index to avoid the false negatives.
+Instead of pruning the input graph heuristically, we can create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
+We then create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph, always crossing from the reverse complement strand to the forward strand.
+We search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index, and combine the results $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ with a range query for all $i$.
+
+While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately.
+For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
+Neither of the options is good for a reference genome.
+With a single copy, we create paths in the reference that do not correspond to any valid genome.
+With the second option, we duplicate sequences that are not only identical but also functionally the same.
+One solution is to switch from graphs to \emph{context-free grammars}.
+As long as the grammar is non-nested, we can handle it with the hypertext index.
+We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
+
 \end{document}

### 2016-04-21 paper/paper.tex
First submitted version.

@@
 
-The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendixes contain more material from the full paper.
+The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendices contain more material from the full paper.
 
@@
 
-GCSA is a generalization of the FM\nobreakdash-index for path graphs.
-Because path graphs are similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them for path graphs.
-
 Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys.
-We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
+We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
 The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
@@
 In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes.
-Determining the number of distinct values with a $\locate$ query can be slow.
-Hence we need additional structures to support efficient \emph{counting queries}.
+As $\locate$ queries can be slow, we need another way to support efficient \emph{counting queries}.
 The problem is similar to determining the \emph{document frequency} of a pattern.
 Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
-We can augment an FM\nobreakdash-index with a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
 
@@
 The \emph{variation graphs} used in vg have strings as node labels.
-The average length of a label is typically around 10~bases.
 Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
@@
 The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index.
-We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries generate duplicates, which must be filtered out.
+We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries have to filter out duplicates.
 A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
@@
 \caption{Query benchmarks using an \orderk{64} index.
-Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
+Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
 \begin{tabular}{c|cc|cc|cccc}
@@
 
-For all positions $j$ with $i \le j < i+\abs{X}$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
+For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
 By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
@@
 Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
-By Lemma~\ref{lemma:dbg-fp}, $X$ is a prefix of the substring.
+By Lemma~\ref{lemma:pg-keys}, $X$ is a prefix of the substring.
 We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
@@
 Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
-As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G''.\gkey(v''_{j+1})$, which is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, key $G''.\gkey(v''_{j})$ is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
+As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, it is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
 When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
@@
 Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}.
-Because graph $G'$ is a pruned de~Bruijn graph, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
+Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
 As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.
@@
 Let $G = (V, E)$ be the input graph.
-We extract all paths of length $k$ (typically for $k = 16$) from graph $G$.
+We extract all paths of length $k$ (typically with $k = 16$) from graph $G$.
 For each path $P = v_{0} \dotsm v_{\abs{P}-1}$, we store several fields.
@@
 We store the set of \emph{predecessor labels} $\set{c \in \Sigma \mid \abs{G.\gpred(v_{0}, c)} > 0}$ as $P.\gpred$.
-For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a new copy of the path and store the node as $P.\gext = v$.
+For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a separate copy of the path and store the node as $P.\gext = v$.
 
@@
 Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys.
-We use two additional arrays: \emph{the LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$.
-The LCP array is stored in a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.
+We use two additional arrays: the \emph{LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$.
+The LCP array is stored as a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.
 
@@
 
-The extension step transforms the current set of paths of length (up to) $k'$ and transforms it into a set of paths of length (up to) $2k'$.
+The extension step transforms the current set of paths of length (up to) $k'$ into a set of paths of length (up to) $2k'$.
 If $P$ is a path with $P.\gext = -1$, we use it as such.
@@
 
-The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
+The doubling steps are followed by the \emph{merging step}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
 It consists of a pruning step, followed by merging the paths with identical labels.} First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
@@
 Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order.
-A doubling step can be done separately for each of the chromosomes.
-A pruning step takes a set of sorted lists paths and merges the lists.
+An extension step can be done separately for each of the chromosomes.
+A pruning step takes a set of sorted lists of paths and merges the lists.
 It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged.
@@
 
-After storing the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$ on disk, we proceed to \emph{building the index}.
+After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we proceed to \emph{building the index}.
 Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing.
@@
 We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
-If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$.
+If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
 Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
@@
 The labels are indexes in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$.
-When we query for patterns $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross the edges.
-In order to find matches crossing one edge, we search for suffixes $X[i, \abs{X}-1]$ in the forward index $F$ and reverses of prefixes $X[0, i-1]$ in the reverse index $R$, for all $1 \le i \le \abs{X}-1$.
-For each value of $i$, we then combine the partial matches into complete matches with a two-dimensional range query $R.\find(X[0, i-1]) \times F.\find(X[i, \abs{X}-1])$ in the edge matrix $E$.
-Matches crossing multiple edges can be found with similar but more complicated techniques.
+When we search for a pattern $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross edges.
+In order to find matches crossing one edge, we search for suffixes $X[i, \abs{X}-1]$ in the forward index $F$ and the reverses of prefixes $X[0, i-1]$ in the reverse index $R$, for all $1 \le i \le \abs{X}-1$.
+For each value of $i$, we combine the partial matches into complete matches with a two-dimensional range query in the edge matrix $E$, using the lexicographic ranges for the reverse of $X[0, i-1]$ and for $X[i, \abs{X}-1]$ as the query ranges.
 
@@
 This causes false negatives: paths that exist in the input graph but not in the index.
-We can generalize the hypertext index to avoid the false negatives.
-Instead of pruning the input graph heuristically, we can create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
+We can avoid the false negatives with a generalization of the hypertext index.
+Instead of pruning the input graph heuristically, we create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
 We then create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph, always crossing from the reverse complement strand to the forward strand.
-We search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index, and combine the results $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ with a range query for all $i$.
+We search for pattern $X$ and its reverse complement $\revcomp{X}$ in the index, and combine the results $\find(X[i, \abs{X}-1])$ and $\find(\revcomp{X}[\abs{X}-i, \abs{X}-1])$ with a range query.
 
-While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately.
+While graphs are a natural formalism for representing genetic variation, they cannot adequately represent certain types of \emph{rearrangements}.
 For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
-Neither of the options is good for a reference genome.
+Neither option is particularly good in a reference genome.
 With a single copy, we create paths in the reference that do not correspond to any valid genome.
@@
 As long as the grammar is non-nested, we can handle it with the hypertext index.
-We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
+We build a hypertext index for a high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
 

### 2016-04-21 paper/paper.tex
Updated submission.

@@
 Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1}) = G_{d}.\gvalue(v_{j+1})$.
-Therefore substring $X[j] \cdot G_{d}.\gkey(v_{j+1})$ exists in $\mathcal{S}$, and the node $v_{j} \in V_{d}$ with $G_{d}.\gkey(v_{j}) = (X[j] \cdot G_{d}.\gkey(v_{j+1}))[0, k-1]$ has $X[j, \abs{X}-1]$ as a prefix of its key.
-As $G''.\gkey(v''_{j})$ is a prefix of $X[j] \cdot G_{d}.\gkey(v_{j+1})$, it is also a prefix of $G_{d}.\gkey(v_{j})$, and hence $v_{j} \in V_{d}(v''_{j})$.
-When $j = 0$, we can extend the substring as above to get a path matching the pattern in $G_{d}$, with $v_{0}$ as the start node.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$.
+Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.
+
+Let $Y$ be the longer of strings $G''.\gkey(v''_{j})$ and $X[j] \cdot G''.\gkey(v''_{j+1})$.
+As $Y$ is a substring of $\mathcal{S}$, there is a node $v_{j} \in V_{d}$ such that key $G_{d}.\gkey(v_{j})$ prefix-matches $Y$.
+Key $G_{d}.\gkey(v_{j})$ has both $X[j, \abs{X}-1]$ and $G''.\gkey(v''_{j})$ as its prefixes, and hence $v_{j} \in V_{d}(v''_{j})$.
+When $j = 0$, we can extend key $G_{d}.\gkey(v_{0})$ to get a path starting from $v_{0}$ and matching $X$ in graph $G_{d}$.
 

### 2016-04-21 paper/paper.tex
Further proofreading.

@@
 
-The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index as GCSA for representing \emph{de~Bruijn graphs} space-efficiently.
+The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently.
 Rødland \cite{Roedland2013} proposed another similar generalization.
@@
 
-The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendices contain more material from the full paper.
+The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendices contain material from the full paper.
 
@@
 Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
-Assume that each right-maximal subtring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that $\#^{k}, \$^{k} \in \mathcal{K}$.
+Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that $\#^{k}, \$^{k} \in \mathcal{K}$.
 The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where

### 2016-04-22 paper/paper.tex
Comparison to FM-index.

@@
 
-The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}. The other appendices contain material from the full paper.
+The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}.
 
@@
 
-\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree.
+\emph{Suffix arrays} (SA) \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree.
 The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}.
-We can find the occurrences of \emph{pattern} string $X$ in the text in $\Oh{\abs{X} \log n}$ time by using \emph{binary search} in the suffix array.
+We find the occurrences of \emph{pattern} string $X$ in the text in $\Oh{\abs{X} \log n}$ time by using \emph{binary search} in the suffix array.
 The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text, while its functionality is more limited than that of the suffix tree.
@@
 As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.
+We chose fast succinct data structures over the smaller but slower compressed data structures for most index components.
+The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0 for the experiments.
+All files were stored on a distributed Lustre file system.
+We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the queries.
 
-The \emph{variation graphs} used in vg have strings as node labels.
-Each node can be traversed in both forward and reverse complement orientations, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around.
+The vg \emph{variation graphs} have strings as node labels.
+Each node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
 When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
-Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.
-
-The following experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
-Input files, final indexes, and temporary files were all stored on a distributed Lustre file system.
-We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the query benchmarks.
+Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples.
 
@@
 We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread.
+We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference.
 The results can be seen in Table~\ref{table:benchmark}.
-The $\find$ and $\locate$ queries are several times slower than in a fast whole-genome FM\nobreakdash-index.
-We have to use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}, while our $\locate$ queries have to filter out duplicates.
+The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components.
 A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
@@
 \begin{center}
-\caption{Query benchmarks using an \orderk{64} index.
-Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
-\begin{tabular}{c|cc|cc|cccc}
+\caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index.
+Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
+\begin{tabular}{c|cccc|cccc}
 \hline
-\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+ & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
 \hline
-16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
-32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
-64 & 351567 & 326101 & 0.33M & 1.35M & 88.9 \textmu{}s & 0.36 \textmu{}s & 0.42 \textmu{}s & 10.3 \textmu{}s \\
+GCSA: 16 & 351584 & 347453 & 1829M &  872M & 17.9 \textmu{}s & 0.51 \textmu{}s & 0.93 \textmu{}s & 10.3 \textmu{}s \\
+GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.6 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
+GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 74.6 \textmu{}s & 0.25 \textmu{}s & 0.32 \textmu{}s & 1.65 \textmu{}s \\
+\hline
+FM:   16 & 351584 & 301538 &    -- &  782M & 7.83 \textmu{}s &              -- &              -- & 3.10 \textmu{}s \\
+FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.16 \textmu{}s \\
+FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.89 \textmu{}s \\
 \hline
@@
 Query performance is several times slower than with a fast FM\nobreakdash-index for sequences.
-GCSA2 can search for patterns at roughly 0.8\nobreakdash--1.0~MB/s and locate 100,000\nobreakdash--300,000 occurrences per second using a single CPU core.
 

### 2016-04-22 paper/paper.tex
Final submission with updated benchmarks.

@@
 
-\emph{Variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
+\emph{Variation graphs} (also called pan-genomes, graph genomes, and reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
 These graphs are eventually expected to replace reference sequences as reference genomes.
@@
 A \emph{path index} is a generalization of text indexes for labeled graphs.
-Given a query string, they find the paths with a label matching the query.
+Given a query string, it finds the paths with a label matching the query.
 Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length.
@@
 
-The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently.
+The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently.
 Rødland \cite{Roedland2013} proposed another similar generalization.
@@
 We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
-A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.
+A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix, and \emph{right-maximal} if it is right-infinite or a suffix.
 
@@
 
-We generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
-We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
-
 We assume that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
@@
 However, this edge is not considered a real edge, and no path can cross it.
+
+We generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
+We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
 
@@
 
-Suffix trees, suffix arrays, and \FMindex{} can be generalized to index multiple texts.
+Suffix trees, suffix arrays, and the \FMindex{} can be generalized to index multiple texts.
 There are also generalizations for other combinatorial structures.
@@
 $\BWT$ stores the labels of the children of each node (leaf nodes require special treatment).
-If a node has $k$ children, we encode that as binary sequence $0^{k-1} 1$.
+If a node has $k$ children, we encode that as a binary sequence $0^{k-1} 1$.
 We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges.
-The labels of the children of node $i$ are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.
+The labels of the children of the $i$th node are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.
 
@@
 We sort the key-value pairs by their keys and store them in array $A$.
-If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of the key $X_{i}$, with binary search.
+If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of key $X_{i}$, with binary search.
 We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.
@@
 If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
-Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
+Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = t:j$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
 
-\begin{definition}[de~Bruijn graph of a graph]\label{def:dbg}
+\begin{definition}[De~Bruijn graph of a graph]\label{def:dbg}
 Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
@@
 \begin{definition}[Path graph as an index]
-Let $G' = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
+Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
 \begin{itemize}
@@
 We use $G'.\find(X)$ to denote the set of nodes $V'_{X} \subseteq V'$ matching the pattern.
-\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v \in V'_{X}} G'.\gvalue(v)$.
+\item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v' \in V'_{X}} G'.\gvalue(v')$.
 We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
@@
 \begin{definition}[Equivalent path graphs]
-Let $G$ and $G'$ be two path graphs, and let $k > 0$ be a parameter value.
-We say that graphs $G$ and $G'$ are \emph{\kequivalent{k}}, if we have $G.\locate(X) = G'.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
+Let $G'$ and $G''$ be two path graphs, and let $k > 0$ be a parameter value.
+We say that graphs $G'$ and $G''$ are \emph{\kequivalent{k}}, if we have $G'.\locate(X) = G''.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
 \end{definition}
@@
 \begin{definition}[Maximally pruned de~Bruijn graph]
-Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it using Lemma~\ref{lemma:dbg-prune} any further.
+Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
 \end{definition}
@@
 As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different.
+See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
 We also extend GCSA with techniques based on suffix trees.
-See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
 Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$.
@@
 
-If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$.
-The labels of predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$.
+If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to that node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$.
+The labels of the predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$.
 Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$.
@@
 
-The range of outgoing edges $(u', v') \in E'$ for node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$.
+The range of outgoing edges $(u', v') \in E'$ from node $u' \in V'$ with lexicographic rank $i'$ is $[sp_{out}, ep_{out}] = [\bvOUT.\select(i', 1) + 1, \bvOUT.\select(i'+1, 1)]$.
 The edges are already sorted by keys $G'.\gkey(u')$.
@@
 We want to find the lexicographic range $G'.\find(X[i, \abs{X}-1])$, which is the union of sets $G'.\gpred(v', X[i])$ over nodes $v' \in G'.\find(X[i+1, \abs{X}-1])$.
-We map the node range $[sp_{i+1}, ep_{i+1}]$ to range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}]$ of nodes $G'.\find(X[i, \abs{X}-1])$:
+We map the node range $[sp_{i+1}, ep_{i+1}]$ to the range $[sp_{in}, ep_{in}]$ of incoming edges; the incoming edges to the corresponding range of outgoing edges $[sp_{out}, ep_{out}]$; and the outgoing edges to the range $[sp_{i}, ep_{i}] = G'.\find(X[i, \abs{X}-1])$:
 \begin{align*}
@@
 We may also sample the values for some nodes on long unary paths for performance reasons.
-If the values $G'.\gvalue(v')$ have not been sampled, we can derive them from sampled values by following the incoming edges.
+If the set $G'.\gvalue(v')$ has not been sampled, we can derive it from sampled values by following the incoming edges.
 
-If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$ .
+If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$.
 If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.
@@
 
-Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by the keys.
+Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
 We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
@@
 
-BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference.
+BWA-MEM \cite{Li2013} aligns reads to a reference genome by finding \emph{maximal exact matches} between the read and the reference.
 It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions.
@@
 
-GCSA2 has been developed as the path index library of vg \cite{Garrison2014-2016}.
+GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}.
 The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
 As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-We chose fast succinct data structures over the smaller but slower compressed data structures for most index components.
-The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0 for the experiments.
+The index uses fast succinct data structures instead of the smaller but slower compressed data structures in most of its components.
+The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
 All files were stored on a distributed Lustre file system.
@@
 We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
-To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial with \texttt{vg mod -p -l -e 4} and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
-We extracted paths of length $16$ from the forward strands of all chromosomes.
+To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
+We extracted all paths of length $16$ from the forward strands of the graphs.
 There were a total of 4.80~billion paths with 1.53~billion distinct labels.
@@
 We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
-The de~Bruijn graph had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes.
+The de~Bruijn graphs had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes.
 The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.
@@
 \caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index.
-Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
+Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
 \begin{tabular}{c|cccc|cccc}
 \hline
- & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+ & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
 \hline
-GCSA: 16 & 351584 & 347453 & 1829M &  872M & 17.9 \textmu{}s & 0.51 \textmu{}s & 0.93 \textmu{}s & 10.3 \textmu{}s \\
-GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.6 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
-GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 74.6 \textmu{}s & 0.25 \textmu{}s & 0.32 \textmu{}s & 1.65 \textmu{}s \\
+GCSA: 16 & 351584 & 347453 & 1829M &  872M & 18.0 \textmu{}s & 0.40 \textmu{}s & 0.93 \textmu{}s & 8.61 \textmu{}s \\
+GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.7 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
+GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 75.3 \textmu{}s & 0.25 \textmu{}s & 0.33 \textmu{}s & 1.66 \textmu{}s \\
 \hline
-FM:   16 & 351584 & 301538 &    -- &  782M & 7.83 \textmu{}s &              -- &              -- & 3.10 \textmu{}s \\
-FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.16 \textmu{}s \\
-FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.89 \textmu{}s \\
+FM:   16 & 351584 & 301538 &    -- &  782M & 6.04 \textmu{}s &              -- &              -- & 2.44 \textmu{}s \\
+FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.17 \textmu{}s \\
+FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.92 \textmu{}s \\
 \hline
@@
 Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
-Hence $u' \in M_{i}$ if and only if key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$.
+Hence if $u' \in M_{i}$, key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$.
 If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
@@
 Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
-As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ for all nodes $v \in G'.\gvalue(v')$.
+As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.
 
@@
 We build an \orderk{k} \emph{de~Bruijn graph} $G_{d} = (V_{d}, E_{d})$ of the path labels and encode it as a GCSA, using the predecessor labels $P.\gpred$ for determining the edges.
-Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys.
+Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by their keys.
 We use two additional arrays: the \emph{LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$.
@@
 We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers.
-The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a $c \in \Sigma$.
+The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a character $c \in \Sigma$.
 We can also sample the nodes for $\locate$ queries during the same scans.

### 2016-06-21 paper/paper.tex
Updated the introduction and some references.

@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Acces}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+%\EventEditors{John Q. Open and Joan R. Acces}
+%\EventNoEds{2}
+%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+%\EventShortTitle{CVIT 2016}
+%\EventAcronym{CVIT}
+%\EventYear{2016}
+%\EventDate{December 24--27, 2016}
+%\EventLocation{Little Whinging, United Kingdom}
+%\EventLogo{}
+%\SeriesVolume{42}
+\ArticleNo{1}
 % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\emph{Variation graphs} (also called pan-genomes, graph genomes, and reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
-These graphs are eventually expected to replace reference sequences as reference genomes.
+\emph{Variation graphs} (also called genome graphs, graph references, and reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
+These graphs are expected to replace reference sequences as reference genomes.
 The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.
@@
 
-Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences.
-Simple variants are encoded with the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site.
-Query times are much higher than in ordinary FM\nobreakdash-indexes due to the powerset alphabet.
-The hypertext index \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings.
+Some path indexes store the graph as a collection of sequences.
+BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding simple variants and creates new sequences with a sufficient amount of context for complex variants.
+Queries are slower than in ordinary FM\nobreakdash-indexes due to using the powerset alphabet.
+vBWT \cite{Maciuca2016} encodes variant sites explicitly in the sequence as  $X(A|B|C)Y$, using distinct separator symbols for each site.
+As the variants must be taken into account during search, queries are slower than in ordinary FM\nobreakdash-indexes.
+The \emph{hypertext index} \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings.
 The strings are indexed using an ordinary FM-index.
@@
 While matches crossing one edge are easy to find, the approach becomes impractical with matches crossing multiple edges.
-There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient way of indexing similar sequences.
+
+There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015,Na2016}) using graphs as a space-efficient way of indexing similar sequences.
+While the problem is different from indexing the paths in a graph, the techniques used are similar.
+
+The above methods can be classified in three categories based on the data models they use.
+Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation.
+Methods based on GCSA can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}.
+GCSA2, the hypertext index, and methods based on de~Bruijn graphs work with \emph{arbitrary graphs}.
 
@@
 
-Let $S$ be a string and $S'$ be be a string or an infinite character sequence over alphabet $\Sigma$.
+Let $S$ be a string and $S'$ be a string or an infinite character sequence over alphabet $\Sigma$.
 We say that sequences $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
@@
 By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
-Hence we can safely use $\parent$ queries with such ranges.
+Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\parent$ query is similar to the $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014}.}
 

### 2016-06-21 paper/paper.tex
Tweaks to Introduction; more detailed index size breakdown.

@@
 
-\emph{Variation graphs} (also called genome graphs, graph references, and reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
+\emph{Variation graphs} (also graph genomes, genome graphs, graph references, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
 These graphs are expected to replace reference sequences as reference genomes.
-The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.
+The shift to graph references will likely take years, as new methods and tools are needed to replace those based on linear references.
 
@@
 Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs.
-It depended on pruning the complex regions, as there were no limits on query length.
+It depends on pruning the complex regions, as there are no limits on query length.
 Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}.
@@
 The pruned graph is then encoded with a generalization of the FM\nobreakdash-index.
-GCSA2 includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
+With two layers of compression, the index often uses less than 1~bit per \kmer{k} for large values of $k$.
+Similar space usage cannot be achieved with the other de~Bruijn graph-based approaches, which explicitly store information for each \kmer{k}.
+GCSA2 also includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
 These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
 
-The proofs of lemmas describing the properties of generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}.
+The proofs of lemmas describing the properties of the generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}.
 

### 2016-06-21 build_gcsa.cpp

@@ -177,13 +177,21 @@ main(int argc, char** argv)
   std::cout << index.sampleCount() << " (at " << index.sampledPositions() << " positions, "
             << index.sampleBits() << " bits each)" << std::endl;
   printHeader("Max query"); std::cout << index.order() << std::endl;
+  std::cout << std::endl;
 
   size_type index_bytes = sdsl::size_in_bytes(index);
-  size_type sample_bytes = sdsl::size_in_bytes(index.stored_samples);
+  size_type sample_bytes =
+    sdsl::size_in_bytes(index.sampled_paths) + sdsl::size_in_bytes(index.sampled_path_rank) +
+    sdsl::size_in_bytes(index.stored_samples) +
+    sdsl::size_in_bytes(index.samples) + sdsl::size_in_bytes(index.sample_select);
+  size_type counter_bytes =
+    sdsl::size_in_bytes(index.extra_pointers) + sdsl::size_in_bytes(index.redundant_pointers);
   size_type lcp_bytes = sdsl::size_in_bytes(lcp);
-  printHeader("Index size"); std::cout << inMegabytes(index_bytes) << " MB" << std::endl;
-  printHeader("Without samples"); std::cout << inMegabytes(index_bytes - sample_bytes) << " MB" << std::endl;
-  printHeader("LCP size"); std::cout << inMegabytes(lcp_bytes) << " MB" << std::endl;
+  printHeader("Core index"); std::cout << inMegabytes(index_bytes - sample_bytes - counter_bytes) << " MB" << std::endl;
+  printHeader("Samples"); std::cout << inMegabytes(sample_bytes) << " MB" << std::endl;
+  printHeader("Counter"); std::cout << inMegabytes(counter_bytes) << " MB" << std::endl;
+  printHeader("LCP array"); std::cout << inMegabytes(lcp_bytes) << " MB" << std::endl;
+  printHeader("Total size"); std::cout << inMegabytes(index_bytes + lcp_bytes) << " MB" << std::endl;
   std::cout << std::endl;
 
   if(verify) { verifyIndex(index, &lcp, graph); }

### 2016-06-22 paper/paper.tex
Further updates to the manuscript.

@@
 Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation.
-Methods based on GCSA can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}.
-GCSA2, the hypertext index, and methods based on de~Bruijn graphs work with \emph{arbitrary graphs}.
+GCSA-based methods can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}.
+GCSA2, the hypertext index, and indexes based on de~Bruijn graphs work with \emph{arbitrary graphs}.
 
@@
 The pruned graph is then encoded with a generalization of the FM\nobreakdash-index.
-With two layers of compression, the index often uses less than 1~bit per \kmer{k} for large values of $k$.
+The index often uses less than 1~bit per \kmer{k} with large values of $k$.
 Similar space usage cannot be achieved with the other de~Bruijn graph-based approaches, which explicitly store information for each \kmer{k}.
@@
 Right: GCSA for graph $G''$.
-Leftward arrows show how backward searching proceeds, with the highlighted arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$.
-Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
+Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$.
+Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
 \end{figure}
@@
 Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$.
-As multiple edges may have the same sort keys, our sorting algorithm must be stable.
+As multiple edges may have the same sort key, our sorting algorithm must be stable.
 We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.
@@
 
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+By selecting $x$ according to cache line size, the queries require $\Oh{\log_{x} \abs{V'}}$ random memory accesses, which in practice dominate the computation time.
 Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
 By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
-Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\parent$ query is similar to the $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014}.}
+Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}
 
@@
 We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The resulting index takes less than 16~gigabytes.
-Query performance is several times slower than with a fast FM\nobreakdash-index for sequences.
+The resulting index takes less than 16~gigabytes, which is less than what the existing path indexes capable of indexing arbitrary graphs require.
+Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
+
+GCSA2 improves upon the existing path indexes in the following ways:
+\begin{itemize}
+
+\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques, but they can only index graphs that have finite \orderk{\infty} path graphs.
+On the other hand, they support queries of arbitrary length.
+
+\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than GCSA2. The current implementation can only handle local variation.
+
+\item vBWT \cite{Maciuca2016} is slower than BWBBLE. It can only handle local variation but supports queries of arbitrary length.
+
+\item Succinct \cite{Bowe2012}, compacted \cite{Cazaux2014}, and compressed \cite{Marcus2014} de~Bruijn graphs must store at least one character for each \kmer{k} and hence require more space than GCSA2.
+
+\item The hypertext index \cite{Thachuk2013} has not been implemented. It supports queries of arbitrary length, but becomes slow with matches crossing more than one edge.
+
+\end{itemize}
 

### 2016-06-23 paper/paper.tex
Fixed a proof; mentioned probabilistic DBGs

@@
 Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
-Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space usage by representing unary paths in the graph as single nodes.
+Compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs achieve similar space usage by representing unary paths in the graph as single nodes.
 A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.
+Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to store the graph.
+They support faster queries in similar space at the expense of producing false positives.
 
@@
 The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely.
-The pruned graph is then encoded with a generalization of the FM\nobreakdash-index.
-The index often uses less than 1~bit per \kmer{k} with large values of $k$.
-Similar space usage cannot be achieved with the other de~Bruijn graph-based approaches, which explicitly store information for each \kmer{k}.
-GCSA2 also includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
-These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
+The pruned graph is encoded with a generalization of the FM\nobreakdash-index.
+GCSA2 often uses less than 1~bit per \kmer{k} with large values of $k$.
+Other de~Bruijn graph-based path indexes cannot achieve similar space usage, as they store some information for each \kmer{k} explicitly.
+GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
+The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
 
+This paper describes the GCSA2 data structure.
+Its uses in the vg toolkit will be discussed in the vg paper.
 The proofs of lemmas describing the properties of the generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}.
@@
 \begin{itemize}
-\item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
+\item each node $v_{X} \in V$ represents a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
 \item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
@@
 We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The resulting index takes less than 16~gigabytes, which is less than what the existing path indexes capable of indexing arbitrary graphs require.
+The resulting index takes less than 16~gigabytes, which is less than 1~bit per \kmer{k} for the \orderk{128} index.
 Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
 
-GCSA2 improves upon the existing path indexes in the following ways:
+GCSA2 compares to the existing path indexes in the following ways:
 \begin{itemize}
 
-\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques, but they can only index graphs that have finite \orderk{\infty} path graphs.
-On the other hand, they support queries of arbitrary length.
+\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques as GCSA2.
+While GCSA2 can index any graph but restricts the length of the queries, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.
 
-\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than GCSA2. The current implementation can only handle local variation.
+\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2.
+The current implementation can only handle local variation.
 
@@
 
-\item The hypertext index \cite{Thachuk2013} has not been implemented. It supports queries of arbitrary length, but becomes slow with matches crossing more than one edge.
+\item Probabilistic de~Bruijn graphs \cite{Pell2012} are faster to query, but they support only queries of length $k$, produce false positives, and require multiple bits per \kmer{k}.
+
+\item The hypertext index \cite{Thachuk2013} has not been implemented. It supports queries of arbitrary length but becomes slow with matches crossing more than one edge.
 
@@
 If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
-Now assume that $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ is the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of $G'.\gkey(v')$.
+Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ bet the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
 
-Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i])$, and assume that it is nonempty.
-Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
-Hence if $u' \in M_{i}$, key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$.
+Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1])$.
+There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
+Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$ for all nodes $u' \in M_{i}$.
+
+Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m]$.
+If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1]$ be a substring represented by node $u'$ in the \kcollection{k} used for building the path graph.
+As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+
+Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$.
 If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.

### 2016-07-12 paper/paper.tex
Updates to the paper; more meaningful kmer counts.

@@
 Compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs achieve similar space usage by representing unary paths in the graph as single nodes.
-A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.
 Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to store the graph.
 They support faster queries in similar space at the expense of producing false positives.
+Read aligners based on various encodings of de~Bruijn graphs have been proposed \cite{Limasset2015,Liu2016}.
 
@@
 BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding simple variants and creates new sequences with a sufficient amount of context for complex variants.
-Queries are slower than in ordinary FM\nobreakdash-indexes due to using the powerset alphabet.
+Queries are slower than in ordinary FM\nobreakdash-indexes because of the powerset alphabet.
 vBWT \cite{Maciuca2016} encodes variant sites explicitly in the sequence as  $X(A|B|C)Y$, using distinct separator symbols for each site.
@@
 
-We assume that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$.
-To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$.
-These characters are only used to label technical nodes similar to nodes $s$ and $t$.
-We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
+If the graph has nodes with indegree or outdegree $0$, we add a \emph{source} node $s$ and a \emph{sink} node $t$ to it.
+To distinguish these technical nodes from the actual nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$, which are not used anywhere else in the graph.
+We add an edge $(s, v)$ to all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ from all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
 We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
@@
 Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k > 0$ be a parameter value.
-We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if the collection contains substrings $\#^{k}$ and $\$^{k}$, and each sequence $S \in \mathcal{S}$
+We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if each sequence $S \in \mathcal{S}$
 (a) is left-infinite or begins with $\#^{k}$;
@@
 We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence.
-If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the start node and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
+If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the source node and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
 Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = t:j$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
@@
 Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
-Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that $\#^{k}, \$^{k} \in \mathcal{K}$.
+Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$.
 The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
@@
 
-Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
+Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
 We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
@@
 
-The doubling steps are followed by the \emph{merging step}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph.
-It consists of a pruning step, followed by merging the paths with identical labels.} First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
+The doubling steps are followed by the \emph{merging step}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.
+First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
 If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$.

### 2016-07-12 algorithms.cpp

@@ -238,37 +238,35 @@ struct ReverseTrieNode
 {
   range_type range;
   size_type  depth;
-  bool       ends_at_sink;
 
   const static size_type SEED_LENGTH = 5; // Create all patterns of that length before parallelizing.
 
-  ReverseTrieNode() : range(0, 0), depth(0), ends_at_sink(false) {}
-  ReverseTrieNode(range_type rng, size_type d, bool sink) : range(rng), depth(d), ends_at_sink(sink) {}
+  ReverseTrieNode() : range(0, 0), depth(0) {}
+  ReverseTrieNode(range_type rng, size_type d) : range(rng), depth(d) {}
 };
 
 struct KMerCounter
 {
-  size_type count, limit;
+  size_type count, depth_limit;
 
-  KMerCounter(size_type _limit) : count(0), limit(_limit) {}
+  KMerCounter(size_type limit) : count(0), depth_limit(limit) {}
 
   inline void add(const ReverseTrieNode& node)
   {
-    if(node.ends_at_sink || node.depth >= this->limit) { this->count++; }
+    if(node.depth >= this->depth_limit) { this->count++; }
   }
 };
 
 struct SeedCollector
 {
-  size_type count, limit;
+  size_type depth_limit;
   std::vector<ReverseTrieNode> seeds;
 
-  SeedCollector(size_type _limit) : count(0), limit(_limit), seeds() {}
+  SeedCollector(size_type limit) : depth_limit(limit), seeds() {}
 
   inline void add(const ReverseTrieNode& node)
   {
-    if(node.depth >= this->limit) { this->seeds.push_back(node); }
-    else if(node.ends_at_sink) { this->count++; } // Seeds ending at the sink are counted later.
+    if(node.depth >= this->depth_limit) { this->seeds.push_back(node); }
   }
 };
 
@@ -282,12 +280,12 @@ processSubtree(const GCSA& index, std::stack<ReverseTrieNode>& node_stack, Handl
     ReverseTrieNode curr = node_stack.top(); node_stack.pop();
     if(Range::empty(curr.range)) { continue; }
     handler.add(curr);
-    if(curr.depth < handler.limit)
+    if(curr.depth < handler.depth_limit)
     {
       index.LF(curr.range, predecessors);
       for(size_type comp = 1; comp + 1 < index.alpha.sigma; comp++)
       {
-        node_stack.push(ReverseTrieNode(predecessors[comp], curr.depth + 1, curr.ends_at_sink));
[…]

### 2016-07-15 paper/paper.tex
Updates to the experiments

@@
 % Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}}
+\title{Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant 098051.}}
 
@@
 
-\subsection{Extensions}
+\subsection{Extensions}\label{sect:extensions}
 
@@
 We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
-The results are summarized in Table~\ref{table:construction}.
-We could build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
-The de~Bruijn graphs had roughly $1.031^{k} \cdot 2.38$~billion nodes, but pruning reduced the size to 4.4\nobreakdash--5.8~billion nodes.
-The index grew more slowly with $k$, as we sampled the values more often in lower order graphs.
+
+Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes.
+We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
+The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them.
+Index size grows even more slowly with $k$, as we sample more values in lower order graphs.
+For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
 
@@
 \caption{GCSA2 construction with different orders of the path graph.
-Billions of nodes in the de~Bruijn graph and the pruned de~Bruijn graph; index size in gigabytes; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
-\begin{tabular}{c|ccc|ccc|cc}
+Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
+\begin{tabular}{c|ccc|cc}
 \hline
-\textbf{Order} & \textbf{Nodes} & \textbf{Pruned} & \textbf{Index} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
+\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
 \hline
- 32 & 6.23G & 4.39G & 13.7 GB & 9.85 h & 63.2 GB & 401 GB & 1.55 TB & 1.05 TB \\
- 64 & 16.9G & 5.27G & 14.2 GB & 13.6 h & 56.1 GB & 424 GB & 2.23 TB & 1.71 TB \\
-128 &  118G & 5.76G & 15.3 GB & 18.0 h & 56.9 GB & 489 GB & 3.00 TB & 2.47 TB \\
+ 32 &        &         &        &         &         \\
+ 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
+128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[t!]
+\begin{center}
+\caption{Index sizes with different orders $k$ of the path graph.
+Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Section \ref{sect:extensions}.}\label{table:indexes}
+\begin{tabular}{c|cc|c|c|c}
+\hline
+\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
+\hline
+ 32 & 6.20G & & & & \\
+ 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
+128 &  116G & 5.76G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
 \hline

### 2016-07-18 paper/paper.tex
Updated the experiments.

@@
 As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-The index uses fast succinct data structures instead of the smaller but slower compressed data structures in most of its components.
+The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components.
 The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
 All files were stored on a distributed Lustre file system.
-We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the queries.
+We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.
 
@@
 \hline
- 32 &        &         &        &         &         \\
+ 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
  64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
@@
 \hline
- 32 & 6.20G & & & & \\
+ 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
  64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
@@
 
-We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread.
+We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread.
 We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference.
@@
 \begin{center}
-\caption{Query benchmarks using an \orderk{64} GCSA and an FM\nobreakdash-index.
+\caption{Query benchmarks using an \orderk{128} GCSA and an FM\nobreakdash-index.
 Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
@@
 \hline
-GCSA: 16 & 351584 & 347453 & 1829M &  872M & 18.0 \textmu{}s & 0.40 \textmu{}s & 0.93 \textmu{}s & 8.61 \textmu{}s \\
-GCSA: 32 & 351555 & 333258 & 52.6M & 34.3M & 37.7 \textmu{}s & 0.27 \textmu{}s & 0.44 \textmu{}s & 4.52 \textmu{}s \\
-GCSA: 64 & 351567 & 326101 & 0.33M & 1.35M & 75.3 \textmu{}s & 0.25 \textmu{}s & 0.33 \textmu{}s & 1.66 \textmu{}s \\
+GCSA:  16 & 351584 & 347453 & 2477M &  872M & 18.3 \textmu{}s & 0.41 \textmu{}s & 0.87 \textmu{}s & 12.1 \textmu{}s \\
+GCSA:  32 & 351555 & 333258 &  112M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 11.1 \textmu{}s \\
+GCSA:  64 & 351567 & 326101 & 2.63M & 1.35M & 76.1 \textmu{}s & 0.26 \textmu{}s & 0.28 \textmu{}s & 5.76 \textmu{}s \\
+GCSA: 128 & 351596 & 316500 & 0.32M & 0.37M &  149 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 5.98 \textmu{}s \\
 \hline
-FM:   16 & 351584 & 301538 &    -- &  782M & 6.04 \textmu{}s &              -- &              -- & 2.44 \textmu{}s \\
-FM:   32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.17 \textmu{}s \\
-FM:   64 & 351567 &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.92 \textmu{}s \\
+FM:    16 & 351584 & 301538 &    -- &  782M & 6.06 \textmu{}s &              -- &              -- & 2.46 \textmu{}s \\
+FM:    32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.19 \textmu{}s \\
+FM:    64 & 351567 &  88184 &    -- & 0.84M & 17.2 \textmu{}s &              -- &              -- & 1.91 \textmu{}s \\
+FM:   128 & 351596 &  35678 &    -- & 0.08M & 23.6 \textmu{}s &              -- &              -- & 3.48 \textmu{}s \\
 \hline
@@
 We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The resulting index takes less than 16~gigabytes, which is less than 1~bit per \kmer{k} for the \orderk{128} index.
+The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
 Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
@@
 
-\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques as GCSA2.
-While GCSA2 can index any graph but restricts the length of the queries, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.
+\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} are based on the same techniques.
+While GCSA2 can index any graph but restricts query length, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.
 
-\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2.
+\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2 would be.
 The current implementation can only handle local variation.

### 2016-07-22 paper/paper.tex
Updates to introduction and conclusions

@@
 Given a query string, it finds the paths with a label matching the query.
-Indexing graphs for path queries is inherently hard, as the number of paths increases exponentially with path length.
-The design of a path index is hence a trade-off between maximum query length, the size of the index, and pruning complex regions of the graph.
+Indexing graphs is inherently hard, as the number of paths increases exponentially with path length.
+The design of a path index is hence a trade-off between maximum query length, index size and performance, and pruning complex regions of the graph.
 
@@
 
-The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} is generalization of the FM\nobreakdash-index for representing \emph{de~Bruijn graphs} space-efficiently.
+\emph{De~Bruijn graphs} can be used as \kmer{k} indexes of other graphs.
+There are already some read aligners based on them \cite{Limasset2015,Liu2016}.
+The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph space-efficiently with generalization of the FM\nobreakdash-index.
 Rødland \cite{Roedland2013} proposed another similar generalization.
-Succinct de~Bruijn graphs can represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
-Compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs achieve similar space usage by representing unary paths in the graph as single nodes.
-Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to store the graph.
-They support faster queries in similar space at the expense of producing false positives.
-Read aligners based on various encodings of de~Bruijn graphs have been proposed \cite{Limasset2015,Liu2016}.
+Succinct de~Bruijn graphs can simulate \orderk{k} de~Bruijn graphs for multiple values of $k$ \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
+Other space-efficient representations include compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs, which represent unary paths in the graph as single nodes or edges.
+Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to support faster queries, at the expense of producing false positives.
+All these space-efficient representations require several bits per \kmer{k} for the graph itself, and more for mapping back to the indexed graph.
 
 Some path indexes store the graph as a collection of sequences.
-BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding simple variants and creates new sequences with a sufficient amount of context for complex variants.
-Queries are slower than in ordinary FM\nobreakdash-indexes because of the powerset alphabet.
+BWBBLE \cite{Huang2013} uses the powerset alphabet for encoding substitutions and creates new sequences with a sufficient amount of context for other variants.
 vBWT \cite{Maciuca2016} encodes variant sites explicitly in the sequence as $X(A|B|C)Y$, using distinct separator symbols for each site.
-As the variants must be taken into account during search, queries are slower than in ordinary FM\nobreakdash-indexes.
-The \emph{hypertext index} \cite{Thachuk2013} assumes that the nodes of the graph are labeled with strings.
-The strings are indexed using an ordinary FM-index.
+Queries in both BWBBLE and vBWT are slower than in ordinary FM\nobreakdash-indexes, as variant sites force the search to branch.
+The \emph{hypertext index} \cite{Thachuk2013} works with graphs that have string labels in nodes.
+The strings are indexed using an FM-index.
 Partial matches in the strings are combined into full matches with range queries in the edge matrix.
@@
 The above methods can be classified in three categories based on the data models they use.
-Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation.
+Many deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation.
 GCSA-based methods can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}.
@@
 The pruned graph is encoded with a generalization of the FM\nobreakdash-index.
-GCSA2 often uses less than 1~bit per \kmer{k} with large values of $k$.
-Other de~Bruijn graph-based path indexes cannot achieve similar space usage, as they store some information for each \kmer{k} explicitly.
+GCSA2 often uses around 1~bit per \kmer{k} with large values of $k$.
+Explicit de~Bruijn graphs cannot achieve similar space usage, as they store some information for each \kmer{k}.
 GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
@@
 
-We built vg graphs from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}.
+We built vg graphs from the human reference genome (GRCh37) and 1000~Genomes Project variation \cite{1000GP2015}.
 To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
@@
 There were a total of 4.80~billion paths with 1.53~billion distinct labels.
-We built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
+We then built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
 
@@
 For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
+Extensions from Section~\ref{sect:extensions} increase index size to 1.12~bits per \kmer{k}.
 
@@
 GCSA2 is a path index for variation graphs.
-It uses a pruned de~Bruijn graph as an index of the variation graph, and encodes that with a generalization of the FM\nobreakdash-index.
-The index includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
-GCSA2 has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
+It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
+The index supports queries of length up to $k$ exactly and longer queries with false positives.
+GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
+The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
 
@@
 
-I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index, and for the new ideas I got from the discussions.
+The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
 

### 2016-08-10 paper/paper.tex
Updated the abstract

@@
 They generalize text indexes for graphs, allowing one to find the paths matching the query string.
-We propose using pruned de Bruijn graphs as path indexes, and encode them space-efficiently with the Burrows-Wheeler transform.
-We also generalize ideas from text indexing literature and use them with graphs.
-The proposed approach has been implemented in the variation graph toolkit vg.
+We propose using de Bruijn graphs as path indexes, compressing them by merging redundant subgraphs, and encoding them with the Burrows-Wheeler transform.
+The resulting fast, space-efficient, and versatile index is used in the variation graph toolkit vg.
 \end{abstract}

### 2016-08-10 paper/paper.tex
Fixed some numbers

--- 2016-07-22 paper/paper.tex
@@
  64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
-128 &  116G & 5.76G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
+128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
 \hline

### 2016-08-10 paper/paper.tex
Started preparing a new version of the paper

@@
-\documentclass[a4paper,UKenglish]{lipics-v2016}
+\documentclass[a4paper,11pt]{llncs}
 
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+
+\usepackage{fullpage}
+\setlength\tabcolsep{4pt}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{textgreek}
 \usepackage{microtype}
-
-\bibliographystyle{plainurl}
-
-% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant 098051.}}
-
-\author[1]{Jouni Sirén}
-\affil[1]{Wellcome Trust Sanger Institute, Hinxton, Cambridge, UK\\
-  \texttt{jouni.siren@iki.fi}}
-\authorrunning{J. Sirén}
-
-\Copyright{Jouni Sirén}
-
-\subjclass{E.1 Data Structures}
-\keywords{Burrows-Wheeler transform, de Bruijn graphs, path indexes, variation graphs}
-% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\EventEditors{John Q. Open and Joan R. Acces}
-%\EventNoEds{2}
-%\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-%\EventShortTitle{CVIT 2016}
-%\EventAcronym{CVIT}
-%\EventYear{2016}
-%\EventDate{December 24--27, 2016}
-%\EventLocation{Little Whinging, United Kingdom}
-%\EventLogo{}
-%\SeriesVolume{42}
-\ArticleNo{1}
-% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{hyperref}
 
@@
 % DNA
-\newcommand{\dnaseq}[1]{\ensuremath{\mathtt{#1}}}
-\newcommand{\baseA}{\dnaseq{A}}
-\newcommand{\baseC}{\dnaseq{C}}
-\newcommand{\baseG}{\dnaseq{G}}
-\newcommand{\baseT}{\dnaseq{T}}
-\newcommand{\baseN}{\dnaseq{N}}
+\newcommand{\baseA}{\mathtt{A}}
+\newcommand{\baseC}{\mathtt{C}}
+\newcommand{\baseG}{\mathtt{G}}
+\newcommand{\baseT}{\mathtt{T}}
+\newcommand{\baseN}{\mathtt{N}}
 \newcommand{\dnacomp}[1]{\ensuremath{\overline{#1}}}
@@
 
+\title{Indexing Variation Graphs\thanks{This work was supported by the Wellcome Trust grant 098051.}}
+
+\author{Jouni Sirén}
+\institute{Wellcome Trust Sanger Institute, UK \\ \email{jouni.siren@iki.fi}}
+
 \begin{document}
 
+\thispagestyle{empty}
+\setcounter{page}{0}
 \maketitle
@@
 
+\clearpage
 \section{Introduction}
@@
 Given a DNA sequence $S$, its \emph{reverse complement} is the sequence $\revcomp{S}$ obtained by reversing the non-technical parts of the sequence and replacing each character with its complement.
-For example, $\revcomp{\dnaseq{GATTACA}\$} = \dnaseq{TGTAATC}\$$.
+For example, $\revcomp{\mathtt{GATTACA}\$} = \mathtt{TGTAATC}\$$.
 
@@
 \includegraphics{gcsa2_text_indexes.pdf}
-\caption{\LFmapping, BWT, suffix array, and LCP array for text $\dnaseq{GCATCATA}\$$.}\label{figure:indexes}
+\caption{\protect\LFmapping, BWT, suffix array, and LCP array for text $\mathtt{GCATCATA}\$$.}\label{figure:indexes}
 \end{figure}
@@
 \caption{Left: Input graph $G = (V, E)$, with each node $v \in V$ labeled with $v:G.\glabel(v)$.
-Right: The \orderk{3} de~Bruijn graph $G' = (V', E')$ of graph $G$, with each node $v' \in V'$ labeled with $G'.\gkey(v')$ and $G'.\gvalue(v')$.
+Right: The \protect\orderk{3} de~Bruijn graph $G' = (V', E')$ of graph $G$, with each node $v' \in V'$ labeled with $G'.\gkey(v')$ and $G'.\gvalue(v')$.
 Both: Edges $(t, s)$ are not shown.
@@
 \includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
-\caption{Left: An \orderk{3} pruned de~Bruijn graph $G''$ \kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}.
+\caption{Left: An \protect\orderk{3} pruned de~Bruijn graph $G''$ \protect\kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}.
 Right: GCSA for graph $G''$.
-Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$.
-Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
+Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$.
+Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
 \end{figure}
@@
 
-\subparagraph*{Acknowledgements.}
+\paragraph*{Acknowledgements.}
 
@@
 
+\bibliographystyle{plainurl}
 \bibliography{paper}

### 2016-08-12 paper/paper.tex
Minor updates to the paper.
The paper is now at 10 pages

@@
 \newcommand{\Oh}[1]{\ensuremath{\mathsf{O}\!\left( #1 \right)}}
+\newcommand{\oh}[1]{\ensuremath{\mathsf{o}\!\left( #1 \right)}}
 
@@
 A \emph{path index} is a generalization of text indexes for labeled graphs.
-Given a query string, it finds the paths with a label matching the query.
+Given a query string, the index finds the paths with a label matching the query.
 Indexing graphs is inherently hard, as the number of paths increases exponentially with path length.
-The design of a path index is hence a trade-off between maximum query length, index size and performance, and pruning complex regions of the graph.
+The design of a path index is hence a trade-off between maximum query length, index size, query performance, and pruning complex regions of the graph.
 
 Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}.
-Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs.
-It depends on pruning the complex regions, as there are no limits on query length.
+Sirén et~al.
+developed \emph{GCSA}~\cite{Siren2014}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs.
+GCSA depends on pruning the complex regions, as there are no limits on query length.
 Kim et~al.~\cite{Kim2015-2016} combined GCSA with the HISAT read aligner \cite{Kim2015}.
@@
 There are already some read aligners based on them \cite{Limasset2015,Liu2016}.
-The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph space-efficiently with generalization of the FM\nobreakdash-index.
+The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph space-efficiently with a generalization of the FM\nobreakdash-index.
 Rødland \cite{Roedland2013} proposed another similar generalization.
-Succinct de~Bruijn graphs can simulate \orderk{k} de~Bruijn graphs for multiple values of $k$ \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly.
+Succinct de~Bruijn graphs can simulate \orderk{k} de~Bruijn graphs for multiple values of $k$ \cite{Boucher2014}, but they still need to store the graph explicitly for the largest value of $k$.
 Other space-efficient representations include compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs, which represent unary paths in the graph as single nodes or edges.
@@
 Queries in both BWBBLE and vBWT are slower than in ordinary FM\nobreakdash-indexes, as variant sites force the search to branch.
-The \emph{hypertext index} \cite{Thachuk2013} works with graphs that have string labels in nodes.
+The hypertext index \cite{Thachuk2013} works with graphs that have string labels on the nodes.
 The strings are indexed using an FM-index.
@@
 GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs.
-Conceptually it uses a de~Bruijn graph as a path index of a variation graph.
-The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely.
+Conceptually it uses a de~Bruijn graph as a \kmer{k} index of a variation graph.
+The de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely.
 The pruned graph is encoded with a generalization of the FM\nobreakdash-index.
@@
 
-This paper describes the GCSA2 data structure.
-Its uses in the vg toolkit will be discussed in the vg paper.
-The proofs of lemmas describing the properties of the generalizations of de~Bruijn graphs are included in Appendix~\ref{appendix:proofs}.
+This paper describes the GCSA2 data structure. Its uses in the vg toolkit will be discussed in the vg paper.
 
@@
 A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$.
-We often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text.
-\emph{Binary} sequences are strings over the alphabet $\set{0, 1}$.
+\emph{Text} strings $T[0, n-1]$ are terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text.
 A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$.
@@
 Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
-The notion of substrings generalizes for such infinite sequences in a natural way.
-We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
+The notion of substring generalizes in a natural way for infinite sequences.
+A substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
 A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix, and \emph{right-maximal} if it is right-infinite or a suffix.
 
-We primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$.
+We are primarily interested in sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$.
 Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base.
-In addition to the endmarker $\$$, the alphabet may also contain other characters for technical purposes.
+The alphabet may contain other technical characters in addition to the enmarker $\$$.
 Each character $c$ of the DNA alphabet has a \emph{complement} $\dnacomp{c}$ defined as $\dnacomp{\baseA} = \baseT$, $\dnacomp{\baseC} = \baseG$, $\dnacomp{\baseG} = \baseC$, $\dnacomp{\baseT} = \baseA$, and $\dnacomp{c} = c$ for other characters $c$.
@@
 \begin{figure}[t!]
+\centering
 \includegraphics{gcsa2_text_indexes.pdf}
@@
 
-The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} is a permutation of the text.
-Given text $T[0, n-1]$, its BWT is a string $\BWT[0, n-1]$, where $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
+The \emph{Burrows-Wheeler transform (BWT)} \cite{Burrows1994} of text $T[0, n-1]$ is a permutation $\BWT[0, n-1]$ such that $\BWT[i] = T[(\SA[i]-1) \bmod n]$.
 Given the \emph{lexicographic rank} $i$ of suffix $T[\SA[i], n-1]$, we can use \emph{\LFmapping} on the BWT to find the lexicographic rank of the previous suffix $T[(\SA[i]-1) \bmod n, n-1]$.
@@
 If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, the range matching suffix $X[i, \abs{X}-1]$ is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$.
-Finding the lexicographic range matching the entire pattern requires $\Oh{\abs{X}}$ $\rank$ queries.
+Matching the entire pattern takes $\Oh{\abs{X}}$ $\rank$ queries.
 
 We use the sampled suffix array pointers for finding the text positions containing the occurrences.
-If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer.
+If $\SA[i]$ is not sampled, we iterate $\LF(i)$ until we find a sampled pointer.
 If we find a sample at $\SA[\LF^{k}(i)]$, we know that
@@
 The \emph{longest-common-prefix array} (LCP array) \cite{Manber1993} is an integer array $\LCP[0, n-1]$, where each value $\LCP[i]$ tells the length of the longest common prefix of suffixes $T[\SA[i-1], n-1]$ and $T[\SA[i], n-1]$ (with $\LCP[0] = 0$).
-It is often used to augment the functionality of the suffix array and the \FMindex.
-In particular, if we augment the \FMindex{} with the LCP array and the topology of the suffix tree, we get the \emph{compressed suffix tree}, which supports the full functionality of the suffix tree in a space-efficient manner \cite{Sadakane2007}.
+It is used to augment the functionality of the suffix array and the \FMindex.
+In particular, if we have the \FMindex{}, the LCP array, and the topology of the suffix tree, we get the \emph{compressed suffix tree}, which supports the full functionality of the suffix tree in a space-efficient manner \cite{Sadakane2007}.
 
@@
 
-\subsection{Generalized indexes}
+\subsection{FM-index for graphs}
 
@@
 The \emph{XBW transform} \cite{Ferragina2009b} is essentially an \FMindex{} for \emph{labeled trees}.
-The nodes of the tree are sorted by path labels from the node to the root of the tree.
+The nodes of the tree are sorted by path labels from the node to the root.
 $\BWT$ stores the labels of the children of each node (leaf nodes require special treatment).
 If a node has $k$ children, we encode that as a binary sequence $0^{k-1} 1$.
-We concatenate these sequences to form bitvector $B$, which is used for mapping lexicographic ranks of nodes to the corresponding BWT ranges.
-The labels of the children of the $i$th node are found in $\BWT[B.\select(i, 1) + 1, B.\select(i + 1, 1)]$.
+We concatenate these sequences to form the indegree bitvector $\bvIN$.
+The labels of the children of the $i$th node are found in $\BWT[\bvIN.\select(i, 1) + 1, \bvIN.\select(i + 1, 1)]$.
 
-The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
-Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them.
+The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} extends the XBW transform to a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
+Before indexing, we transform the graph into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them.
 The transformation increases the size of the graph exponentially in the worst case.
-After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences.
-$\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform.
+In addition to sequences $\BWT$ and $\bvIN$, GCSA also uses the outdegree bitvector $\bvOUT$, which is encoded in the same way as $\bvIN$.
 \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
@@
 
-The design of a path index is a trade-off between maximum query length and the size of the index.
-In the worst case, a graph may have up to $\sigma^{k}$ distinct labels of paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and cyclic graphs have paths of unlimited length.
-GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary.
-The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.
+The proofs of the lemmas in this section can be found in Appendix~\ref{appendix:proofs}.
 
-\subsection{Basic indexes}
+\subsection{De Bruijn graphs as path indexes}
 
 The \emph{\kmer{k} index} is the simplest path index.
-In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ and $V_{X} \subseteq V$.
-Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $V_{X}$ as matches if we find it.
-Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.
+It consists of a set of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ is a \kmer{k} and $V_{X} \subseteq V$ is the set of start nodes of the paths matching the \kmer{k}.
+If we store the pairs in a hash table, we can quickly search for patterns of length $k$.
+If we store the pairs as a sorted list, queries become slower, as they require binary search.
+On the other hand, we gain the ability to search for patterns shorter than $k$ characters.
+The main drawback of these basic \kmer{k} indexes is their size, as they have to store the key-value pairs explicitly.
 
-Another representation trades query performance for the ability to search for short patterns.
-We sort the key-value pairs by their keys and store them in array $A$.
-If $X \in \Sigma^{\ast}$ is a pattern with $\abs{X} \le k$, we find the range $A[sp, ep]$ of pairs $(X_{i}, V_{i})$, where $X$ is a prefix of key $X_{i}$, with binary search.
-We then list the union $\bigcup_{i=sp}^{ep} V_{i}$ as the set of matches.
-
-We can use de~Bruijn graphs and \kmer{k} indexes to simulate each other.
+We can represent \kmer{k} indexes as de~Bruijn graphs.
 For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph.
@@
 If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s \in V$ is the source node and $j$ is the distance to the nearest non-inserted $\#$ in $S$.
-Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = t:j$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.
 
@@
 
-A \kmer{(k+1)} index of graph $G = (V, E)$ can simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$.
-We represent the nodes of the de~Bruijn graph with the adjacent edges.
-If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$.
-Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.
-
-We can also use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ as a \kmer{k} index of $G$.
-Let $\mathcal{S}$ be the \kcollection{k} in Definition~\ref{def:dbg}.
+Let $G' = (V', E')$ be a de~Bruijn graph of graph $G = (V, E)$, and let $\mathcal{S}$ be the \kcollection{k} used to define it.
 We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ matching pattern $G'.\gkey(v)$ in graph $G$.
-There will be no false negatives, regardless of pattern length.
-There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results of such queries in the input graph.
+There will be no \emph{false negatives} (path labels that exist in the input graph but not in the index), regardless of pattern length.
+There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results in the input graph.
 See Figure~\ref{figure:graph-dbg} for an example.
@@
 \begin{figure}[t!]
-\includegraphics[width=396pt]{gcsa2_graph_dbg.pdf}
+\includegraphics[width=\textwidth]{gcsa2_graph_dbg.pdf}
 \caption{Left: Input graph $G = (V, E)$, with each node $v \in V$ labeled with $v:G.\glabel(v)$.
@@
 
+We can encode the de~Bruijn graph as a GCSA using $(\log \sigma + 2)(1 + \oh{1})$ bits per \kmer{k}.
+By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that typically uses a couple of bytes per \kmer{k} (see the \orderk{32} index in Section~\ref{sect:experiments}).
+While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.
+
 \subsection{Path graphs}
 
-Path graphs generalize de~Bruijn graphs by allowing node keys of different lengths.
+When a path of length $k' < k$ has a unique label, its extensions become \emph{redundant} nodes in an \orderk{k} de~Bruijn graph, if the path branches after the first $k'$ characters.
+By using shorter keys when possible, we can build a smaller graph equivalent to the de~Bruijn graph as a path index.
 
@@
 \item If $V'_{X} \subseteq V'$ is the set of nodes matching pattern $X$, the set of \emph{occurrences} for the pattern is $G'.\locate(V'_{X}) = \bigcup_{v' \in V'_{X}} G'.\gvalue(v')$.
-We use $G'.\locate(X)$ as a shorthand for $G'.\locate(G'.\find(X))$.
+We use $G'.\locate(X)$ to denote $G'.\locate(G'.\find(X))$.
 \end{itemize}
@@
 
-When used as an index, a path graph will not produce false negatives.
-On the other hand, we may encounter false positives already at path length $k'+1$, where $k'$ is the length of the shortest key in the key set.
-This is not very convenient for a path index.
-In the next section, we consider a class of path graphs that can be proven to be equivalent to de~Bruijn graphs.
+A path graph may produce false positives with patterns longer than $k'$ characters, where $k'$ is the length of the shortest key.
+In the next section, we define a class of path graphs that can be proven to be equivalent to de~Bruijn graphs.
 
@@
 
-De~Bruijn graphs often have nodes that are redundant when the graph is used as a path index.
-By carefully pruning the graphs, we get smaller graphs that are equivalent to \orderk{k} de~Bruijn graphs with patterns of length $k$ or less.
-These pruned de~Bruijn graphs naturally arise from GCSA construction.
-They are similar to manifold de Bruijn graphs \cite{Lin2014}.
+We can prune de~Bruijn graphs by merging keys sharing a common prefix, if the values are identical.
+These pruned de~Bruijn graphs, which arise naturally from GCSA construction, are similar to manifold de Bruijn graphs \cite{Lin2014}.
+As path indexes, they are equivalent to de~Bruijn graphs with patterns of length up to $k$ characters.
 
@@
 \begin{figure}[t!]
-\includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
+\includegraphics[width=\textwidth]{gcsa2_pruned_index.pdf}
 \caption{Left: An \protect\orderk{3} pruned de~Bruijn graph $G''$ \protect\kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}.
@@
 
-We now show how a GCSA of a path graph works.
-As path graphs are not exactly the same as the prefix-sorted automata used with the original GCSA \cite{Siren2014}, the technical details will be different.
-See Figure~\ref{figure:pruned-index} for an example of a GCSA index.
-We also extend GCSA with techniques based on suffix trees.
-Appendix~\ref{appendix:construction} gives a construction algorithm for the GCSA of a pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph, starting from a set of paths of length $k$.
-While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.
+GCSA2 construction starts from paths of length $k$ in the input graph.
+We build a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph using a similar \emph{prefix-doubling} algorithm as in the original GCSA \cite{Siren2014}, and encode the result as a GCSA.
+To avoid excessive memory usage, we keep the paths and the graphs on disk, and read only a single chromosome at a time into memory.
+The details of the construction algorithm can be found in Appendix~\ref{appendix:construction}.
 
-\subsection{GCSA for path graphs}
+As in the original GCSA, we sort the nodes of the path graph in lexicographic order, encode the indegrees and outdegrees in bitvectors $\bvIN$ and $\bvOUT$, and store the predecessor labels in $\BWT$.
+See Figure~\ref{figure:pruned-index} for an example.
+If lexicographic range $[sp_{i+1}, ep_{i+1}]$ matches suffix $X[i+1, \abs{X}-1]$ of pattern $X$, we can find the range matching suffix $X[i, \abs{X}-1]$ as
+\begin{align*}
+[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
+[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
+[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+\end{align*}
+In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the values of the predecessor.
+We may also sample other values to improve query performance.
+The sampled nodes are marked in bitvector $B_{S}$, the number of values in each sample is encoded in unary in bitvector $B_{V}$, and the sampled values are stored in integer array $V_{S}$.
+A detailed description of the data structure can be found in Appendix~\ref{appendix:gcsa2}.
+
+The compacted trie of keys resembles a suffix tree.
+We can simulate it space-efficiently by using the LCP array \cite{Abouelhoda2004,Fischer2009a}, and thus extend GCSA2 to support many suffix tree operations.
+For example, we can search for \emph{maximal exact matches} by using \LFmapping{} and $\parent$ queries \cite{Ohlebusch2010a}, and use that as a basis for a read aligner similar to BWA-MEM \cite{Li2013}.
+We can also use \emph{document counting} techniques \cite{Sadakane2007a} to quickly count the number of distinct matches in a lexicographic range.
+Further details of these extensions can be found in Appendix~\ref{appendix:extensions}.
+
+\section{Implementation and Experiments}\label{sect:experiments}
+
+GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}.
+The implementation is written in C++, and the source code is available on GitHub (\href{https://github.com/jltsiren/gcsa2}{https://github.com/jltsiren/gcsa2}).
+As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components.
+The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
+All files were stored on a distributed Lustre file system.
+We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.
+
+The vg \emph{variation graphs} have strings as node labels.
+Each node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
+When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
+Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples.
+
+We built vg graphs from the human reference genome (GRCh37) and 1000~Genomes Project variation \cite{1000GP2015}.
+To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
+We extracted all paths of length $16$ from the forward strands of the graphs.
+There were a total of 4.80~billion paths with 1.53~billion distinct labels.
+We then built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
+
+Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes.
+We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
+The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them.
+Index size grows even more slowly with $k$, as we sample more values in lower order graphs.
+For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
+Extensions from Appendix~\ref{appendix:extensions} increase index size to 1.12~bits per \kmer{k}.
+
+\begin{table}[t!]
+\begin{center}
+\caption{GCSA2 construction with different orders of the path graph.
+Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
+\begin{tabular}{c|ccc|cc}
+\hline
+\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
+\hline
+ 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
+ 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
+128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[t!]
+\begin{center}
+\caption{Index sizes with different orders $k$ of the path graph.
+Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
+\begin{tabular}{c|cc|c|c|c}
+\hline
+\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
+\hline
+ 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
+ 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
+128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread.
+We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference.
+The results can be seen in Table~\ref{table:benchmark}.
+The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components.
+A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
+
+\begin{table}[t!]
+\begin{center}
+\caption{Query benchmarks using an \orderk{128} GCSA and an FM\nobreakdash-index.
+Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
+\begin{tabular}{c|cccc|cccc}
+\hline
+ & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+\hline
+GCSA:  16 & 351584 & 347453 & 2477M &  872M & 18.3 \textmu{}s & 0.41 \textmu{}s & 0.87 \textmu{}s & 12.1 \textmu{}s \\
+GCSA:  32 & 351555 & 333258 &  112M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 11.1 \textmu{}s \\
+GCSA:  64 & 351567 & 326101 & 2.63M & 1.35M & 76.1 \textmu{}s & 0.26 \textmu{}s & 0.28 \textmu{}s & 5.76 \textmu{}s \\
+GCSA: 128 & 351596 & 316500 & 0.32M & 0.37M &  149 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 5.98 \textmu{}s \\
+\hline
+FM:    16 & 351584 & 301538 &    -- &  782M & 6.06 \textmu{}s &              -- &              -- & 2.46 \textmu{}s \\
+FM:    32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.19 \textmu{}s \\
+FM:    64 & 351567 &  88184 &    -- & 0.84M & 17.2 \textmu{}s &              -- &              -- & 1.91 \textmu{}s \\
+FM:   128 & 351596 &  35678 &    -- & 0.08M & 23.6 \textmu{}s &              -- &              -- & 3.48 \textmu{}s \\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\section{Conclusions}
+
+GCSA2 is a path index for variation graphs.
+It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
+The index supports queries of length up to $k$ exactly and longer queries with false positives.
+GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
+The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
+
+We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
+The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
+Queries are several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
+
+The primary desing goals in GCSA2 have been query performance and index size.
+The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions.
+Other path indexes work with a more restricted class of graphs \cite{Siren2014,Huang2013,Kim2015-2016,Maciuca2016}, are at least an order of magnitude slower \cite{Huang2013,Maciuca2016}, require much more space \cite{Bowe2012,Pell2012,Cazaux2014,Marcus2014}, or are theoretical proposals that have never been implemented \cite{Thachuk2013}.
+
+The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice.
+As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently.
+We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.
+
+\paragraph*{Acknowledgements.}
+
+The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
+
+\bibliographystyle{plainurl}
+\bibliography{paper}
+
+\clearpage
+\appendix
+\section{Proofs of Lemmas}\label{appendix:proofs}
+
+\begin{proof}[Lemma~\ref{lemma:pg-fn}: No false negatives]
+Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
+The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
+
+For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
+By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
+Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
+As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Lemma~\ref{lemma:pg-context}: Context length]
+If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
+Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ bet the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
+
+Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1])$.
+There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
+Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$ for all nodes $u' \in M_{i}$.
+
+Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m]$.
+If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1]$ be a substring represented by node $u'$ in the \kcollection{k} used for building the path graph.
+As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+
+Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$.
+If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
+Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
+\end{proof}
+
+\begin{proof}[Lemma~\ref{lemma:pg-keys}: Short keys]
+(a) If node $v'$ has multiple predecessors with label $c$, the keys of the predecessors must be longer than string $c \cdot G'.\gkey(v')$, as the key set is prefix-free.
+
+(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
+\end{proof}
+
+\begin{proof}[Lemma~\ref{lemma:dbg-fp}: No short false positives]
+We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
+
+Let $v' \in G'.\find(X)$ be a node.
+By Lemma~\ref{lemma:pg-keys}, pattern $X$ prefix-matches key $G'.\gkey(v')$.
+For every node $v \in G'.\gvalue(v')$, there is a substring $S[i, i+k-1] = G'.\gkey(v'_{0})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$, with $\mathcal{S}.\gnode(S, i) = v$.
+Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
+\end{proof}
+
+\begin{proof}[Lemma~\ref{lemma:dbg-prune}: Pruning]
+Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+
+Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
+By Lemma~\ref{lemma:pg-keys}, $X$ is a prefix of the substring.
+We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
+The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
+If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
+
+Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
+If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
+Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
+As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$.
+Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.
+
+Let $Y$ be the longer of strings $G''.\gkey(v''_{j})$ and $X[j] \cdot G''.\gkey(v''_{j+1})$.
+As $Y$ is a substring of $\mathcal{S}$, there is a node $v_{j} \in V_{d}$ such that key $G_{d}.\gkey(v_{j})$ prefix-matches $Y$.
+Key $G_{d}.\gkey(v_{j})$ has both $X[j, \abs{X}-1]$ and $G''.\gkey(v''_{j})$ as its prefixes, and hence $v_{j} \in V_{d}(v''_{j})$.
+When $j = 0$, we can extend key $G_{d}.\gkey(v_{0})$ to get a path starting from $v_{0}$ and matching $X$ in graph $G_{d}$.
+
+Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
+We can also start from path $P''$ and determine the node $v_{0}$.
+As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
+\end{proof}
+
+\begin{proof}[Lemma~\ref{lemma:dbg-maximal}: Maximal pruning]
+Let $G_{d} = (V_{d}, E_{d})$ the de~Bruijn graph of graph $G$ with the same order $k$ as graph $G'$, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+If $v' \in V'$ is a node, then $G_{d}.\gvalue(v) = G'.\gvalue(v')$ for all nodes $v \in V_{d}(v')$.
+
+Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$.
+Then string $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$.
+There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
+
+Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}.
+Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
+As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.
+
+The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$.
+Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
+\end{proof}
+
+\section{GCSA for Path Graphs}\label{appendix:gcsa2}
 
@@
 The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.
-
-\subsection{Extensions}\label{sect:extensions}
-
-Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
-We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
-The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
-The compacted trie can be considered the \emph{suffix tree} of graph $G'$.
-
-The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
-Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
-These ranges can be determined from the LCP array.
-If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
-
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
-By selecting $x$ according to cache line size, the queries require $\Oh{\log_{x} \abs{V'}}$ random memory accesses, which in practice dominate the computation time.
-Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
-By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
-Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}
-
-BWA-MEM \cite{Li2013} aligns reads to a reference genome by finding \emph{maximal exact matches} between the read and the reference.
-It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions.
-GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$).
-However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
-
-In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
-In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes.
-As $\locate$ queries can be slow, we need another way to support efficient \emph{counting queries}.
-The problem is similar to determining the \emph{document frequency} of a pattern.
-Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
-We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
-
-Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
-If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, the number of \emph{redundant} documents in them is $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
-We create an array $R[0, n-2]$ based on the \emph{inorder traversal} of the suffix tree.
-If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
-We set $R[j] = 0$ for any subsequent visits to the same node.
-Range $R[sp, ep-1]$ covers the internal nodes in the subtree with node $v$ as the root.
-We can determine document frequencies as $\countq(v) = (ep+1-sp) - \sum R[sp, ep-1]$.
-If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute the sums with $\select$ queries:
-$$
-\sum R[a, b] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
-$$
-
-For value counting, we use array $R$ to store the number of redundant values as above.
-We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above.
-The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
-$$
-G'.\countq(X) = G'.\countq([sp, ep]) = (ep+1-sp) + \sum A[sp, ep] - \sum R[sp, ep-1].
-$$
-The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
-
-\section{Implementation and Experiments}
-
-GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}.
-The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
-As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components.
-The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
-All files were stored on a distributed Lustre file system.
-We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.
-
-The vg \emph{variation graphs} have strings as node labels.
-Each node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
-When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
-Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples.
-
-We built vg graphs from the human reference genome (GRCh37) and 1000~Genomes Project variation \cite{1000GP2015}.
-To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
-We extracted all paths of length $16$ from the forward strands of the graphs.
-There were a total of 4.80~billion paths with 1.53~billion distinct labels.
-We then built GCSA with 1\nobreakdash--3 doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes.
-
-Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes.
-We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
-The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them.
-Index size grows even more slowly with $k$, as we sample more values in lower order graphs.
-For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
-Extensions from Section~\ref{sect:extensions} increase index size to 1.12~bits per \kmer{k}.
-
-\begin{table}[t!]
-\begin{center}
-\caption{GCSA2 construction with different orders of the path graph.
-Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
-\begin{tabular}{c|ccc|cc}
-\hline
-\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
-\hline
- 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
- 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
-128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-\begin{table}[t!]
-\begin{center}
-\caption{Index sizes with different orders $k$ of the path graph.
-Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Section \ref{sect:extensions}.}\label{table:indexes}
-\begin{tabular}{c|cc|c|c|c}
-\hline
-\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
-\hline
- 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
- 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
-128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread.
-We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference.
-The results can be seen in Table~\ref{table:benchmark}.
-The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components.
-A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
-
-\begin{table}[t!]
-\begin{center}
-\caption{Query benchmarks using an \orderk{128} GCSA and an FM\nobreakdash-index.
-Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
-\begin{tabular}{c|cccc|cccc}
-\hline
- & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
-\hline
-GCSA:  16 & 351584 & 347453 & 2477M &  872M & 18.3 \textmu{}s & 0.41 \textmu{}s & 0.87 \textmu{}s & 12.1 \textmu{}s \\
-GCSA:  32 & 351555 & 333258 &  112M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 11.1 \textmu{}s \\
-GCSA:  64 & 351567 & 326101 & 2.63M & 1.35M & 76.1 \textmu{}s & 0.26 \textmu{}s & 0.28 \textmu{}s & 5.76 \textmu{}s \\
-GCSA: 128 & 351596 & 316500 & 0.32M & 0.37M &  149 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 5.98 \textmu{}s \\
-\hline
-FM:    16 & 351584 & 301538 &    -- &  782M & 6.06 \textmu{}s &              -- &              -- & 2.46 \textmu{}s \\
-FM:    32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.19 \textmu{}s \\
-FM:    64 & 351567 &  88184 &    -- & 0.84M & 17.2 \textmu{}s &              -- &              -- & 1.91 \textmu{}s \\
-FM:   128 & 351596 &  35678 &    -- & 0.08M & 23.6 \textmu{}s &              -- &              -- & 3.48 \textmu{}s \\
-\hline
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{Conclusions}
-
-GCSA2 is a path index for variation graphs.
-It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
-The index supports queries of length up to $k$ exactly and longer queries with false positives.
-GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
-The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
-
-We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
-Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
-
-GCSA2 compares to the existing path indexes in the following ways:
-\begin{itemize}
-
-\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} are based on the same techniques.
-While GCSA2 can index any graph but restricts query length, GCSA and HISAT2 support queries of any length but require that the graph has an \orderk{\infty} pruned de~Bruijn graph of reasonable size.
-
-\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than a similar implementation of GCSA2 would be.
-The current implementation can only handle local variation.
-
-\item vBWT \cite{Maciuca2016} is slower than BWBBLE. It can only handle local variation but supports queries of arbitrary length.
-
-\item Succinct \cite{Bowe2012}, compacted \cite{Cazaux2014}, and compressed \cite{Marcus2014} de~Bruijn graphs must store at least one character for each \kmer{k} and hence require more space than GCSA2.
-
-\item Probabilistic de~Bruijn graphs \cite{Pell2012} are faster to query, but they support only queries of length $k$, produce false positives, and require multiple bits per \kmer{k}.
-
-\item The hypertext index \cite{Thachuk2013} has not been implemented. It supports queries of arbitrary length but becomes slow with matches crossing more than one edge.
-
-\end{itemize}
-
-The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice.
-As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently.
-We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.
-
-\paragraph*{Acknowledgements.}
-
-The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
-
-\bibliographystyle{plainurl}
-\bibliography{paper}
-
-\clearpage
-\appendix
-\section{Proofs of Lemmas}\label{appendix:proofs}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:pg-fn} (No false negatives)]
-Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
-The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
-
-For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
-By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
-Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
-As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
-Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:pg-context} (Context length)]
-If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
-Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ bet the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
-
-Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1])$.
-There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
-Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$ for all nodes $u' \in M_{i}$.
-
-Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m]$.
-If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
-Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1]$ be a substring represented by node $u'$ in the \kcollection{k} used for building the path graph.
-As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
-
-Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$.
-If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
-Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:pg-keys} (Short keys)]
-(a) If node $v'$ has multiple predecessors with label $c$, the keys of the predecessors must be longer than string $c \cdot G'.\gkey(v')$, as the key set is prefix-free.
-
-(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-fp} (No short false positives)]
-We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
-
-Let $v' \in G'.\find(X)$ be a node.
-By Lemma~\ref{lemma:pg-keys}, pattern $X$ prefix-matches key $G'.\gkey(v')$.
-For every node $v \in G'.\gvalue(v')$, there is a substring $S[i, i+k-1] = G'.\gkey(v'_{0})$ in the \kcollection{k} $\mathcal{S}$ used for building graph $G'$, with $\mathcal{S}.\gnode(S, i) = v$.
-Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-prune} (Pruning)]
-Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
-
-Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
-By Lemma~\ref{lemma:pg-keys}, $X$ is a prefix of the substring.
-We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
-The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
-If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
-
-Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
-If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
-Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$.
-Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.
-
-Let $Y$ be the longer of strings $G''.\gkey(v''_{j})$ and $X[j] \cdot G''.\gkey(v''_{j+1})$.
-As $Y$ is a substring of $\mathcal{S}$, there is a node $v_{j} \in V_{d}$ such that key $G_{d}.\gkey(v_{j})$ prefix-matches $Y$.
-Key $G_{d}.\gkey(v_{j})$ has both $X[j, \abs{X}-1]$ and $G''.\gkey(v''_{j})$ as its prefixes, and hence $v_{j} \in V_{d}(v''_{j})$.
-When $j = 0$, we can extend key $G_{d}.\gkey(v_{0})$ to get a path starting from $v_{0}$ and matching $X$ in graph $G_{d}$.
-
-Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
-We can also start from path $P''$ and determine the node $v_{0}$.
-As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
-Let $G_{d} = (V_{d}, E_{d})$ the de~Bruijn graph of graph $G$ with the same order $k$ as graph $G'$, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
-If $v' \in V'$ is a node, then $G_{d}.\gvalue(v) = G'.\gvalue(v')$ for all nodes $v \in V_{d}(v')$.
-
-Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$.
-Then string $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$.
-There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
-
-Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}.
-Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
-As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.
-
-The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$.
-Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
-\end{proof}
 
@@
 
+\section{Suffix Tree of a Path Graph}\label{appendix:extensions}
+
+Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
+We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
+The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
+The compacted trie can be considered the \emph{suffix tree} of graph $G'$.
+
+The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
+Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
+These ranges can be determined from the LCP array.
+If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
+
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+By selecting $x$ according to cache line size, the queries require $\Oh{\log_{x} \abs{V'}}$ random memory accesses, which in practice dominate the computation time.
+Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
+By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
+Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}
+
+BWA-MEM \cite{Li2013} aligns reads to a reference genome by finding \emph{maximal exact matches} between the read and the reference.
+It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions.
+GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$).
+However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
+
+In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range.
+In GCSA, each node $v' \in V'$ may have multiple values (pointers), and a value may occur in multiple nodes.
+As $\locate$ queries can be slow, we need another way to support efficient \emph{counting queries}.
+The problem is similar to determining the \emph{document frequency} of a pattern.
+Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
+We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+
+Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$, and let $\countq(v)$ be the document frequency of the label $\ell(v)$.
+If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, the number of \emph{redundant} documents in them is $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$.
+We create an array $R[0, n-2]$ based on the \emph{inorder traversal} of the suffix tree.
+If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$.
+We set $R[j] = 0$ for any subsequent visits to the same node.
+Range $R[sp, ep-1]$ covers the internal nodes in the subtree with node $v$ as the root.
+We can determine document frequencies as $\countq(v) = (ep+1-sp) - \sum R[sp, ep-1]$.
+If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute the sums with $\select$ queries:
+$$
+\sum R[a, b] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
+$$
+
+For value counting, we use array $R$ to store the number of redundant values as above.
+We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above.
+The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
+$$
+G'.\countq(X) = G'.\countq([sp, ep]) = (ep+1-sp) + \sum A[sp, ep] - \sum R[sp, ep-1].
+$$
+The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
+
 \section{Alternate GCSA Encoding}\label{appendix:encoding}

### 2016-08-14 paper/paper.tex
Almost final submission.

@@
 \usepackage{microtype}
-\usepackage{hyperref}
+
+% hyperref does not work with LNCS
+%\usepackage{hyperref}
+\usepackage{url}
 
@@
 
-There are also several structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015,Na2016}) using graphs as a space-efficient way of indexing similar sequences.
+There are also structures (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015,Na2016}) using graphs as a space-efficient way of indexing similar sequences.
 While the problem is different from indexing the paths in a graph, the techniques used are similar.
@@
 The pruned graph is encoded with a generalization of the FM\nobreakdash-index.
-GCSA2 often uses around 1~bit per \kmer{k} with large values of $k$.
-Explicit de~Bruijn graphs cannot achieve similar space usage, as they store some information for each \kmer{k}.
-GCSA2 also includes extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph.
+GCSA2 often uses less space (e.g.~1~bit per \kmer{k}) than other de~Bruijn graph-based indexes, which have to store some information for each \kmer{k} explicitly.
+The index also includes extensions based on compressed suffix trees.
 The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
@@
 
-\subsection{Text indexes}
+\subsection{Text Indexes}
 
@@
 It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges.
-If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$.
 Although fast and versatile, suffix trees are impractical with large texts, as they require much more space than the text itself.
@@
 \end{itemize}
-We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
+We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
 \end{definition}
@@
 
-\subsection{FM-index for graphs}
+\subsection{FM-Index for Graphs}
 
@@
 
-\subsection{De Bruijn graphs as path indexes}
+\subsection{De Bruijn Graphs as Path Indexes}
 
@@
 
-\subsection{Path graphs}
+\subsection{Path Graphs}
 
@@
 
-\subsection{Pruned de Bruijn graphs}
+\subsection{Pruned de Bruijn Graphs}
 
@@
 Right: GCSA for graph $G''$.
-Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$.
-Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
+Leftward arrows show how backward searching works, with the red arrows showing it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$.
+Rightward arrows match nodes with sampled values, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
 \end{figure}
@@
 \end{align*}
-In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the values of the predecessor.
+In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the predecessor.
 We may also sample other values to improve query performance.
@@
 
-GCSA2 has been developed as the path indexing library of vg \cite{Garrison2014-2016}.
-The implementation is written in C++, and the source code is available on GitHub (\href{https://github.com/jltsiren/gcsa2}{https://github.com/jltsiren/gcsa2}).
-As the implementation uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
-The index uses fast succinct data structures instead of the smaller but slower compressed data structures for most of its components.
-The experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0.
-All files were stored on a distributed Lustre file system.
+GCSA2 is the path indexing library of vg \cite{Garrison2014-2016}.
+The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
+Most index components are encoded using fast non-compressed bitvectors.
+As GCSA2 uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+
+We used a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory for the experiments, and stored all files on a distributed Lustre file system.
+The system was running Ubuntu~12.04 on Linux kernel~3.2.0.
 We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.
 
+\subsection{Construction}
+
 The vg \emph{variation graphs} have strings as node labels.
-Each node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
-When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels.
-Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples.
+A node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
+For indexing, the graph is implicitly converted into an input graph with single-character labels.
+We always sample the input graph nodes corresponding to the initial offsets of variation graph node labels.
 
@@
 
-Table~\ref{table:construction} summarizes index construction, while Table~\ref{table:indexes} contains details on the size of the indexes.
-We can build whole-genome indexes overnight using less than 96~gigabytes of memory, including disk cache.
-The indexes contain roughly $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the pruned de~Bruijn graphs use only 4.4\nobreakdash--5.8~billion nodes to represent them.
-Index size grows even more slowly with $k$, as we sample more values in lower order graphs.
+Tables~\ref{table:construction} and \ref{table:indexes} show construction requirements and index sizes, respectively.
+We can build a whole-genome index overnight using less than 96~gigabytes of memory, including disk cache.
+The index contains $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the path graph only uses 4.4\nobreakdash--5.8~billion nodes to represent them.
 For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
-Extensions from Appendix~\ref{appendix:extensions} increase index size to 1.12~bits per \kmer{k}.
+Extensions based on suffix trees increase the size to 1.12~bits per \kmer{k}.
 
@@
 \begin{center}
-\caption{GCSA2 construction with different orders of the path graph.
-Construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
+\caption{GCSA2 construction.
+Order of the path graph; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
 \begin{tabular}{c|ccc|cc}
@@
 \begin{center}
-\caption{Index sizes with different orders $k$ of the path graph.
-Number of \kmer{k}s and nodes in the pruned de~Bruijn graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
+\caption{GCSA2 index sizes.
+Order $k$ of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
 \begin{tabular}{c|cc|c|c|c}
@@
 
+Index construction uses more memory with $k = 32$ than with larger values of $k$.
+The \orderk{32} path graph has more nodes, where we cannot derive the values from the predecessor node.
+As we sample more values, we need more memory in the final phase of construction.
+With larger values of $k$, the path graph resembles the input graph better, and we sample less values.
+For the same reason, index size decreases with large values of $k$, even though the graph requires more space.
+
+\subsection{Queries}
+
 We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread.
-We also ran the same queries in an FM\nobreakdash-index (\texttt{csa\_wt<>} from SDSL with SA sample density 17) of the reference.
+We also ran the same queries in an FM\nobreakdash-index of the reference.
 The results can be seen in Table~\ref{table:benchmark}.
-The $\find$ and $\locate$ queries are several times slower than in the FM\nobreakdash-index, which uses the same SDSL components.
-A simplified encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
+The FM\nobreakdash-index (\texttt{csa\_wt<>} with SA sample density 17) uses the same SDSL components as GCSA2, allowing us to compare the fundamental techniques and ignore most implementation-specific details.
 
@@
 
-\section{Conclusions}
+With \kmer{16}s, where both indexes match a similar number of patterns, $\find$ queries in GCSA2 are about 3x slower than in the FM\nobreakdash-index.
+With longer patterns, the difference grows larger, as the query stops early when the index cannot match the pattern.
+$\select$ queries on bitvector $\bvIN$ are the main source of the difference, as they are much slower than $\rank$ queries.
+A simplified GCSA encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
+
+When the number of distinct occurrences is high, $\locate$ queries are about 5x slower than in the FM\nobreakdash-index.
+With less occurrences, unpredictable factors such as SA order sampling in the FM\nobreakdash-index or nodes with multiple values and filtering out duplicates in GCSA2 affect the results.
+Queries $\parent$ and $\countq$ are both clearly faster than a single step of backward searching.
+
+\section{Discussion}
 
 GCSA2 is a path index for variation graphs.
-It uses a de~Bruijn graph as an index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
+It uses a de~Bruijn graph as a \kmer{k} index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
 The index supports queries of length up to $k$ exactly and longer queries with false positives.
-GCSA2 includes extensions based on suffix trees; other extensions have also been considered but not implemented yet (see Appendix~\ref{appendix:hypertext}).
-The index has been integrated in the variation graph toolkit vg, where it is used for e.g.~read alignment based on maximal exact matches.
+GCSA2 also includes extensions based on suffix trees; other extensions have been considered but not implemented (see Appendix~\ref{appendix:hypertext}).
+The index is used in the variation graph toolkit vg for e.g.~read alignment based on maximal exact matches.
 
@@
 The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
-Queries are several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.
+Queries are several times slower than in a similar implementation of the FM\nobreakdash-index for sequences.
 
-The primary desing goals in GCSA2 have been query performance and index size.
+The primary desing goals in GCSA2 were query performance and index size.
 The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions.
@@
 
-The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice.
+We may want to determine whether a pattern matches known \emph{haplotypes} or only their recombinations.
+As GCSA2 does not support this directly, vg checks it afterwards using a separate structure \cite{Novak2016}.
+The \emph{FM\nobreakdash-index of alignment} \cite{Na2015,Na2016} embeds the haplotypes directly in a GCSA-like index and reports the haplotypes matching the $\find$ query.
+While the solution depends on the specific properties of the graph, it could be possible to extend it to work with any GCSA.
+
+% Maybe in the full version
+%The compressed suffix tree used in GCSA2 may also be useful as a text index, as it is both easy to implement and fast in practice.
 As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently.
@@
 
-\paragraph*{Acknowledgements.}
+\paragraph{Acknowledgements.} The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
 
-The author thanks Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions while developing the GCSA2 index.
-
-\bibliographystyle{plainurl}
+\bibliographystyle{splncs03}
 \bibliography{paper}
@@
 As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
-Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$. \qed
 \end{proof}
@@
 Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
+\qed
 \end{proof}
@@
 
-(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
+(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$. \qed
 \end{proof}
@@
 We may assume without loss of generality that graph $G'$ is a de~Bruijn graph.
-
 Let $v' \in G'.\find(X)$ be a node.
@@
 Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
+\qed
 \end{proof}
@@
 If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
-Now assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
+Assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
 As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$.
@@
 As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
+\qed
 \end{proof}
@@
 Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
+\qed
 \end{proof}
@@
 
-Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V, E)$ and the last character array $L$.
+Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V_{d}, E_{d})$ and the last character array $L$.
 If the lexicographic rank of \kmer{k} $X$ is $i$, the lexicographic rank of \kmer{k} $(cX)[0, k-1]$ is $G_{d}.\LF(i, c)$.
@@
 String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.
+
+\emph{Complex regions} of the variation graph must be pruned before indexing.
+While this happens before index construction begins, the construction algorithm has features that can make the pruning less destructive.
+Pruning heuristics often create regions that are completely missing from the index.
+The same mechanism that saves memory by having each chromosome in a separate file can be used to index \emph{overlapping subgraphs} without indexing any paths between them.
+By having the pruned graph in one file and the reference path in another file, we can guarantee that no region is completely missing from the index.
+We can also index \emph{selected paths} in complex regions by duplicating nodes for prefix-doubling and mapping the duplicates back to the original nodes during the merging step.
+If we index the paths corresponding to \emph{known haplotypes}, we can guarantee that the index contains all observed variation.
 
@@
 The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree.
-Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$.
+Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the path label $\ell(v)$ from the root to the node.
 These ranges can be determined from the LCP array.
-If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
+If we use the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
 
@@
 We can avoid the false negatives with a generalization of the hypertext index.
-Instead of pruning the input graph heuristically, we create a primary graph based on known \emph{haplotypes} and build a GCSA index for both strands of the graph.
+Instead of pruning the input graph heuristically, we create a primary graph based on known haplotypes and build a GCSA index for both strands of the graph.
 We then create a matrix of additional edges corresponding to potential \emph{recombinations} in the path graph, always crossing from the reverse complement strand to the forward strand.

### 2016-08-15 paper/paper.tex
The submitted version.

@@
 Path indexes are one of the most important tools for working with variation graphs.
-They generalize text indexes for graphs, allowing one to find the paths matching the query string.
+They generalize text indexes to graphs, allowing one to find the paths matching the query string.
 We propose using de Bruijn graphs as path indexes, compressing them by merging redundant subgraphs, and encoding them with the Burrows-Wheeler transform.
@@
 Sirén et~al.
-developed \emph{GCSA}~\cite{Siren2014}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) for directed acyclic graphs.
+developed \emph{GCSA}~\cite{Siren2014}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) to directed acyclic graphs.
 GCSA depends on pruning the complex regions, as there are no limits on query length.
@@
 The pruned graph is encoded with a generalization of the FM\nobreakdash-index.
-GCSA2 often uses less space (e.g.~1~bit per \kmer{k}) than other de~Bruijn graph-based indexes, which have to store some information for each \kmer{k} explicitly.
-The index also includes extensions based on compressed suffix trees.
+GCSA2 often uses less space (e.g.~less than 1~bit per \kmer{k}) than other de~Bruijn graph-based indexes, which have to store some information for each \kmer{k} explicitly.
+The index also includes extensions based on suffix trees.
 The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
@@
 Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
-The notion of substring generalizes in a natural way for infinite sequences.
+The notion of substring generalizes in a natural way to infinite sequences.
 A substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
@@
 $$
-where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition for any character $c \in \Sigma$:
+where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition to any character $c \in \Sigma$:
 $$
@@
 
-We generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
+We generalize the definition to infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
 We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
@@
 \item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
-\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
+\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = XY[k-1]$ occurring in $\mathcal{S}$.
 \end{itemize}
@@
 Suffix trees, suffix arrays, and the \FMindex{} can be generalized to index multiple texts.
-There are also generalizations for other combinatorial structures.
-The \emph{XBW transform} \cite{Ferragina2009b} is essentially an \FMindex{} for \emph{labeled trees}.
+There are also generalizations to other combinatorial structures.
+The \emph{XBW transform} \cite{Ferragina2009b} is an \FMindex{} for \emph{labeled trees}.
 The nodes of the tree are sorted by path labels from the node to the root.
-$\BWT$ stores the labels of the children of each node (leaf nodes require special treatment).
+$\BWT$ stores the labels of the children of each node, while leaf nodes are marked with special characters.
 If a node has $k$ children, we encode that as a binary sequence $0^{k-1} 1$.
@@
 On the other hand, we gain the ability to search for patterns shorter than $k$ characters.
-The main drawback of these basic \kmer{k} indexes is their size, as they have to store the key-value pairs explicitly.
+The main drawback of these basic \kmer{k} indexes is their size, as they store the key-value pairs explicitly.
 
@@
 We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ matching pattern $G'.\gkey(v)$ in graph $G$.
-There will be no \emph{false negatives} (path labels that exist in the input graph but not in the index), regardless of pattern length.
+The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index), regardless of pattern length.
 There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results in the input graph.
@@
 We can encode the de~Bruijn graph as a GCSA using $(\log \sigma + 2)(1 + \oh{1})$ bits per \kmer{k}.
-By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that typically uses a couple of bytes per \kmer{k} (see the \orderk{32} index in Section~\ref{sect:experiments}).
+By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses typically a couple of bytes per \kmer{k} (see the \orderk{32} index in Section~\ref{sect:experiments}).
 While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.
@@
 \begin{definition}[Path graph]
-Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
-Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$.
+Let $\mathcal{S}$ be the \kcollection{k} of the labels of the maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
+Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$ and that $\$^{k} \in \mathcal{S}$.
 The \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$, where
@@
 Right: GCSA for graph $G''$.
-Leftward arrows show how backward searching works, with the red arrows showing it from pattern $\mathtt{T}$ to pattern $\mathtt{AT}$.
-Rightward arrows match nodes with sampled values, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
+Leftward arrows illustrate backward searching, with the red arrows showing it from $\mathtt{T}$ to $\mathtt{AT}$.
+Rightward arrows mark the samples belonging to each node, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
 \end{figure}
@@
 See Figure~\ref{figure:pruned-index} for an example.
-If lexicographic range $[sp_{i+1}, ep_{i+1}]$ matches suffix $X[i+1, \abs{X}-1]$ of pattern $X$, we can find the range matching suffix $X[i, \abs{X}-1]$ as
+If lexicographic range $[sp_{i+1}, ep_{i+1}]$ matches suffix $X[i+1, \abs{X}-1]$ of pattern $X$, we can find the range $[sp_{i}, ep_{i}]$ matching suffix $X[i, \abs{X}-1]$ as
 \begin{align*}
@@
 
-The vg \emph{variation graphs} have strings as node labels.
+The vg \emph{variation graphs} use strings as node labels.
 A node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
@@
 \caption{GCSA2 index sizes.
-Order $k$ of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
+Order $k$ of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (the graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
 \begin{tabular}{c|cc|c|c|c}
@@
 
-With \kmer{16}s, where both indexes match a similar number of patterns, $\find$ queries in GCSA2 are about 3x slower than in the FM\nobreakdash-index.
+With \kmer{16}s, where both indexes match a similar number of patterns, $\find$ queries are about 3x slower in GCSA2 than in the FM\nobreakdash-index.
 With longer patterns, the difference grows larger, as the query stops early when the index cannot match the pattern.
@@
 
-The primary desing goals in GCSA2 were query performance and index size.
+The primary desing goals for GCSA2 were query performance and index size.
 The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions.
@@
 We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$.
-Then we apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.
+We then apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.
 
@@
 
-After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we proceed to \emph{building the index}.
+After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we \emph{build the index}.
 Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing.
@@
 If \kmer{2k} $X$ is encoded with \kmer{k} ranks $(i, j)$, we can encode string $cX$ as a lexicographic range of \kmer{k} rank sequences, with $(G_{d}.\LF(i, c), G_{d}.\LF(j, L[i]), G_{d}.\LF(0, L[j]))$ as the lower bound and $G_{d}.\LF(\abs{V_{d}}-1, L[j])$ as the last rank in the upper bound.
-String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.
+String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only if the lexicographic ranges of the \kmer{k} rank sequences overlap.
 
@@
 We can also index \emph{selected paths} in complex regions by duplicating nodes for prefix-doubling and mapping the duplicates back to the original nodes during the merging step.
-If we index the paths corresponding to \emph{known haplotypes}, we can guarantee that the index contains all observed variation.
+If we index the paths corresponding to \emph{known haplotypes} in complex regions, we can guarantee that the index contains all observed variation.
 
@@
 The hypertext index \cite{Thachuk2013} is based on graphs $G = (V, E)$, where the label $G.\glabel(v)$ of a node $v \in V$ is a string over alphabet $\Sigma$.
-The labels are indexes in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$.
+The labels are indexed in FM\nobreakdash-index $F$, while the reverse labels are indexed in FM\nobreakdash-index $R$.
 When we search for a pattern $X \in \Sigma^{\ast}$ of length $\abs{X} \ge 2$, some of the matches may cross edges.

### 2016-10-07 paper/paper.tex
Updating the paper.

@@
 The extensions are used for e.g.~finding maximal exact matches in the vg read aligner.
+
+The main differences to the original GCSA are:
+\begin{enumerate}
+
+\item The graph encoding in GCSA2 has been optimized for small alphabets, improving query performance by up to an order of magnitude.
+
+\item The construction algorithm stores the graphs on disk, reducing the memory requirements for building a whole-genome human index from hundreds of gigabytes to tens of gigabytes.
+
+\item GCSA2 can index denser graphs, including cyclic graphs, by limiting maximum query length.
+
+\item GCSA2 extends the functionality of the FM-ideas with suffix tree operations.
+
+\end{enumerate}
+The first two improvements can also be used with the original GCSA.
+The third point represents a different approach to indexing graphs.
+It was not possible the with the original GCSA, as the construction algorithm required a prefix-range-sorted graph.
+The extendend functionality depends on the new construction algorithm and on limiting maximum query length.
 
@@
 
-GCSA2 construction starts from paths of length $k$ in the input graph.
-We build a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph using a similar \emph{prefix-doubling} algorithm as in the original GCSA \cite{Siren2014}, and encode the result as a GCSA.
-To avoid excessive memory usage, we keep the paths and the graphs on disk, and read only a single chromosome at a time into memory.
-The details of the construction algorithm can be found in Appendix~\ref{appendix:construction}.
-
 As in the original GCSA, we sort the nodes of the path graph in lexicographic order, encode the indegrees and outdegrees in bitvectors $\bvIN$ and $\bvOUT$, and store the predecessor labels in $\BWT$.
@@
 
+GCSA2 construction starts from paths of length $k$ in the input graph.
+We build a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph using a similar \emph{prefix-doubling} algorithm as in the original GCSA \cite{Siren2014}, and encode the result as a GCSA.
+To avoid excessive memory usage, we keep the paths and the graphs on disk, and read only a single chromosome at a time into memory.
+The details of the construction algorithm can be found in Appendix~\ref{appendix:construction}.
+
+We can improve the query performance with (maximally pruned) de~Bruijn graphs by using a \emph{simplified encoding} (see Appendix~\ref{appendix:encoding}).
+We replace $\BWT$ and bitvector $\bvIN$ with bitvectors $B_{c}$ for all $c \in \Sigma$ such that $B_{c}[j] = 1$ if and only if node $j$ in lexicographic order has a predecessor with label $c$.
+This simplifies backward searching to
+\begin{align*}
+sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
+ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1;\ \textrm{and} \\
+[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+\end{align*}
+
 The compacted trie of keys resembles a suffix tree.
@@
 The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
-Most index components are encoded using fast non-compressed bitvectors.
-As GCSA2 uses \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile.
+It depends on \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}.
+We use the simplified encoding (Appendix~\ref{appendix:encoding}) with fast non-compressed bitvectors in most index components.
+Bitvectors $B_{c}$ for rare characters ($\baseN$, $\#$, and $\$$) are compressed as sparse bitvectors \cite{Okanohara2007}.
 
@@
 The system was running Ubuntu~12.04 on Linux kernel~3.2.0.
-We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.7 for the benchmarks.
+We used vg version~1.3.0 for processing the graphs and GCSA2 version~0.8 using SDSL version~2.1.1 for the benchmarks.
+All code was compiled with gcc/g++ version~4.9.2.
 
@@
 
-We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, and queried for them in the \orderk{128} index using a single thread.
-We also ran the same queries in an FM\nobreakdash-index of the reference.
+We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence.
+SSA is the SDSL implementation (\texttt{csa\_wt<>} with SA sample period 17) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT.
+As the default FM-index in SDSL, it prioritizes query performance over compression.
+BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with SA sample period 32).
+Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.
+
+As building the original GCSA requires around $65n$ bytes of memory for a path graph with $n$ nodes, we could not compare GCSA and GCSA2 directly on a system with 256~gigabytes of memory.
+Instead, we used \emph{RLCSA} \cite{Maekinen2010} (May~2016 version with SA sample period 32) as a proxy.
+The RLCSA is an FM-index for repetitive sequence collections using the same basic components as the original GCSA.
+With a mixed query load, RLCSA is 1.5x to 3x faster than GCSA, depending on algorithmic overhead and the mix of $\find$ and $\locate$ queries \cite{Siren2014}.
+
+We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, filtered out the \kmer{k}s consisting entirely of $\baseN$s, and queried for the remaining \kmer{k}s using a single thread.
 The results can be seen in Table~\ref{table:benchmark}.
-The FM\nobreakdash-index (\texttt{csa\_wt<>} with SA sample density 17) uses the same SDSL components as GCSA2, allowing us to compare the fundamental techniques and ignore most implementation-specific details.
 
@@
 \begin{center}
-\caption{Query benchmarks using an \orderk{128} GCSA and an FM\nobreakdash-index.
-Index type and pattern length; number of patterns, matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
-\begin{tabular}{c|cccc|cccc}
+\caption{Query benchmarks using an \orderk{128} GCSA2 and various FM\nobreakdash-indexes.
+Pattern length; the number of patterns; index type; the number matching patterns, matching nodes, and distinct occurrences; the average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and the average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
+\begin{tabular}{cc|cccc|cccc}
 \hline
- & \textbf{Patterns} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
+$k$ & \textbf{Patterns} & \textbf{Index} & \textbf{Found} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
 \hline
-GCSA:  16 & 351584 & 347453 & 2477M &  872M & 18.3 \textmu{}s & 0.41 \textmu{}s & 0.87 \textmu{}s & 12.1 \textmu{}s \\
-GCSA:  32 & 351555 & 333258 &  112M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 11.1 \textmu{}s \\
-GCSA:  64 & 351567 & 326101 & 2.63M & 1.35M & 76.1 \textmu{}s & 0.26 \textmu{}s & 0.28 \textmu{}s & 5.76 \textmu{}s \\
-GCSA: 128 & 351596 & 316500 & 0.32M & 0.37M &  149 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 5.98 \textmu{}s \\
+ 16 & 351584 & GCSA2 & 347453 & 2477M &  872M & 4.75 \textmu{}s & 0.42 \textmu{}s & 0.87 \textmu{}s & 5.85 \textmu{}s \\
+    &        & SSA   & 301538 &    -- &  782M & 6.00 \textmu{}s &              -- &              -- & 2.43 \textmu{}s \\
+    &        & BWA   & 320764 &    -- & 1564M & 3.64 \textmu{}s &              -- &              -- & 4.65 \textmu{}s \\
+    &        & RLCSA & 301538 &    -- &  782M & 23.7 \textmu{}s &              -- &              -- & 8.12 \textmu{}s \\
 \hline
-FM:    16 & 351584 & 301538 &    -- &  782M & 6.06 \textmu{}s &              -- &              -- & 2.46 \textmu{}s \\
-FM:    32 & 351555 & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.19 \textmu{}s \\
-FM:    64 & 351567 &  88184 &    -- & 0.84M & 17.2 \textmu{}s &              -- &              -- & 1.91 \textmu{}s \\
-FM:   128 & 351596 &  35678 &    -- & 0.08M & 23.6 \textmu{}s &              -- &              -- & 3.48 \textmu{}s \\
+ 32 & 351555 & GCSA2 & 333258 &  112M & 34.3M & 10.8 \textmu{}s & 0.28 \textmu{}s & 0.38 \textmu{}s & 5.44 \textmu{}s \\
+    &        & SSA   & 153957 &    -- & 26.6M & 10.9 \textmu{}s &              -- &              -- & 2.16 \textmu{}s \\
+    &        & BWA   & 156080 &    -- & 52.9M & 6.57 \textmu{}s &              -- &              -- & 3.19 \textmu{}s \\
+    &        & RLCSA & 153957 &    -- & 26.6M & 47.6 \textmu{}s &              -- &              -- & 5.87 \textmu{}s \\
+\hline
+  64 & 351567 & GCSA2 & 326101 & 2.63M & 1.35M & 22.5 \textmu{}s & 0.26 \textmu{}s & 0.29 \textmu{}s & 2.92 \textmu{}s \\
+     &        & SSA   &  88184 &    -- & 0.84M & 17.1 \textmu{}s &              -- &              -- & 1.89 \textmu{}s \\
+     &        & BWA   &  88786 &    -- & 1.60M & 10.3 \textmu{}s &              -- &              -- & 2.34 \textmu{}s \\
+     &        & RLCSA &  88184 &    -- & 0.84M & 74.3 \textmu{}s &              -- &              -- & 5.97 \textmu{}s \\
+\hline
+ 128 & 351596 & GCSA2 & 316500 & 0.32M & 0.37M & 45.3 \textmu{}s & 0.26 \textmu{}s & 0.26 \textmu{}s & 3.13 \textmu{}s \\
+     &        & SSA   &  35678 &    -- & 0.08M & 23.5 \textmu{}s &              -- &              -- & 3.47 \textmu{}s \\
+     &        & BWA   &  35741 &    -- & 0.12M & 14.0 \textmu{}s &              -- &              -- & 3.46 \textmu{}s \\
+     &        & RLCSA &  35678 &    -- & 0.08M & 91.7 \textmu{}s &              -- &              -- & 12.9 \textmu{}s \\
 \hline
@@
 
-With \kmer{16}s, where both indexes match a similar number of patterns, $\find$ queries are about 3x slower in GCSA2 than in the FM\nobreakdash-index.
-With longer patterns, the difference grows larger, as the query stops early when the index cannot match the pattern.
-$\select$ queries on bitvector $\bvIN$ are the main source of the difference, as they are much slower than $\rank$ queries.
-A simplified GCSA encoding (see Appendix~\ref{appendix:encoding}) would make $\find$ queries faster and $\locate$ queries slower.
+Backward searching in an FM-index stops early if there are no matches.
+In order to compare the $\find$ performance of the indexes reliably, we must hence concentrate on the \kmer{16}s, where the fraction of matching patterns is similar for all indexes.
+GCSA2 and the fast FM-indexes (SSA and BWA) all have similar performance, while RLCSA is several times slower.
+As a result, we can estimate that $\find$ queries in GCSA2 are an order of magnitude faster than in GCSA.
 
-When the number of distinct occurrences is high, $\locate$ queries are about 5x slower than in the FM\nobreakdash-index.
-With less occurrences, unpredictable factors such as SA order sampling in the FM\nobreakdash-index or nodes with multiple values and filtering out duplicates in GCSA2 affect the results.
-Queries $\parent$ and $\countq$ are both clearly faster than a single step of backward searching.
+When comparing the $\locate$ performance of different FM-indexes, the distribution of the query positions should be close to uniform.
+Otherwise the biases from e.g.~different suffix array sampling strategies or the variation in the number of distinct occurrences per node in GCSA2 can make the results unreliable.
+As the \kmer{k}s have been sampled uniformly from the variation graph, we get the best results with the \kmer{16}s, where all indexes can match most of the patterns.
+
+GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6.
+On the average, each sample contains 1.53~values and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7.
+SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates.
+BWA is closer to GCSA2 in $\locate$ performance, which can be explained by sparser sampling than in SSA.
+RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges.
+Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does not have the optimizations, we can estimate that GCSA is 4x slower than GCSA2.
+
+The remaining queries, $\parent$ and $\countq$, take a fraction of a microsecond.
+As a $\parent$ query takes comparable time to a single step of backward searching, it will not be a bottleneck in finding maximal exact matches.
+Counting the number of distinct occurrences with a $\countq$ query is clearly faster than retrieving even a single occurrence.
 
@@
 The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
-Queries are several times slower than in a similar implementation of the FM\nobreakdash-index for sequences.
+Query performance is comparable to that of fast FM-indexes for sequences.
 
-The primary desing goals for GCSA2 were query performance and index size.
-The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece, and provides several options for dealing with complex regions.
+The primary design goals for GCSA2 were query performance and index size.
+The index works with arbitrary graphs, supports queries that are long enough to map short reads in one piece without false positives, and provides several options for dealing with complex regions.
 Other path indexes work with a more restricted class of graphs \cite{Siren2014,Huang2013,Kim2015-2016,Maciuca2016}, are at least an order of magnitude slower \cite{Huang2013,Maciuca2016}, require much more space \cite{Bowe2012,Pell2012,Cazaux2014,Marcus2014}, or are theoretical proposals that have never been implemented \cite{Thachuk2013}.
@@
 We may want to determine whether a pattern matches known \emph{haplotypes} or only their recombinations.
-As GCSA2 does not support this directly, vg checks it afterwards using a separate structure \cite{Novak2016}.
+As GCSA2 does not support this directly, vg must determine this afterwards using a separate structure \cite{Novak2016}.
 The \emph{FM\nobreakdash-index of alignment} \cite{Na2015,Na2016} embeds the haplotypes directly in a GCSA-like index and reports the haplotypes matching the $\find$ query.
@@
 
+\section{Simplified GCSA Encoding}\label{appendix:encoding}
+
+Let $G' = (V', E')$ be a path graph with $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$.
+This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}).
+We can use the simplified encoding of the original GCSA \cite{Siren2014} for the GCSA of such path graphs.
+
+We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
+If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
+Backward searching is simplified to:
+\begin{align*}
+sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
+ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1;\ \textrm{and} \\
+[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+\end{align*}
+We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a cheap $B_{c}.\rank()$.
+
+Computing $G'.\LF(i)$ can more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine the character used in the backward step.
+However, if the alphabet is small, this is still faster than the $\select$ queries in the general encoding.
+We can further improve the time/space trade-off by compressing the bitvectors $B_{c}$ for rare characters (e.g.~$\baseN$, $\#$, and $\$$) and checking $B_{c}[i]$ first for the frequent characters (e.g.~bases) when computing $G'.\LF(i)$.
+
 \section{Suffix Tree of a Path Graph}\label{appendix:extensions}
@@
 
-\section{Alternate GCSA Encoding}\label{appendix:encoding}
-
-Let $G' = (V', E')$ be a path graph with $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$.
-This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}).
-We can use the simplified encoding of the original GCSA \cite{Siren2014} for the GCSA of such path graphs.
-
-We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
-If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
-Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
-$$
-G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
-$$
-We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector.
-On the other hand, computing $G'.\LF(i)$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$.
-The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.
-
 \section{Using the Hypertext Index}\label{appendix:hypertext}

### 2016-10-10 paper/paper.tex
Paper in ALENEX format

@@
-\documentclass[a4paper,11pt]{llncs}
+\documentclass[twoside,leqno,twocolumn]{article}
+\usepackage{ltexpprt}
 
@@
 
-\usepackage{fullpage}
 \setlength\tabcolsep{4pt}
@@
 
-% hyperref does not work with LNCS
-%\usepackage{hyperref}
 \usepackage{url}
@@
 
-\title{Indexing Variation Graphs\thanks{This work was supported by the Wellcome Trust grant 098051.}}
+\title{Indexing Variation Graphs\thanks{Supported by the Wellcome Trust grant 098051.}}
 
-\author{Jouni Sirén}
-\institute{Wellcome Trust Sanger Institute, UK \\ \email{jouni.siren@iki.fi}}
+\author{Jouni Sirén\thanks{Wellcome Trust Sanger Institute, UK. \texttt{jouni.siren@iki.fi}.}}
+
+\date{}
 
@@
 
-\thispagestyle{empty}
-\setcounter{page}{0}
 \maketitle
 
-\begin{abstract}
+\begin{abstract}\small\baselineskip=9pt
 Variation graphs, which represent genetic variation within a population, are replacing sequences as reference genomes.
@@
 
-\clearpage
 \section{Introduction}
@@
 
-Given string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$.
-We also define $S.\select(i, c) = \arg \max_{j \le n} S.\rank(j, c) < i$ as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}.
+Given string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define
+$$
+S.\select(i, c) = \max \set{ j \le n \mid S.\rank(j, c) < i }
+$$
+as the position of the occurrence of character $c$ with rank $i > 0$.\footnote{These definitions are used in the SDSL library \cite{Gog2014b}.
 We assume for convenience that $S.\select(0, c) = -1$.} A \emph{bitvector} is a binary sequence supporting efficient $\rank$/$\select$ queries.
@@
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \centering
@@
 \caption{\protect\LFmapping, BWT, suffix array, and LCP array for text $\mathtt{GCATCATA}\$$.}\label{figure:indexes}
-\end{figure}
+\end{figure*}
 
@@
 $$
-where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We can also generalize the definition to any character $c \in \Sigma$:
+where $\Carray[c]$ is the number of occurrences of characters $c' < c$ in the BWT. Then $\SA[\LF(i)] = (\SA[i]-1) \bmod n$. We generalize the definition to any character $c \in \Sigma$:
 $$
@@
 $$
-Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes smaller than string $cX$ in lexicographic order.
+Let $X$ be a string. If there are $i$ suffixes $S'$ of text $T$ such that $S' < X$ in lexicographic order, then there are $\LF(i, c)$ suffixes $S'$ such that $S' < cX$.
 
@@
 To distinguish these technical nodes from the actual nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$, which are not used anywhere else in the graph.
-We add an edge $(s, v)$ to all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ from all nodes $v \in V \setminus \set{t}$ with no other outgoing edges.
+We add an edge $(s, v)$ to all nodes $v \in V \setminus \set{s}$ with no incoming edges, and an edge $(v, t)$ from all nodes $v \in V \setminus \set{t}$ with no outgoing edges.
 We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.
@@
 
-\begin{definition}[\kcollection{k}]
+\begin{Definition}[\kcollection{k}]
 Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k > 0$ be a parameter value.
@@
 (c) contains no other occurrences of characters $\#$ and $\$$.
-\end{definition}
+\end{Definition}
 
-\begin{definition}[de~Bruijn graph]
-Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
+\begin{Definition}[de~Bruijn graph]
+Let $k > 0$, and let $\mathcal{S}$ be a \kcollection{k}. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
 \begin{itemize}
@@
 We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
-\end{definition}
+\end{Definition}
 
@@
 If we store the pairs as a sorted list, queries become slower, as they require binary search.
-On the other hand, we gain the ability to search for patterns shorter than $k$ characters.
+On the other hand, we can also search for patterns shorter than $k$ characters.
 The main drawback of these basic \kmer{k} indexes is their size, as they store the key-value pairs explicitly.
@@
 
-\begin{definition}[De~Bruijn graph of a graph]\label{def:dbg}
+\begin{Definition}[de~Bruijn graph of a graph]\label{def:dbg} ~\\
 Let $G$ be a labeled graph, and let $\mathcal{S}$ be the \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is the \orderk{k} de~Bruijn graph of graph $G$.
-\end{definition}
+\end{Definition}
 
 Let $G' = (V', E')$ be a de~Bruijn graph of graph $G = (V, E)$, and let $\mathcal{S}$ be the \kcollection{k} used to define it.
-We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value}: $G'.\gvalue(v) = \set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.$ Apart from some technicalities near the source/sink nodes, set $G'.\gvalue(v)$ is the set of start nodes of the paths $P$ matching pattern $G'.\gkey(v)$ in graph $G$.
-The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index), regardless of pattern length.
+We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value} $G'.\gvalue(v)$ defined as the set
+$$
+\set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
+$$
+Apart from some technicalities near the source/sink nodes, $G'.\gvalue(v)$ is the set of start nodes of the paths matching pattern $G'.\gkey(v)$ in graph $G$.
+The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index).
 There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results in the input graph.
@@
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \includegraphics[width=\textwidth]{gcsa2_graph_dbg.pdf}
@@
 The highlighted path in the de~Bruijn graph is a false positive, as it consists of two disjoint paths in the input graph.}\label{figure:graph-dbg}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{definition}[Path graph]
-Let $\mathcal{S}$ be the \kcollection{k} of the labels of the maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
+\begin{Definition}[Path graph]
+Let $k > 0$, let $\mathcal{S}$ be the \kcollection{k} of the labels of the maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of substrings of length $k$ or less from $\mathcal{S}$.
 Assume that each right-maximal substring $S \in \mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$ and that $\$^{k} \in \mathcal{S}$.
@@
 We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
-\end{definition}
+\end{Definition}
 
-\begin{definition}[Path graph as an index]
+\begin{Definition}[Path graph as an index] ~\\
 Let $G = (V, E)$ be a graph, and let $G' = (V', E')$ be a path graph of $G$.
@@
 \end{itemize}
-\end{definition}
+\end{Definition}
 
 \begin{lemma}[No false negatives]\label{lemma:pg-fn}
-Let $G' = (V', E')$ be a path graph of graph $G = (V, E)$, and let $X \in \patternset$ be a pattern string.
-Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
+Let $G' = (V', E')$ be a path graph of $G = (V, E)$, and let $X \in \patternset$ be a pattern. Set $G'.\locate(X)$ contains the start nodes of all paths $P$ in graph $G$ with $G.\glabel(P) = X$.
 \end{lemma}
@@
 
-\begin{definition}[Equivalent path graphs]
+\begin{Definition}[Equivalent path graphs]
 Let $G'$ and $G''$ be two path graphs, and let $k > 0$ be a parameter value.
 We say that graphs $G'$ and $G''$ are \emph{\kequivalent{k}}, if we have $G'.\locate(X) = G''.\locate(X)$ for all patterns $X \in \patternset$ with $1 \le \abs{X} \le k$.
-\end{definition}
+\end{Definition}
 
-\begin{definition}[Pruned de~Bruijn graph]
+\begin{Definition}[Pruned de~Bruijn graph]
 Let $G$ be a graph, and let $G'$ be an \orderk{k} path graph of $G$. Path graph $G'$ is an \orderk{k} \emph{pruned de~Bruijn graph}, if it is \kequivalent{k} to the \orderk{k} de~Bruijn graph of $G$.
-\end{definition}
+\end{Definition}
 
-\begin{lemma}[No short false positives]\label{lemma:dbg-fp}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$.
-Then $G'.\locate(X)$ is a set of start nodes $v\in V$ of paths $P$ matching the pattern in graph $G$.
+\begin{lemma}[No short false positives]\label{lemma:dbg-fp} ~\\
+Let $G = (V, E)$ be a graph, let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph of graph $G$, and let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$.
+Then $G'.\locate(X)$ is a set of start nodes $v\in V$ of paths matching the pattern in graph $G$.
 \end{lemma}
@@
 \begin{lemma}[Pruning]\label{lemma:dbg-prune}
-Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v' \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v')$.
+Let $G = (V, E)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v' \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v')$.
 If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u', v' \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
@@
 
-\begin{figure}[t!]
+\begin{figure*}[t!]
 \includegraphics[width=\textwidth]{gcsa2_pruned_index.pdf}
@@
 Rightward arrows mark the samples belonging to each node, with the blue ones showing them for node $\mathtt{CAT}$.}\label{figure:pruned-index}
-\end{figure}
+\end{figure*}
 
-\begin{definition}[Maximally pruned de~Bruijn graph]
+\begin{Definition}[Maximally pruned graph] ~\\
 Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
-\end{definition}
+\end{Definition}
 
 \begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
-Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of graph $G = (V, E)$. Then $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$.
+Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of $G = (V, E)$. Then $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all $(u', v') \in E'$.
 \end{lemma}
@@
 \begin{align*}
-[sp_{in}, ep_{in}] & = [\bvIN.\select(sp_{i+1}, 1) + 1, \bvIN.\select(ep_{i+1}+1, 1)]; \\
-[sp_{out}, ep_{out}] & = [\LF(sp_{in}, X[i]), \LF(ep_{in}+1, X[i]) - 1];\ \textrm{and} \\
-[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
+[sp_{in}, ep_{in}] = [&\bvIN.\select(sp_{i+1}, 1) + 1, \\
+ & \bvIN.\select(ep_{i+1}+1, 1)]; \\
+[sp_{out}, ep_{out}] = [&\LF(sp_{in}, X[i]), \\
+ & \LF(ep_{in}+1, X[i]) - 1]; \\
+[sp_{i}, ep_{i}] = [&\bvOUT.\rank(sp_{out}, 1), \\
+ & \bvOUT.\rank(ep_{out}, 1)].
 \end{align*}
@@
 
-We can improve the query performance with (maximally pruned) de~Bruijn graphs by using a \emph{simplified encoding} (see Appendix~\ref{appendix:encoding}).
-We replace $\BWT$ and bitvector $\bvIN$ with bitvectors $B_{c}$ for all $c \in \Sigma$ such that $B_{c}[j] = 1$ if and only if node $j$ in lexicographic order has a predecessor with label $c$.
+We can improve the query performance with (maximally pruned) de~Bruijn graphs by using a \emph{simplified encoding} (Appendix~\ref{appendix:encoding}).
+We replace $\BWT$ and $\bvIN$ with bitvectors $B_{c}$ for all $c \in \Sigma$, such that $B_{c}[j] = 1$ if and only if node $j$ in lexicographic order has a predecessor with label $c$.
 This simplifies backward searching to
@@
 sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
-ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1;\ \textrm{and} \\
+ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1; \\
 [sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
@@
 GCSA2 is the path indexing library of vg \cite{Garrison2014-2016}.
-The implementation is written in C++, and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}).
-It depends on \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}.
+The implementation is written in C++, and the source code is available on GitHub.\footnote{\url{https://github.com/jltsiren/gcsa2}} It depends on \emph{SDSL} \cite{Gog2014b} and \emph{libstdc++ parallel mode}.
 We use the simplified encoding (Appendix~\ref{appendix:encoding}) with fast non-compressed bitvectors in most index components.
@@
 
-The vg \emph{variation graphs} use strings as node labels.
+\emph{Variation graphs}, as defined in vg, use strings as node labels.
 A node can be traversed in both forward and reverse complement orientations, and edges may cross between the orientations.
@@
 
-\begin{table}[t!]
+\begin{table}
 \begin{center}
-\caption{GCSA2 construction.
-Order of the path graph; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
 \begin{tabular}{c|ccc|cc}
 \hline
-\textbf{Order} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
+$k$ & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
 \hline
@@
 \end{tabular}
+\caption{GCSA2 construction.
+Order of the path graph; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
 \end{center}
@@
 
-\begin{table}[t!]
+\begin{table*}[t]
 \begin{center}
-\caption{GCSA2 index sizes.
-Order $k$ of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($\BWT$, $\bvIN$, and $\bvOUT$), the index (the graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
 \begin{tabular}{c|cc|c|c|c}
 \hline
-\textbf{Order} & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
+$k$ & \textbf{\kmer{k}s} & \textbf{Nodes} & \textbf{Graph} & \textbf{Index} & \textbf{With extensions} \\
 \hline
@@
 \end{tabular}
+\caption{GCSA2 index sizes.
+Order of the path graph; number of \kmer{k}s and nodes in the path graph in billions; index size in gigabytes and in bits per \kmer{k} for the graph ($B_{c}$ and $\bvOUT$), the index (the graph, $B_{S}$, $B_{V}$, and $V_{S}$), and the index with the extensions from Appendix~\ref{appendix:extensions}.}\label{table:indexes}
 \end{center}
-\end{table}
+\end{table*}
 
@@
 
-We extracted \kmer{16}s, \kmer{32}s, \kmer{64}s, and \kmer{128}s from the (non-pruned) vg graphs using \texttt{vg sim}, filtered out the \kmer{k}s consisting entirely of $\baseN$s, and queried for the remaining \kmer{k}s using a single thread.
+We extracted \kmer{k}s for $k \in \set{16, 32, 64, 128}$ from the (non-pruned) vg graphs by using \texttt{vg sim}, filtered out \kmer{k}s consisting entirely of $\baseN$s, and queried for the remaining \kmer{k}s using a single thread.
 The results can be seen in Table~\ref{table:benchmark}.
 
-\begin{table}[t!]
+\begin{table*}[t]
 \begin{center}
-\caption{Query benchmarks using an \orderk{128} GCSA2 and various FM\nobreakdash-indexes.
-Pattern length; the number of patterns; index type; the number matching patterns, matching nodes, and distinct occurrences; the average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and the average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
 \begin{tabular}{cc|cccc|cccc}
@@
 \end{tabular}
+\caption{Query benchmarks using an \orderk{128} GCSA2 and various FM\nobreakdash-indexes.
+Pattern length; number of patterns; index type; number matching patterns, matching nodes, and distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time per value for $\locate(sp, ep)$ queries in microseconds.}\label{table:benchmark}
 \end{center}
-\end{table}
+\end{table*}
 
@@
 
-\bibliographystyle{splncs03}
+\bibliographystyle{abbrv}
 \bibliography{paper}
 
-\clearpage
 \appendix
@@
 \begin{proof}[Lemma~\ref{lemma:pg-fn}: No false negatives]
-Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from node $v_{0} \in V$ with $G.\glabel(P) = X$.
+Let $\mathcal{S}$ be the \kcollection{k} used for building the path graph, and let $P$ be a path starting from $v_{0} \in V$ with $G.\glabel(P) = X$.
 The collection contains a sequence $S \in \mathcal{S}$ such that $S[i, i+\abs{X}-1] = X$ and $\mathcal{S}.\gnode(S, i) = v_{0}$.
@@
 For all positions $j$ with $i \le j \le i+\abs{X}-1$, there is a node $v'_{j} \in V'$ with $G'.\glabel(v'_{j}) = S[j]$ and $G'.\gkey(v'_{j}) = S[j, j+\abs{G'.\gkey(v'_{j})}-1]$.
-By definition, the path graph has an edge $(v'_{j}, v'_{j+1}) \in E'$ for $i \le j < i+\abs{X}-1$.
+By definition, path graph $G'$ has an edge $(v'_{j}, v'_{j+1}) \in E'$ for all such positions $j$.
 Hence $P' = v'_{i} \dotsm v'_{i+\abs{X}-1}$ is a path in $G'$ with $G'.\glabel(P') = X$.
 As path $P'$ starts from node $v'_{i} \in V'$, node $v'_{i}$ is included in the set $G'.\find(X)$.
-Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$. \qed
+Furthermore, $v_{0} = \mathcal{S}.\gnode(S, i) \in G'.\gvalue(v'_{i}) \subseteq G'.\locate(X)$.
 \end{proof}
@@
 
-Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1])$.
+Consider the set
+$$
+M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1]).
+$$
 There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
@@
 Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
-\qed
 \end{proof}
@@
 
-(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$. \qed
+(b) By the construction in the proof of Lemma~\ref{lemma:pg-context}, the context length for pattern $X$ in graph $G'$ is $\min(\abs{X}, \abs{G'.\gkey(v')})$.
 \end{proof}
@@
 Prefix $S[i, i+\abs{X}-1] = X$ of the substring corresponds to a path with label $X$ starting from node $v$ in graph $G$.
-\qed
 \end{proof}
@@
 
-Given $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
+Given a node $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
 We can also start from path $P''$ and determine the node $v_{0}$.
 As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
-\qed
 \end{proof}
@@
 
-The above is true for all nodes $x' \in G'.\gpred(v', G'.\glabel(u'))$.
-Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
-\qed
+The above is true for all $x' \in G'.\gpred(v', G'.\glabel(u'))$. Hence we can prune graph $G'$ further using string $G'.\glabel(u') \cdot G'.\gkey(v')$ as the new key in Lemma~\ref{lemma:dbg-prune}.
 \end{proof}
@@
 
-GCSA construction \cite{Siren2014} is based on the original \emph{prefix-doubling} algorithm for suffix array construction \cite{Manber1993}.
+GCSA construction \cite{Siren2014} is based on the \emph{prefix-doubling} algorithm for suffix array construction \cite{Manber1993}.
 The original GCSA started from paths of length $1$ in the input graph, and then repeatedly \emph{joined} paths of length $k$ into paths of length $2k$, until each path had a distinct label.
-The resulting path graph was essentially an unlimited-order pruned de~Bruijn graph and supported queries of arbitrary length.
+The resulting path graph was essentially an \orderk{\infty} pruned de~Bruijn graph and supported queries of any length.
 
@@
 sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
-ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1;\ \textrm{and} \\
+ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1; \\
 [sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
 \end{align*}
-We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a cheap $B_{c}.\rank()$.
+Two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) are replaced with a cheap $B_{c}.\rank()$.
 
-Computing $G'.\LF(i)$ can more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine the character used in the backward step.
+Computing $G'.\LF(i)$ can expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine the character used in the backward step.
 However, if the alphabet is small, this is still faster than the $\select$ queries in the general encoding.
@@
 Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern.
-We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.
+We can use a bitvector of length $2n-d-1$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute the frequencies efficiently \cite{Sadakane2007a}.
 
@@
 We can determine document frequencies as $\countq(v) = (ep+1-sp) - \sum R[sp, ep-1]$.
-If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute the sums with $\select$ queries:
+If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}$, where we can compute sums $\sum_{i=a}^{b} R[i]$ with $\select$ queries as
 $$
-\sum R[a, b] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
+(B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
 $$
@@
 The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
-$$
-G'.\countq(X) = G'.\countq([sp, ep]) = (ep+1-sp) + \sum A[sp, ep] - \sum R[sp, ep-1].
-$$
+\begin{align*}
+G'.\countq(X) & = G'.\countq([sp, ep]) \\
+ & = \sum_{i = sp}^{ep} (A[i] + 1) - \sum_{i = sp}^{ep - 1} R[i].
+\end{align*}
 The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
@@
 
-We often have to prune complex regions of the input graph before indexing it with GCSA.
+We often have to prune complex regions of the input graph before indexing it.
 This causes false negatives: paths that exist in the input graph but not in the index.
@@
 For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
-Neither option is particularly good in a reference genome.
+Neither option is good in a reference genome.
 With a single copy, we create paths in the reference that do not correspond to any valid genome.
 With the second option, we duplicate sequences that are not only identical but also functionally the same.
-One solution is to switch from graphs to \emph{context-free grammars}.
+One solution is to use \emph{context-free grammars}.
 As long as the grammar is non-nested, we can handle it with the hypertext index.

### 2016-10-12 paper/paper.tex
Another round of proofreading.

@@
 \emph{Variation graphs} (also graph genomes, genome graphs, graph references, or reference graphs), which encode the genetic variation within a population as a graph, have been proposed as a solution to the reference bias \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}.
-These graphs are expected to replace reference sequences as reference genomes.
-The shift to graph references will likely take years, as new methods and tools are needed to replace those based on linear references.
+These graphs are expected to replace sequences as reference genomes.
+The shift to graphs will likely take years, as new methods and tools are needed to replace those based on linear references.
 
@@
 Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}.
-Sirén et~al.
-developed \emph{GCSA}~\cite{Siren2014}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) to directed acyclic graphs.
+Sirén et~al.\ developed \emph{GCSA}~\cite{Siren2014}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} (a text index based on the Burrows-Wheeler transform \cite{Burrows1994}) to directed acyclic graphs.
 GCSA depends on pruning the complex regions, as there are no limits on query length.
@@
 \emph{De~Bruijn graphs} can be used as \kmer{k} indexes of other graphs.
-There are already some read aligners based on them \cite{Limasset2015,Liu2016}.
-The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph space-efficiently with a generalization of the FM\nobreakdash-index.
+There are already read aligners based on them \cite{Limasset2015,Liu2016}.
+The \emph{succinct de~Bruijn graph} of Bowe et.~al.~\cite{Bowe2012} encodes the graph with a generalization of the FM\nobreakdash-index.
 Rødland \cite{Roedland2013} proposed another similar generalization.
 Succinct de~Bruijn graphs can simulate \orderk{k} de~Bruijn graphs for multiple values of $k$ \cite{Boucher2014}, but they still need to store the graph explicitly for the largest value of $k$.
-Other space-efficient representations include compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs, which represent unary paths in the graph as single nodes or edges.
+Other representations include compacted \cite{Cazaux2014} and compressed \cite{Marcus2014} de~Bruijn graphs, which represent unary paths in the graph as single nodes or edges.
 Probabilistic de~Bruijn graphs \cite{Pell2012} use Bloom filters to support faster queries, at the expense of producing false positives.
-All these space-efficient representations require several bits per \kmer{k} for the graph itself, and more for mapping back to the indexed graph.
+All these space-efficient representations require several bits per \kmer{k} for the graph, and more for mapping back to the indexed graph.
 
@@
 
-\item The construction algorithm stores the graphs on disk, reducing the memory requirements for building a whole-genome human index from hundreds of gigabytes to tens of gigabytes.
+\item The construction algorithm stores the graphs on disk, reducing the memory requirements of building a whole-genome human index from hundreds of gigabytes to tens of gigabytes.
 
@@
 
-\item GCSA2 extends the functionality of the FM-ideas with suffix tree operations.
+\item GCSA2 extends the functionality of the FM-index with suffix tree operations.
 
@@
 Sometimes we consider \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous infinite subset of $\mathbb{Z}$.
-The notion of substring generalizes in a natural way to infinite sequences.
+The notion of substring generalizes to infinite sequences in a natural way.
 A substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right.
@@
 
-Given string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define
+Given a string $S[0, n-1]$, we define $S.\rank(i, c)$ as the number of occurrences of character $c$ in the prefix $S[0, i-1]$. We also define
 $$
@@
 We find the occurrences of \emph{pattern} string $X$ in the text in $\Oh{\abs{X} \log n}$ time by using \emph{binary search} in the suffix array.
-The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text, while its functionality is more limited than that of the suffix tree.
+The suffix array requires $n \log n$ bits of space in addition to the text, while its functionality is more limited than that of the suffix tree.
 See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.
 
-\begin{figure*}[t!]
-\centering
+\begin{figure}
+%\centering
 \includegraphics{gcsa2_text_indexes.pdf}
 \caption{\protect\LFmapping, BWT, suffix array, and LCP array for text $\mathtt{GCATCATA}\$$.}\label{figure:indexes}
-\end{figure*}
+\end{figure}
 
@@
 The \emph{\FMindex} \cite{Ferragina2005a} combines a representation of the BWT supporting $\rank$/$\select$ queries, the $\Carray$ array, and a set of \emph{sampled pointers} from the suffix array.
-It finds the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix) with a process called \emph{backward searching}.
+It uses \emph{backward searching} to find the \emph{lexicographic range} of suffixes matching pattern $X$ (having $X$ as a prefix).
 If the lexicographic range matching suffix $X[i+1, \abs{X}-1]$ of the pattern is $\SA[sp, ep]$, the range matching suffix $X[i, \abs{X}-1]$ is $\SA[\LF(sp, X[i]), \LF(ep+1, X[i]) - 1]$.
@@
 The \emph{longest-common-prefix array} (LCP array) \cite{Manber1993} is an integer array $\LCP[0, n-1]$, where each value $\LCP[i]$ tells the length of the longest common prefix of suffixes $T[\SA[i-1], n-1]$ and $T[\SA[i], n-1]$ (with $\LCP[0] = 0$).
-It is used to augment the functionality of the suffix array and the \FMindex.
-In particular, if we have the \FMindex{}, the LCP array, and the topology of the suffix tree, we get the \emph{compressed suffix tree}, which supports the full functionality of the suffix tree in a space-efficient manner \cite{Sadakane2007}.
+If we have the \FMindex{}, the LCP array, and the topology of the suffix tree, we get the \emph{compressed suffix tree}, which supports the full functionality of the suffix tree in a space-efficient manner \cite{Sadakane2007}.
 
@@
 We say that $v_{0}$ is the \emph{start} node and $v_{\abs{P}-1}$ is the \emph{end} node of the path.
-The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
+The label of path $P$ is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.
 
@@
 
-We generalize the definition to infinite paths $P = (v_{i})_{i \in Z}$ in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
+We also consider infinite paths $P = (v_{i})_{i \in Z}$, generalizing the definition in a similar way as we did with infinite character sequences in Section~\ref{sect:strings}.
 We say that path $P$ is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
@@
 The \emph{XBW transform} \cite{Ferragina2009b} is an \FMindex{} for \emph{labeled trees}.
-The nodes of the tree are sorted by path labels from the node to the root.
+The nodes of the tree are sorted by the path labels from the node to the root.
 $\BWT$ stores the labels of the children of each node, while leaf nodes are marked with special characters.
@@
 The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} extends the XBW transform to a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs.
-Before indexing, we transform the graph into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them.
+Before indexing, we transform the graph into an equivalent graph, where the nodes can be sorted unambiguously by the labels of the right-maximal paths starting from them.
 The transformation increases the size of the graph exponentially in the worst case.
-In addition to sequences $\BWT$ and $\bvIN$, GCSA also uses the outdegree bitvector $\bvOUT$, which is encoded in the same way as $\bvIN$.
+In addition to sequences $\BWT$ and $\bvIN$, GCSA also uses an outdegree bitvector $\bvOUT$, which is encoded in the same way as $\bvIN$.
 \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.
@@
 A \emph{path index} is a generalization of text indexes for \emph{labeled graphs}.
-Given a path index for \emph{input graph} $G = (V, E)$, we want to use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} the pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
+Given a path index for \emph{input graph} $G = (V, E)$, we use the index to find the start nodes $v_{0} \in V$ of the paths $P = v_{0} \dotsm v_{\abs{X}-1}$ \emph{matching} pattern $X$ (paths $P$ with $G.\glabel(P) = X$).
 
-The proofs of the lemmas in this section can be found in Appendix~\ref{appendix:proofs}.
+The proofs of the lemmas can be found in Appendix~\ref{appendix:proofs}.
 
@@
 If we store the pairs in a hash table, we can quickly search for patterns of length $k$.
-If we store the pairs as a sorted list, queries become slower, as they require binary search.
-On the other hand, we can also search for patterns shorter than $k$ characters.
+If we use binary search in a sorted list of pairs, queries become slower, but we gain the ability to search for patterns shorter than $k$ characters.
 The main drawback of these basic \kmer{k} indexes is their size, as they store the key-value pairs explicitly.
@@
 For that purpose, we define the de~Bruijn graph of graph $G = (V, E)$ by using the collection $\mathcal{S}$ of the labels of the maximal paths in the graph.
-If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$.
+If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$.
 We transform $\mathcal{S}$ into a \kcollection{k} by inserting characters $\#$ to the beginning of each non-left-infinite sequence when necessary, and characters $\$$ to the end of each non-right-infinite sequence.
@@
 Let $G' = (V', E')$ be a de~Bruijn graph of graph $G = (V, E)$, and let $\mathcal{S}$ be the \kcollection{k} used to define it.
-We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value} $G'.\gvalue(v)$ defined as the set
+We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value} $G'.\gvalue(v)$:
 $$
@@
 
-We can encode the de~Bruijn graph as a GCSA using $(\log \sigma + 2)(1 + \oh{1})$ bits per \kmer{k}.
-By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses typically a couple of bytes per \kmer{k} (see the \orderk{32} index in Section~\ref{sect:experiments}).
+We can encode the de~Bruijn graph as a GCSA using $(\abs{V'} + \oh{\abs{V'}})(\log \sigma + 2)$ bits.
+By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses a couple of bytes per \kmer{k} for a typical variation graphs (see Section~\ref{sect:experiments}).
 While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.
@@
 \begin{itemize}
-\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
+\item each node $v_{K} \in V'$ represents a distinct key $K \in \mathcal{K}$, with $G'.\glabel(v_{K}) = K[0]$;
 \item each node has a key $G'.\gkey(v_{K}) = K$ and a value $G'.\gvalue(v_{K}) = V_{K}$, where $V_{K}$ is the set of nodes $\mathcal{S}.\gnode(S, i) \in V$ for $S \in \mathcal{S}$ and positions $i$ such that $S[i, i+\abs{K}-1] = K$; and
@@
 
-We can prune de~Bruijn graphs by merging keys sharing a common prefix, if the values are identical.
+We can compress de~Bruijn graphs structurally by merging keys sharing a common prefix, if the corresponding values are identical.
 These pruned de~Bruijn graphs, which arise naturally from GCSA construction, are similar to manifold de Bruijn graphs \cite{Lin2014}.
@@
 Let $G = (V, E)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v' \in V'$ having string $K$ as a proper prefix of $G'.\gkey(v')$.
-If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u', v' \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus V'_{K}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
+If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u', v' \in V'_{K}$, the path graph with key set $(\mathcal{K} \setminus \set{ G'.\gkey(v') \mid v' \in V'_{K}}) \cup \set{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
 \end{lemma}
 
-We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
+We can compress a de~Bruijn graph structurally by merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold.
 Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $G'' = (V'', E'')$ be the same graph after further pruning.
@@
 \end{align*}
+
 In order to support $\locate$ queries, we \emph{sample} the values of a node, if the node has multiple incoming edges or the values cannot be derived from the predecessor.
@@
 We can improve the query performance with (maximally pruned) de~Bruijn graphs by using a \emph{simplified encoding} (Appendix~\ref{appendix:encoding}).
-We replace $\BWT$ and $\bvIN$ with bitvectors $B_{c}$ for all $c \in \Sigma$, such that $B_{c}[j] = 1$ if and only if node $j$ in lexicographic order has a predecessor with label $c$.
+We replace $\BWT$ and $\bvIN$ with bitvectors $B_{c}$ for all $c \in \Sigma$, where $B_{c}[j] = 1$ if and only if node $j$ in lexicographic order has a predecessor with label $c$.
 This simplifies backward searching to
@@
 To avoid excessive growth, we removed paths where \kmer{16}s crossed more than $4$ nontrivial edges with \texttt{vg mod -p -l -e 4}, and subgraphs shorter than $100$ bases with \texttt{vg mod -S -l 100}.
-We extracted all paths of length $16$ from the forward strands of the graphs.
+We extracted all paths of length $16$ from the forward strand of the graph.
 There were a total of 4.80~billion paths with 1.53~billion distinct labels.
@@
 We can build a whole-genome index overnight using less than 96~gigabytes of memory, including disk cache.
-The index contains $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the path graph only uses 4.4\nobreakdash--5.8~billion nodes to represent them.
-For $k = 128$, GCSA2 requires 0.68~bits per \kmer{k}, out of which 0.32~bits is used for the path graph.
-Extensions based on suffix trees increase the size to 1.12~bits per \kmer{k}.
+The index contains $1.031^{k} \cdot 2.348$~billion \kmer{k}s, but the path graph only uses 4.4\nobreakdash--5.7~billion nodes to represent them.
+For $k = 128$, GCSA2 requires 0.63~bits per \kmer{k}, out of which 0.28~bits is used for the path graph.
+Extensions based on suffix trees increase the size to 1.08~bits per \kmer{k}.
 
@@
 \hline
- 32 & 8.62 h & 63.1 GB & 387 GB & 1.37 TB & 0.88 TB \\
- 64 & 12.1 h & 55.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
-128 & 16.1 h & 56.6 GB & 478 GB & 2.78 TB & 2.25 TB \\
+ 32 & 7.44 h & 59.8 GB & 387 GB & 1.37 TB & 0.88 TB \\
+ 64 & 10.4 h & 51.9 GB & 415 GB & 2.03 TB & 1.51 TB \\
+128 & 14.1 h & 52.3 GB & 478 GB & 2.78 TB & 2.25 TB \\
 \hline
@@
 \hline
- 32 & 6.20G & 4.37G & 3.28 GB / 4.55 bits & 9.97 GB / 13.8 bits & 13.6 GB / 18.9 bits \\
- 64 & 16.7G & 5.24G & 3.93 GB / 2.03 bits & 9.19 GB / 4.74 bits & 14.1 GB / 7.27 bits \\
-128 &  116G & 5.73G & 4.29 GB / 0.32 bits & 9.17 GB / 0.68 bits & 15.2 GB / 1.12 bits \\
+ 32 & 6.20G & 4.37G & 2.89 GB / 4.00 bits & 9.50 GB / 13.2 bits & 13.2 GB / 18.2 bits \\
+ 64 & 16.7G & 5.24G & 3.46 GB / 1.78 bits & 8.64 GB / 4.46 bits & 13.6 GB / 6.99 bits \\
+128 &  116G & 5.73G & 3.78 GB / 0.28 bits & 8.58 GB / 0.63 bits & 14.6 GB / 1.08 bits \\
 \hline
@@
 With larger values of $k$, the path graph resembles the input graph better, and we sample less values.
-For the same reason, index size decreases with large values of $k$, even though the graph requires more space.
+For the same reason, index size decreases with larger values of $k$, even though the graph requires more space.
 
@@
 RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges.
-Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does not have the optimizations, we can estimate that GCSA is 4x slower than GCSA2.
+Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.
 
@@
 As a $\parent$ query takes comparable time to a single step of backward searching, it will not be a bottleneck in finding maximal exact matches.
-Counting the number of distinct occurrences with a $\countq$ query is clearly faster than retrieving even a single occurrence.
+Counting the number of distinct occurrences with a $\countq$ query is faster than retrieving even a single occurrence.
 
@@
 We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space.
-The resulting index takes less than 16~gigabytes, or 1.12~bits per \kmer{k} for the \orderk{128} index with extensions.
+The resulting index takes less than 15~gigabytes, or 1.08~bits per \kmer{k} for the \orderk{128} index with extensions.
 Query performance is comparable to that of fast FM-indexes for sequences.
@@
 We may want to determine whether a pattern matches known \emph{haplotypes} or only their recombinations.
-As GCSA2 does not support this directly, vg must determine this afterwards using a separate structure \cite{Novak2016}.
+As GCSA2 does not support this directly, vg must determine it afterwards using a separate structure \cite{Novak2016}.
 The \emph{FM\nobreakdash-index of alignment} \cite{Na2015,Na2016} embeds the haplotypes directly in a GCSA-like index and reports the haplotypes matching the $\find$ query.
-While the solution depends on the specific properties of the graph, it could be possible to extend it to work with any GCSA.
+While the solution depends on specific properties of the graph, it could be possible to extend it to work with any GCSA.
 
@@
 If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
-Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ bet the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
+Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ be the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
 
@@
 If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
-Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1]$ be a substring represented by node $u'$ in the \kcollection{k} used for building the path graph.
+Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of the \kcollection{k} used for building the path graph.
 As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
 
-Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ is a prefix of key $G'.\gkey(u')$.
+Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ prefix-matches key $G'.\gkey(u')$.
 If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
@@
 
+\begin{Definition}[Equivalent paths] ~\\
+Let $G'$ and $G''$ be path graphs of the same graph, and let $P' = v'_{0} \dotsm v'_{n-1}$ and $P'' = v'_{0} \dotsm v''_{n-1}$ be paths in graphs $G'$ and $G''$, respectively.
+We say that paths $P'$ and $P''$ are \emph{equivalent}, if for $0 \le i < n$, keys $G'.\gkey(v'_{i})$ and $G''.\gkey(v''_{i})$ have a common prefix $K_{i}$ such that $G'.\gvalue(v') = G''.\gvalue(v'')$ for all nodes $v' \in V'$ and $v'' \in V''$ having $K_{i}$ as a prefix of their keys.
+\end{Definition}
+
 \begin{proof}[Lemma~\ref{lemma:dbg-prune}: Pruning]
-Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of $G$, let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, and let $\mathcal{S}$ be the \kcollection{k} used for building the path graphs.
+Let $G'' = (V'', E'')$ be the path graph corresponding to the new key set, let $v''_{K} \in V''$ be the node with key $K$, and let $\mathcal{S}$ be the \kcollection{k} used to define the path graphs.
 
-Let $X \in \patternset$ be a pattern with $1 \le \abs{X} \le k$, and let $S[i, i+k-1] = G_{d}.\gkey(v_{0})$ be a substring of $\mathcal{S}$, where $v_{0} \in G_{d}.\find(X)$.
-By Lemma~\ref{lemma:pg-keys}, $X$ is a prefix of the substring.
-We extend the substring to length $\abs{X}+k-1$ without changing its starting position.
-The extended substring $S[i, i+\abs{X}+k-2]$ corresponds to a path $P = v_{0} \dotsm v_{\abs{X}}$ with label $X$ in the de~Bruijn graph, with $G_{d}.\gkey(v_{j}) = S[i+j, i+j+k-1]$.
-If we change each node $v_{j}$ to the corresponding equivalence class $v''_{j} \in V''$, we get a path with label $X$ in graph $G''$.
+Consider the edge $(u', v') \in E'$ defined by substring $S'$ of $\mathcal{S}$.
+The same substring also defines an edge $(u'', v'') \in E''$, where either (a) $G''.\gkey(u'') = G'.\gkey(u')$ or (b) $u'' = v''_{K}$ and $u' \in V'_{K}$, and the same holds for nodes $v''$ and $v'$.
+We can hence transform any path in graph $G'$ into an equivalent path in graph $G''$ by replacing nodes $v' \in V'_{K}$ with node $v''_{K}$.
 
-Now let $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ be a path matching pattern $X$ in graph $G''$.
-If $\abs{X} = 1$, we can choose any $v_{0} \in V_{d}(v''_{0})$.
-Assume that we have chosen node $v_{j+1}$ from the equivalence class $V_{d}(v''_{j+1})$ and that $X[j+1, \abs{X}-1]$ is a prefix of $G_{d}.\gkey(v_{j+1})$.
-As edge $(v''_{j}, v''_{j+1}) \in E''$ exists, it must be that $\abs{G.\gpred(v, X[j])} > 0$ for a node $v \in G''.\gvalue(v''_{j+1})$.
-Therefore substring $X[j] \cdot G''.\gkey(v''_{j+1})$ exists in $\mathcal{S}$.
+Let $P'' = v''_{0} \dotsm v''_{\abs{P''}-1}$ be a in graph $G''$.
+We transform it into an equivalent path $P' = v'_{0} \dotsm v'_{\abs{P''}-1}$ in graph $G'$.
+There are two cases for $\abs{P''} = 1$.
+If $v''_{0} = v''_{K}$, we can replace it with any $v'_{0} \in V'_{K}$.
+Otherwise we use the node $v'_{0} \in V'$ such that $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$.
 
-Let $Y$ be the longer of strings $G''.\gkey(v''_{j})$ and $X[j] \cdot G''.\gkey(v''_{j+1})$.
-As $Y$ is a substring of $\mathcal{S}$, there is a node $v_{j} \in V_{d}$ such that key $G_{d}.\gkey(v_{j})$ prefix-matches $Y$.
-Key $G_{d}.\gkey(v_{j})$ has both $X[j, \abs{X}-1]$ and $G''.\gkey(v''_{j})$ as its prefixes, and hence $v_{j} \in V_{d}(v''_{j})$.
-When $j = 0$, we can extend key $G_{d}.\gkey(v_{0})$ to get a path starting from $v_{0}$ and matching $X$ in graph $G_{d}$.
+In the general case $\abs{P''} > 1$, assume that we have transformed the suffix $v''_{1} \dotsm v''_{\abs{P''}-1}$ of path $P''$ into an equivalent path $v'_{1} \dotsm v'_{\abs{P''}-1}$ in graph $G'$.
+Because $G'.\gvalue(v'_{1}) = G''.\gvalue(v''_{1})$, node $v'_{1}$ must have a predecessor with label $c = G''.\glabel(v''_{0})$.
+We can choose any such predecessor as node $v'_{0}$.
 
-Given a node $v_{0} \in G_{d}.\find(X)$, we can build a path $P'' = v''_{0} \dotsm v''_{\abs{X}-1}$ matching the pattern in graph $G''$, with $v_{0} \in V_{d}(v''_{0})$.
-We can also start from path $P''$ and determine the node $v_{0}$.
-As $G_{d}.\gvalue(v_{0}) = G''.\gvalue(v''_{0})$, we have $G_{d}.\locate(X) = G''.\locate(X)$.
+Consider the predecessors $v' \in G'.\gpred(v'_{1}, c)$ and $v'' \in G''.\gpred(v''_{1}, c)$. Their keys prefix-match string $Y = c \cdot K_{1}$. There are three cases:
+\begin{enumerate}
+\item If $Y$ is a prefix of $G'.\gkey(v')$, the key $G'.\gkey(w')$ of every successor $w'$ of node $v'$ prefix-matches $K_{1}$ and hence has $K_{1}$ as a prefix.
+Therefore $G'.\gvalue(v')$ is the union of sets $G.\gpred(v, c)$ over $v \in G'.\gvalue(v'_{1})$.
+If $Y$ is also a prefix of $G''.\gkey(v'')$, nodes $v'$ and $v''$ have identical value sets by the same reasoning.
+
+\item If $Y$ is a prefix of $G'.\gkey(v')$ and $G''.\gkey(v'')$ is a proper prefix of $Y$, there is only one possible predecessor $v'' = v''_{0}$. Hence $v' \in V'_{K}$ and $v''_{0} = v''_{K}$.
+
+\item If key $G'.\gkey(v')$ is a proper prefix of string $Y$, there is only one possible predecessor $v' = v'_{0}$.
+Because key $G''.\gkey(v''_{0})$ prefix-matches string $Y$, it must also be a proper prefix of the string.
+Hence either $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$ or $v'_{0} \in V'_{K}$ and $v''_{0} = v''_{K}$.
+\end{enumerate}
+The shorter of $Y$ and $G''.\gkey(v''_{0})$ can be used as $K_{0}$.
+
+We can transform any path $P'$ in graph $G'$ into an equivalent path $P''$ in graph $G''$, and the other way around.
+Because the labels of equivalent paths and the value sets of their start nodes are identical, we have $G'.\locate(X) = G''.\locate(X)$ for all patterns $X \in \patternset$ with $\abs{X} > 0$.
 \end{proof}
@@
 We sort the nodes $V'$ by their keys in lexicographic order and generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$ from the nodes in that order.
-For each node $v' \in V'$, we append $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$; $\bvIN$ with the indegree encoded as $0^{G.\gindegree(v')-1} 1$; and $\bvOUT$ with the outdegree as $0^{G.\goutdegree(v')-1} 1$.
+For each node $v' \in V'$, we append $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$; $\bvIN$ with the indegree encoded as $0^{G'.\gindegree(v')-1} 1$; and $\bvOUT$ with the outdegree as $0^{G'.\goutdegree(v')-1} 1$.
 
@@
 For every $j \in [sp_{in}, ep_{in}]$ for a node $v' \in V'$, having $\LF(j) \in [sp_{out}, ep_{out}]$ for a node $u' \in V'$ implies an edge $(u', v') \in E'$.
-If we sort the incoming edges $(u', v') \in E'$ stably by strings $G'.\glabel(u') \cdot G'.\gkey(v')$, we have $(u'_{1}, v'_{1}) < (u'_{2}, v'_{2})$ in that order, if $G'.\gkey(u'_{1}) < G'.\gkey(u'_{2})$.
 
@@
 Query $G'.\locate(X)$ retrieves the values $G'.\gvalue(v')$ for nodes $v' \in G'.\find(X)$ and filters out duplicates.
-Storing the values explicitly for all nodes would require too much space.
-Like FM\nobreakdash-indexes, GCSA uses a \emph{sampling} scheme to avoid that.
+Instead of storing the values explicitly for all nodes, GCSA uses a \emph{sampling} scheme to save space.
 We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are conveniently chosen integers.
@@
 If node $v' \in V'$ with lexicographic rank $i$ has only one predecessor, the lexicographic rank of the predecessor is $G'.\LF(i) = \bvOUT.\rank(\LF(\bvIN.\select(i, 1) + 1), 1)$.
-If lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.
+If the lexicographic rank $G'.\LF^{k}(i)$ corresponding to node $w' \in V'$ is the first sampled node we encounter, we know that $G'.\gvalue(v') = \set{w+k \mid w \in G'.\gvalue(w')}$.
 
@@
 
-Because we have to store path labels explicitly in the keys, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$.
+Because we store path labels explicitly, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$.
 After $d$ doubling steps, the length of the paths is $2^{d} k$, and we can use them to build a maximally pruned \orderk{(2^{d} k)} de~Bruijn graph.
 Each doubling step consists of a \emph{pruning} step, followed by an \emph{extension} step.
-The pruning step uses a limited form of Lemma~\ref{lemma:dbg-prune} and applies it to all possible ranges of paths.
+The pruning step applies a limited form of Lemma~\ref{lemma:dbg-prune} to lexicographic ranges of paths.
 Given two paths $P$ and $P'$, we can determine the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array.
-If all paths with a certain prefix of their labels start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.
+If all paths sharing a prefix start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.
 
@@
 
-The doubling steps are followed by the \emph{merging step}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.
-First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
+The doubling steps are followed by the \emph{merging step}, which transforms the paths into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$.
+We merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$.
 If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$.
@@
 Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs.
-To avoid that, we store them on \emph{disk} when possible.
-Each chromosome forms a subgraph that is disjoint from the other chromosomes, except near the source node and the sink node.
-Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order.
-An extension step can be done separately for each of the chromosomes.
-A pruning step takes a set of sorted lists of paths and merges the lists.
+To avoid that, we keep them on \emph{disk} when possible.
+The subgraph corresponding to each chromosome is stored in a separate file, with paths sorted by their labels in lexicographic order.
+Extension steps are done separately for each of the chromosomes.
+A pruning step merges the sorted files.
 It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged.
 The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer.
-The merging step is mostly the same as a pruning step, but the output format is different.
+The merging step works in a similar way.
 
 After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we \emph{build the index}.
-Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing.
-Because graph $G''$ is a path graph, there is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$.
+Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$, while the outdegree sequence $\bvOUT$ requires further processing.
+There is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$.
 We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers.
 The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a character $c \in \Sigma$.
-We can also sample the nodes for $\locate$ queries during the same scans.
+We also sample the nodes for $\locate$ queries during the same scans.
 
@@
 We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
-If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
-Backward searching is simplified to:
+If node $v' \in V$ with lexicographic rank $i$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
+The backward step becomes:
 \begin{align*}
@@
 Computing $G'.\LF(i)$ can expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine the character used in the backward step.
-However, if the alphabet is small, this is still faster than the $\select$ queries in the general encoding.
+If the alphabet is small, this is still faster than the $\select$ queries in the general encoding.
 We can further improve the time/space trade-off by compressing the bitvectors $B_{c}$ for rare characters (e.g.~$\baseN$, $\#$, and $\$$) and checking $B_{c}[i]$ first for the frequent characters (e.g.~bases) when computing $G'.\LF(i)$.
@@
 
-Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys.
-We define the \emph{LCP array} $\LCP[0, \abs{V'}-1]$ of graph $G'$ as an array, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
-The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree.
-The compacted trie can be considered the \emph{suffix tree} of graph $G'$.
+Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order.
+The \emph{LCP array} of graph $G'$ is an array $\LCP[0, \abs{V'}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
+If we build a trie of keys $G'.\gkey(v')$ for $v' \in V'$ and compact the unary paths into single edges, the resulting tree can be considered the \orderk{k} \emph{suffix tree} of graph $G'$.
 
@@
 These ranges can be determined from the LCP array.
-If we use the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.
+If we use the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we can support the full functionality of the suffix tree \cite{Fischer2009a}.
 
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
-By selecting $x$ according to cache line size, the queries require $\Oh{\log_{x} \abs{V'}}$ random memory accesses, which in practice dominate the computation time.
-Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
+If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time with $\Oh{\log_{x} \abs{V'}}$ random memory accesses, while using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+Because the nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
 By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern.
@@
 BWA-MEM \cite{Li2013} aligns reads to a reference genome by finding \emph{maximal exact matches} between the read and the reference.
-It stores both the reference and its reverse complement in the same index, allowing it to extend the pattern in both directions.
+It indexes both the reference and its reverse complement, allowing it to extend the pattern in both directions.
 GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains key $\revcomp{K}$).
-However, we can search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
+However, we can search for maximal exact matches by using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}.
 
@@
 While graphs are a natural formalism for representing genetic variation, they cannot adequately represent certain types of \emph{rearrangements}.
-For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph.
+For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can either have a single copy or multiple copies of $S$ in the graph.
 Neither option is good in a reference genome.

### 2016-10-13 paper/paper.tex
Yet another round of proofreading.

@@
 A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$.
-We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s.
+We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and refer to substrings of length $k$ as \kmer{k}s.
 Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$.
@@
 \item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
-\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = XY[k-1]$ occurring in $\mathcal{S}$.
+\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{(k+1)} $X[0]Y = XY[k-1]$ occurring in $\mathcal{S}$.
 \end{itemize}
@@
 The \emph{\kmer{k} index} is the simplest path index.
-It consists of a set of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ is a \kmer{k} and $V_{X} \subseteq V$ is the set of start nodes of the paths matching the \kmer{k}.
+It consists of a set of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ is a \kmer{k} and $V_{X} \subseteq V$ is the set of the start nodes of the paths matching the \kmer{k}.
 If we store the pairs in a hash table, we can quickly search for patterns of length $k$.
@@
 $$
-Apart from some technicalities near the source/sink nodes, $G'.\gvalue(v)$ is the set of start nodes of the paths matching pattern $G'.\gkey(v)$ in graph $G$.
+Apart from some technicalities near the source/sink nodes, $G'.\gvalue(v)$ is the set of the start nodes of the paths matching pattern $G'.\gkey(v)$ in graph $G$.
 The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index).
@@
 We can encode the de~Bruijn graph as a GCSA using $(\abs{V'} + \oh{\abs{V'}})(\log \sigma + 2)$ bits.
-By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses a couple of bytes per \kmer{k} for a typical variation graphs (see Section~\ref{sect:experiments}).
+By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses a couple of bytes per \kmer{k} for typical variation graphs (see Section~\ref{sect:experiments}).
 While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.
@@
 
-When a path of length $k' < k$ has a unique label, its extensions become \emph{redundant} nodes in an \orderk{k} de~Bruijn graph, if the path branches after the first $k'$ characters.
-By using shorter keys when possible, we can build a smaller graph equivalent to the de~Bruijn graph as a path index.
+When a path of length $k' < k$ has a unique label, its extensions become \emph{redundant} nodes in the \orderk{k} de~Bruijn graph, if the path branches after the first $k'$ characters.
+By using shorter keys when possible, we can build a smaller graph that is equivalent to the de~Bruijn graph as a path index.
 
@@
 \begin{Definition}[Maximally pruned graph] ~\\
-Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
+Let $G'$ be a a pruned de~Bruijn graph of graph $G$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
 \end{Definition}
@@
 As the default FM-index in SDSL, it prioritizes query performance over compression.
-BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with SA sample period 32).
+BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with the default SA sample period 32).
 Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.
@@
 GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6.
-On the average, each sample contains 1.53~values and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7.
+On the average, each sample contains 1.53~values, and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7.
 SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates.
@@
 It uses a de~Bruijn graph as a \kmer{k} index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index.
-The index supports queries of length up to $k$ exactly and longer queries with false positives.
+The index supports queries of length up to $k$ exactly, and longer queries with false positives.
 GCSA2 also includes extensions based on suffix trees; other extensions have been considered but not implemented (see Appendix~\ref{appendix:hypertext}).
@@
 If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$.
-Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ be the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
+Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ be the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m_{i+1}]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.
 
@@
 There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$.
-Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$ for all nodes $u' \in M_{i}$.
+Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m_{i+1}]$ for all nodes $u' \in M_{i}$.
 
-Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m]$.
-If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m_{i+1}]$.
+If the key is a prefix of string $X[i, i+m_{i+1}]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
 Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of the \kcollection{k} used for building the path graph.
-As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
+As $S[j+1, j+m_{i+1}] = X[i+1, i+m_{i+1}]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.
 
-Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ prefix-matches key $G'.\gkey(u')$.
-If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$.
-Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
+Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m_{i+1}]$ prefix-matches key $G'.\gkey(u')$.
+If $\abs{M_{i}} > 1$, string $X[i, i+m_{i+1}]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can set $m_{i} = m_{i+1}+1$.
+Otherwise we set $m_{i} = \min(m_{i+1}+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
 \end{proof}
@@
 If $v''_{0} = v''_{K}$, we can replace it with any $v'_{0} \in V'_{K}$.
-Otherwise we use the node $v'_{0} \in V'$ such that $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$.
+Otherwise we use the node $v'_{0} \in V'$ with $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$.
 
@@
 \end{enumerate}
-The shorter of $Y$ and $G''.\gkey(v''_{0})$ can be used as $K_{0}$.
+In every case, $G'.\gvalue(v'_{0}) = G''.\gvalue(v''_{0})$, and we can use shorter of strings $Y$ and $G''.\gkey(v''_{0})$ as $K_{0}$.
 
@@
 Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$.
-Then string $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$.
+String $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$.
 There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.
 
-Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}.
-Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$.
+Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of $\mathcal{S}$.
+Because $G'.\gkey(v')$ is a substring of $G'.\gkey(u')$ and $G'$ is a pruned de~Bruijn graph, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in $\mathcal{S}$ is $G'.\gvalue(v')$.
 As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.
@@
 Instead of storing the values explicitly for all nodes, GCSA uses a \emph{sampling} scheme to save space.
-We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are conveniently chosen integers.
+We assume that the nodes of the input graph $G = (V, E)$ are conveniently chosen integers.
 If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.
@@
 We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$.
-For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$.
+For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as for bitvectors $\bvIN$ and $\bvOUT$.
 We store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each.
@@
 The pruning step applies a limited form of Lemma~\ref{lemma:dbg-prune} to lexicographic ranges of paths.
-Given two paths $P$ and $P'$, we can determine the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array.
+Given two paths $P$ and $P'$, we can determine the length of the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array.
 If all paths sharing a prefix start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.
@@
 If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$.
-We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$.
+We also store the union of predecessor labels as $v'.\gpred = \bigcup_{i=0}^{m-1} P_{i}.\gpred$.
 We then apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.
 
-Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs.
+Storing the paths and the graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs.
 To avoid that, we keep them on \emph{disk} when possible.
@@
 Extension steps are done separately for each of the chromosomes.
-A pruning step merges the sorted files.
+The pruning step merges the sorted files.
 It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged.
@@
 After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we \emph{build the index}.
-Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$, while the outdegree sequence $\bvOUT$ requires further processing.
-There is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$.
+Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $v''.\gpred$, while the outdegree sequence $\bvOUT$ requires further processing.
+There is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in v''.\gpred$.
 We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers.
@@
 This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}).
-We can use the simplified encoding of the original GCSA \cite{Siren2014} for the GCSA of such path graphs.
+We can use the simplified encoding of the original GCSA \cite{Siren2014} with such path graphs.
 
-We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
+We replace the sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$.
 If node $v' \in V$ with lexicographic rank $i$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$.
@@
 The \emph{LCP array} of graph $G'$ is an array $\LCP[0, \abs{V'}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$.
-If we build a trie of keys $G'.\gkey(v')$ for $v' \in V'$ and compact the unary paths into single edges, the resulting tree can be considered the \orderk{k} \emph{suffix tree} of graph $G'$.
+If we build a trie of keys $G'.\gkey(v')$ for $v' \in V'$ and compact the unary paths into single edges, the resulting tree can be considered the \emph{suffix tree} of graph $G'$.
 
@@
 
-If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time with $\Oh{\log_{x} \abs{V'}}$ random memory accesses, while using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
+If we build an $x$\nobreakdash-ary tree over the LCP array, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time with $\Oh{\log_{x} \abs{V'}}$ random memory accesses, while using $\frac{x}{x-1} \abs{V'} \log k$ bits of space.
 Because the nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use.
@@
 \begin{align*}
-G'.\countq(X) & = G'.\countq([sp, ep]) \\
+G'.\countq(X) & = G'.\countq(sp, ep) \\
  & = \sum_{i = sp}^{ep} (A[i] + 1) - \sum_{i = sp}^{ep - 1} R[i].
@@
 With a single copy, we create paths in the reference that do not correspond to any valid genome.
-With the second option, we duplicate sequences that are not only identical but also functionally the same.
+With the second option, we lose the information that both copies of $S$ are the same sequence.
 One solution is to use \emph{context-free grammars}.

### 2016-10-13 paper/paper.tex
Spellchecking

@@
 It was not possible the with the original GCSA, as the construction algorithm required a prefix-range-sorted graph.
-The extendend functionality depends on the new construction algorithm and on limiting maximum query length.
+The extended functionality depends on the new construction algorithm and on limiting maximum query length.
 
@@
 Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base.
-The alphabet may contain other technical characters in addition to the enmarker $\$$.
+The alphabet may contain other technical characters in addition to the endmarker $\$$.
 Each character $c$ of the DNA alphabet has a \emph{complement} $\dnacomp{c}$ defined as $\dnacomp{\baseA} = \baseT$, $\dnacomp{\baseC} = \baseG$, $\dnacomp{\baseG} = \baseC$, $\dnacomp{\baseT} = \baseA$, and $\dnacomp{c} = c$ for other characters $c$.

### 2016-11-01 paper/paper.tex
Final ALENEX 2017 version of the paper

@@
 We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence.
-SSA is the SDSL implementation (\texttt{csa\_wt<>} with SA sample period 17) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT.
+SSA is the SDSL implementation (\texttt{csa\_wt<>}) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT.
 As the default FM-index in SDSL, it prioritizes query performance over compression.
+We used SSA with SA sample period 17 for good $\locate$ performance.
 BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with the default SA sample period 32).
@@
 As building the original GCSA requires around $65n$ bytes of memory for a path graph with $n$ nodes, we could not compare GCSA and GCSA2 directly on a system with 256~gigabytes of memory.
-Instead, we used \emph{RLCSA} \cite{Maekinen2010} (May~2016 version with SA sample period 32) as a proxy.
+Instead, we used \emph{RLCSA} \cite{Maekinen2010} (May~2016 version) as a proxy.
 The RLCSA is an FM-index for repetitive sequence collections using the same basic components as the original GCSA.
-With a mixed query load, RLCSA is 1.5x to 3x faster than GCSA, depending on algorithmic overhead and the mix of $\find$ and $\locate$ queries \cite{Siren2014}.
+Under a mixed query load, RLCSA with SA sample period 32 is 1.5x to 3x faster than GCSA, depending on algorithmic overhead and the mix of $\find$ and $\locate$ queries \cite{Siren2014}.
 
@@
 GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6.
-On the average, each sample contains 1.53~values, and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7.
-SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates.
-BWA is closer to GCSA2 in $\locate$ performance, which can be explained by sparser sampling than in SSA.
+On the average, GCSA2 calls $\locate$ for 2.84~nodes per distinct value, making the amount of work comparable to sample period 30.2.
+SSA with sample period 17 is 2.4x faster than GCSA2, mostly because it has to do less work.
+BWA with sample period 32 is closer to GCSA2 in $\locate$ performance.
 RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges.
-Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.
+Assuming that $\locate$ queries are 3x slower in GCSA than in RLCSA, as GCSA does not use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.
 
